### 2018-04-26 paper/gd1.tex
add paper files

% This file is part of the stream_information project.
% Copyright 2017 the authors. All rights reserved.

% # style notes
% - it is Cram\'er--Rao not Cram\'er-Rao. And yet Fisher-matrix not Fisher--matrix.

\documentclass[modern]{aastex62}

% \usepackage{amsmath}

% typography
\setlength{\parindent}{1.\baselineskip}
% \newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\package}[1]{\textsl{#1}}

% aastex parameters
% \received{not yet; THIS IS A DRAFT}
%\revised{not yet}
%\accepted{not yet}
% % Adds "Submitted to " the arguement.
% \submitjournal{ApJ}
\shorttitle{GD-1 in Gaia DR2}
\shortauthors{price-whelan \& bonaca}

%@arxiver{}

\begin{document}\sloppy\sloppypar\raggedbottom\frenchspacing % trust me

\title{Gaia DR2 view of the GD-1 stellar stream}

\author[0000-0003-0872-7098]{Adrian~M.~Price-Whelan}
\affiliation{Department of Astrophysical Sciences,
             Princeton University, Princeton, NJ 08544, USA}
\email{adrn@astro.princeton.edu}
\correspondingauthor{Adrian M. Price-Whelan}

\author[0000-0002-7846-9787]{Ana Bonaca}
\affil{Harvard--Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA}

\begin{abstract}\noindent % trust me

\end{abstract}

\keywords{Galaxy: halo --- dark matter}

\section{Introduction}
\label{sec:intro}

\section{Data}
\label{sec:data}

\section{Results}
\label{sec:results}

\subsection{Density along the stream}
\label{sec:density}

\subsection{Kinematic profile}
\label{sec:kinematics}

\section{Discussion}
\label{sec:discussion}

\acknowledgements{
thanks: Casey, Geha, Hogg, Spergel
CCA for hospitality
AB acknowledges generous support from the Institute for Theory and Computation at Harvard University.
All code used in this work and all results are available at \url{https://github.com/adrn/GD1-DR2}.
}

\software{
    \package{Astropy} \citep{astropy},
    \package{gala} \citep{gala},
    \package{IPython} \citep{ipython},
    \package{matplotlib} \citep{mpl},
    \package{numpy} \citep{numpy},
    \package{scipy} \citep{scipy}
}

\bibliographystyle{aasjournal}
\bibliography{gd1}

\end{document}

### 2018-04-26 paper/gd1.tex
regroup results
plot and notebooks

@@
 \newcommand{\package}[1]{\textsl{#1}}
+\newcommand{\gaia}{\textsl{Gaia}}
 
@@
 
+(\citealt{Schlegel:1998, Schlafly:2011}; hereafter SFD)
+
+Maximum $A_G$ is $\approx$0.07 mag in both the under-density and gap regions.
+
+% Notebook: GD1-dust-and-completeness
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=\textwidth]{sfd.pdf}
+\end{center}
+\caption{%
+Colored background shows the \gaia\ $G$-band extinction, $A_G$, in the GD-1
+coordinate system.
+Vertical lines roughly show the regions identified as the under-density (UD) and
+gap.
+The maximum extinction in the UD or Gap regions is $\approx$0.07 mag.
+\label{fig:sfd}
+}
+\end{figure}
+
 \section{Results}
@@
 
-\subsection{Density along the stream}
-\label{sec:density}
+\subsection{Global properties}
+\label{sec:res_global}
 
-\subsection{Kinematic profile}
-\label{sec:kinematics}
+\subsection{Gap}
+\label{sec:res_gap}
+
+\subsection{Underdensity}
+\label{sec:res_underdensity}
 
@@
 \acknowledgements{
-thanks: Casey, Geha, Hogg, Spergel
+thanks: Casey, Geha, Hogg, Spergel, Koposov
 CCA for hospitality
@@
 
+\clearpage
+
+\appendix
+\section{Completeness check and data validation}
+\label{sec:validate}
+
+% % Notebook:
+% \begin{figure}[h]
+% \begin{center}
+% \includegraphics[width=0.7\textwidth]{nvisits.pdf}
+% \end{center}
+% \caption{%
+% TODO
+% \label{fig:TODO}
+% }
+% \end{figure}
+
 \end{document}

### 2018-04-26 paper/gd1.tex
dust

@@
 % - it is Cram\'er--Rao not Cram\'er-Rao. And yet Fisher-matrix not Fisher--matrix.
+
+% TODO:
+% - Reference for the dustmaps package? See software section.
 
@@
 
-(\citealt{Schlegel:1998, Schlafly:2011}; hereafter SFD)
-
-Maximum $A_G$ is $\approx$0.07 mag in both the under-density and gap regions.
+Most of the clearly identifiable portions of the GD-1 stream are located at high
+Galactic latitudes ($b > 35^\circ$), and we therefore do not expect significant
+dust extinction or variations in extinction along the stream.
+\figurename~\ref{fig:sfd} shows the $V$-band extinction in the region around the
+GD-1 stream, computed from the Schlegel-Finkbeiner-Davis extinction map
+(\cite{Schlegel:1998}; hereafter SFD).
+Vertical lines indicate the two regions clearly identifiable as under-densities
+from the proper-motion-selected stream members.
+The maximum $A_V$ is $\approx$0.07 mag in both the under-density and gap
+regions.
+The dispersion in $A_V$ in the region $-60^\circ < \phi_1 < 0^\circ$ and
+$-1^\circ < \phi_1 < 1^\circ$ is $\std_{A_V} \approx 0.03~\textrm{mag}$.
 
@@
     \package{Astropy} \citep{astropy},
+    \package{dustmaps}\footnote{\url{https://github.com/gregreen/dustmaps}},
     \package{gala} \citep{gala},

### 2018-04-26 paper/gd1.tex
acknowledgments

--- 2018-04-26 paper/gd1.tex
@@
 thanks: Casey, Geha, Hogg, Spergel
-CCA for hospitality
+This research was started at the NYC Gaia DR2 Workshop at the Center for Computational Astrophysics of the Flatiron Institute in 2018 April.
 AB acknowledges generous support from the Institute for Theory and Computation at Harvard University.

### 2018-04-26 paper/gd1.tex

--- 2018-04-26 paper/gd1.tex
@@
 \acknowledgements{
-thanks: Casey, Geha, Hogg, Spergel, Koposov
-CCA for hospitality
+thanks: Belokurov, Casey, Geha, Hogg, Johnston, Koposov, Spergel
+This research was started at the NYC Gaia DR2 Workshop at the Center for Computational Astrophysics of the Flatiron Institute in 2018 April.
 AB acknowledges generous support from the Institute for Theory and Computation at Harvard University.

### 2018-04-26 paper/gd1.tex
ack eddie

@@
 
-% \usepackage{amsmath}
+\usepackage{amsmath}
 
@@
 \acknowledgements{
-thanks: Belokurov, Casey, Geha, Hogg, Johnston, Koposov, Spergel
+thanks: Belokurov, Casey, Geha, Hogg, Johnston, Koposov, Schlafly, Spergel
 This research was started at the NYC Gaia DR2 Workshop at the Center for Computational Astrophysics of the Flatiron Institute in 2018 April.

### 2018-04-26 paper/gd1.tex
words

--- 2018-04-26 paper/gd1.tex
@@
 \setlength{\parindent}{1.\baselineskip}
-% \newcommand{\acronym}[1]{{\small{#1}}}
+\newcommand{\acronym}[1]{{\small{#1}}}
 \newcommand{\package}[1]{\textsl{#1}}
 \newcommand{\gaia}{\textsl{Gaia}}
+\newcommand{\DR}{\acronym{DR2}}
 
@@
 \begin{abstract}\noindent % trust me
-
+Tidally-disrupted globular clusters leave behind thin, dynamically-cold streams
+of stars that are extremely valuable tracers of the large- and small-scale
+properties of mass around the Galaxy.
+Many thin stellar streams have been discovered around the Milky Way, the
+majority of which exist in the Galactic halo and are therefore primarily
+sensitive to the distribution of dark matter.
+Here we present astrometric data for one of the longest stellar streams --- the
+``GD-1'' stream --- from data release 2 (\DR) of the \gaia\ mission.
+We re-discover significant density variations along the stream by selecting
+candidate member stars using the exquisite \gaia\ data alone, including the
+existence of a prominent, apparent gap in the stream.
+<Using photometric data from the \pans, we refine our selection ...>
+These density variations --- and especially the gap --- are possible indications
+of past interactions with large, massive perturbers.
+<estimate the mass of interaction...>
 \end{abstract}
@@
 \label{sec:intro}
+
+Dynamically cold stellar streams, likely formed from the tidal disruption of globular clusters, are sensitive to the large and small scale distribution of matter in the halo
+
+Thin stellar streams form from the tidal disruption of low-dispersion stellar systems such as globular clusters \citep[e.g.,][]{TODO}.
+To date, $\approx$13--14 thin streams have been discovered by searching large-area sky surveys such as the Sloan Digital Sky Survey (SDSS) and Pan-STARRS1 survey (PS1).
+These previously known thin streams span a wide range of lengths \citep[$\sim$$2$--$60$ deg;][]{bernard14,grillmair06} and masses \citep[$\sim$$10^3$--$10^5~\msun$;][]{TODO,TODO}, however, only two such streams have clear progenitor systems: NGC 5466 and Palomar 5 \citep{TODO,TODO}.
+Though it has been shown with many independent methods that modeling these thin streams can place strong constraints on the gravitational potential in which they form \citep{apw14,TODO}, many efforts to apply these models to real streams around the Milky Way have been hampered by the small number of known progenitor systems.
 
@@
 \acknowledgements{
-thanks: Belokurov, Casey, Geha, Hogg, Johnston, Koposov, Spergel
-This research was started at the NYC Gaia DR2 Workshop at the Center for Computational Astrophysics of the Flatiron Institute in 2018 April.
+thanks: Casey, Geha, Hogg, Spergel, Koposov
+CCA for hospitality
 AB acknowledges generous support from the Institute for Theory and Computation at Harvard University.

### 2018-04-27 paper/gd1.tex
words

@@
 We re-discover significant density variations along the stream by selecting
-candidate member stars using the exquisite \gaia\ data alone, including the
-existence of a prominent, apparent gap in the stream.
-<Using photometric data from the \pans, we refine our selection ...>
+candidate member stars using the exquisite \gaia\ data alone.
+Combined with filtering using Pan-STARRS photometry, we see clear evidence of a
+prominent, apparent gap in the stream, and at least XX other under-densities.
 These density variations --- and especially the gap --- are possible indications
-of past interactions with large, massive perturbers.
-<estimate the mass of interaction...>
+of past interactions with large, massive perturbers such as dark matter
+sub-halos.
 \end{abstract}

### 2018-04-27 paper/gd1.tex
abstract

--- 2018-04-26 paper/gd1.tex
@@
 We re-discover significant density variations along the stream by selecting
-candidate member stars using the exquisite \gaia\ data alone, including the
-existence of a prominent, apparent gap in the stream.
-<Using photometric data from the \pans, we refine our selection ...>
-These density variations --- and especially the gap --- are possible indications
-of past interactions with large, massive perturbers.
-<estimate the mass of interaction...>
+candidate member stars first using only the exquisite \gaia\ data.
+Combined with filtering using Pan-STARRS photometry, we see clear evidence of
+multiple prominent under-densities and haps in the stream.
+These density variations --- and especially the gaps --- are possible
+indications of past interactions with large, massive perturbers.
+Given that the orbit of the stream crosses the Milky Way disk at large radius,
+these perturbers are likely halo objects and have masses consistent with XX,
+suggestive of dark matter sub-halos.
 \end{abstract}
@@
 
-Dynamically cold stellar streams, likely formed from the tidal disruption of globular clusters, are sensitive to the large and small scale distribution of matter in the halo
+Dynamically cold stellar streams, likely formed from the tidal disruption of
+globular clusters, have been discovered throughout the Milky Way halo (e.g.,
+\citealt{TODO}).
 
-Thin stellar streams form from the tidal disruption of low-dispersion stellar systems such as globular clusters \citep[e.g.,][]{TODO}.
 To date, $\approx$13--14 thin streams have been discovered by searching large-area sky surveys such as the Sloan Digital Sky Survey (SDSS) and Pan-STARRS1 survey (PS1).
@@
 \label{sec:data}
+
+- Transform to GD-1 coordinates (Koposov et al.)
+    - Can see deviation from great circle
+- Cut to be close to latitude = 0
+- Cut to get proper motion over-density
+- Color-magnitude filtering
+
+Figure 1:
+- PM cut (panel a)
+- PM + CMD cut (panel b)
+- SFD (panel c)??
+
+Comment on these here - add figure for both in Appendix?
+- Dust / no significant extinction - combine with other figure
+- Scanning pattern - just discuss, no figure
 

### 2018-04-27 paper/gd1.tex
More words

@@
 Given that the orbit of the stream crosses the Milky Way disk at large radius,
-these perturbers are likely halo objects and have masses consistent with XX,
+these perturbers are likely halo objects and would have to have masses XX,
 suggestive of dark matter sub-halos.
@@
 
-Dynamically cold stellar streams, likely formed from the tidal disruption of
-globular clusters, have been discovered throughout the Milky Way halo (e.g.,
-\citealt{TODO}).
+Dynamically cold stellar streams are formed from the tidal disruption of stellar
+systems by the gravitational field of their host galaxy (e.g.,
+\citealt{Johnston:1998}).
+Over 30 candidate stellar streams have been discovered throughout the halo of
+the Milky Way and have a wide range of properties such as length, velocity
+dispersion, and density (e.g., \citealt{SoManyPeople,Bonaca:2012}; see also the
+recent review \citealt{Grillmair:2006, Newberg:2016}).
+Of the thin, likely globular-cluster-origin streams, only two have clear
+progenitor systems: NGC 5466 (\citealt{TODO}) and Palomar 5 (\citealt{TODO}).
 
-To date, $\approx$13--14 thin streams have been discovered by searching large-area sky surveys such as the Sloan Digital Sky Survey (SDSS) and Pan-STARRS1 survey (PS1).
-These previously known thin streams span a wide range of lengths \citep[$\sim$$2$--$60$ deg;][]{bernard14,grillmair06} and masses \citep[$\sim$$10^3$--$10^5~\msun$;][]{TODO,TODO}, however, only two such streams have clear progenitor systems: NGC 5466 and Palomar 5 \citep{TODO,TODO}.
+Stellar streams are
+
+\citealt{Bonaca:2018}
+
+Dynamical modeling of stellar streams or models of mean stream tracks have been used
+
 Though it has been shown with many independent methods that modeling these thin streams can place strong constraints on the gravitational potential in which they form \citep{apw14,TODO}, many efforts to apply these models to real streams around the Milky Way have been hampered by the small number of known progenitor systems.
+
+We revisit...do not try to model or fit for Galactic potential. Focus on stream properties, data.
 

### 2018-04-27 paper/gd1.tex
Stuff
Title

@@
 
-\title{Gaia DR2 view of the GD-1 stellar stream}
+\title{A first look at the GD-1 stellar stream with Gaia DR2}
 
@@
 The dispersion in $A_V$ in the region $-60^\circ < \phi_1 < 0^\circ$ and
-$-1^\circ < \phi_1 < 1^\circ$ is $\std_{A_V} \approx 0.03~\textrm{mag}$.
+$-1^\circ < \phi_1 < 1^\circ$ is $\sigma_{A_V} \approx 0.03~\textrm{mag}$.
 

### 2018-04-28 paper/gd1.tex
Story
abstract edits

--- 2018-04-27 paper/gd1.tex
@@
 \newcommand{\gaia}{\textsl{Gaia}}
+\newcommand{\pans}{\textsl{PanSTARRS}}
 \newcommand{\DR}{\acronym{DR2}}
+\newcommand{\msun}{\rm M_\odot}
 
@@
 \begin{abstract}\noindent % trust me
-Tidally-disrupted globular clusters leave behind thin, dynamically-cold streams
+Tidally-disrupted globular clusters are transformed into thin, dynamically-cold streams
 of stars that are extremely valuable tracers of the large- and small-scale
 properties of mass around the Galaxy.
-Many thin stellar streams have been discovered around the Milky Way, the
-majority of which exist in the Galactic halo and are therefore primarily
-sensitive to the distribution of dark matter.
-Here we present astrometric data for one of the longest stellar streams --- the
-``GD-1'' stream --- from data release 2 (\DR) of the \gaia\ mission.
-We re-discover significant density variations along the stream by selecting
-candidate member stars first using only the exquisite \gaia\ data.
-Combined with filtering using Pan-STARRS photometry, we see clear evidence of
-multiple prominent under-densities and haps in the stream.
-These density variations --- and especially the gaps --- are possible
-indications of past interactions with large, massive perturbers.
-Given that the orbit of the stream crosses the Milky Way disk at large radius,
-these perturbers are likely halo objects and have masses consistent with XX,
-suggestive of dark matter sub-halos.
+Most of such streams discovered in the Milky Way reside in its halo, and are therefore primarily sensitive to the distribution of dark matter.
+% Many thin stellar streams have been discovered around the Milky Way, the
+% majority of which exist in the Galactic halo and are therefore primarily
+- cleanest sample, foreground removed w gaia parallaxes; members selected on combination of gaia pm and panstarrs colors
+- we thus got the crispest view of the stream to date, revealing in high resolution the rich structure which was only hinted at in the previous generation of data
+% We present astrometric data for one of the longest stellar streams --- the
+% GD-1 stream --- from the second data release (\DR) of the \gaia\ mission.
+% We re-discover significant density variations along the stream by selecting
+% candidate member stars using the exquisite \gaia\ data alone, including the
+% existence of a prominent, apparent gap in the stream.
+% <Using photometric data from the \pans, we refine our selection ...>
+These density variations --- and especially the gap --- are possible indications
+of past interactions with large, massive perturbers.
+<estimate the mass of interaction...>
 \end{abstract}
@@
 
-Dynamically cold stellar streams, likely formed from the tidal disruption of
-globular clusters, have been discovered throughout the Milky Way halo (e.g.,
-\citealt{TODO}).
+Dynamically cold stellar streams, likely formed from the tidal disruption of globular clusters, are sensitive to the large and small scale distribution of matter in the halo
 
+Thin stellar streams form from the tidal disruption of low-dispersion stellar systems such as globular clusters \citep[e.g.,][]{TODO}.
 To date, $\approx$13--14 thin streams have been discovered by searching large-area sky surveys such as the Sloan Digital Sky Survey (SDSS) and Pan-STARRS1 survey (PS1).
-These previously known thin streams span a wide range of lengths \citep[$\sim$$2$--$60$ deg;][]{bernard14,grillmair06} and masses \citep[$\sim$$10^3$--$10^5~\msun$;][]{TODO,TODO}, however, only two such streams have clear progenitor systems: NGC 5466 and Palomar 5 \citep{TODO,TODO}.
+These previously known thin streams span a wide range of lengths \citep[$\sim2$--$60$ deg;][]{bernard14,grillmair06} and masses \citep[$\sim10^3$--$10^5~\msun$;][]{TODO,TODO}, however, only two such streams have clear progenitor systems: NGC 5466 and Palomar 5 \citep{TODO,TODO}.
 Though it has been shown with many independent methods that modeling these thin streams can place strong constraints on the gravitational potential in which they form \citep{apw14,TODO}, many efforts to apply these models to real streams around the Milky Way have been hampered by the small number of known progenitor systems.
@@
 \label{sec:data}
-
-- Transform to GD-1 coordinates (Koposov et al.)
-    - Can see deviation from great circle
-- Cut to be close to latitude = 0
-- Cut to get proper motion over-density
-- Color-magnitude filtering
-
-Figure 1:
-- PM cut (panel a)
-- PM + CMD cut (panel b)
-- SFD (panel c)??
-
-Comment on these here - add figure for both in Appendix?
-- Dust / no significant extinction - combine with other figure
-- Scanning pattern - just discuss, no figure
 
@@
 \caption{%
-Colored background shows the \gaia\ $G$-band extinction, $A_G$, in the GD-1
+Colored background shows the $V$-band extinction, $A_V$, in the GD-1
 coordinate system.
-Vertical lines roughly show the regions identified as the under-density (UD) and
+Vertical lines roughly show the extents the identified under-density (UD) and
 gap.
@@
 \acknowledgements{
-thanks: Belokurov, Casey, Geha, Hogg, Johnston, Koposov, Schlafly, Spergel
+Gaia
+Belokurov, Casey, Geha, Hogg, Johnston, Koposov, Lisanti, Schlafly, Spergel
 This research was started at the NYC Gaia DR2 Workshop at the Center for Computational Astrophysics of the Flatiron Institute in 2018 April.

### 2018-04-28 paper/gd1.tex

@@
 <estimate the mass of interaction...>
+% 
+% Many thin stellar streams have been discovered around the Milky Way, the
+% majority of which exist in the Galactic halo and are therefore primarily
+% sensitive to the distribution of dark matter.
+% Here we present astrometric data for one of the longest stellar streams --- the
+% ``GD-1'' stream --- from data release 2 (\DR) of the \gaia\ mission.
+% We re-discover significant density variations along the stream by selecting
+% candidate member stars first using only the exquisite \gaia\ data.
+% Combined with filtering using Pan-STARRS photometry, we see clear evidence of
+% multiple prominent under-densities and haps in the stream.
+% These density variations --- and especially the gaps --- are possible
+% indications of past interactions with large, massive perturbers.
+% Given that the orbit of the stream crosses the Milky Way disk at large radius,
+% these perturbers are likely halo objects and would have to have masses XX,
+% suggestive of dark matter sub-halos.
 \end{abstract}
@@
 
-Dynamically cold stellar streams, likely formed from the tidal disruption of globular clusters, are sensitive to the large and small scale distribution of matter in the halo
+Dynamically cold stellar streams are formed from the tidal disruption of stellar
+systems by the gravitational field of their host galaxy (e.g.,
+\citealt{Johnston:1998}).
+Over 30 candidate stellar streams have been discovered throughout the halo of
+the Milky Way and have a wide range of properties such as length, velocity
+dispersion, and density (e.g., \citealt{SoManyPeople,Bonaca:2012}; see also the
+recent review \citealt{Grillmair:2006, Newberg:2016}).
+Of the thin, likely globular-cluster-origin streams, only two have clear
+progenitor systems: NGC 5466 (\citealt{TODO}) and Palomar 5 (\citealt{TODO}).
 
-Thin stellar streams form from the tidal disruption of low-dispersion stellar systems such as globular clusters \citep[e.g.,][]{TODO}.
-To date, $\approx$13--14 thin streams have been discovered by searching large-area sky surveys such as the Sloan Digital Sky Survey (SDSS) and Pan-STARRS1 survey (PS1).
-These previously known thin streams span a wide range of lengths \citep[$\sim2$--$60$ deg;][]{bernard14,grillmair06} and masses \citep[$\sim10^3$--$10^5~\msun$;][]{TODO,TODO}, however, only two such streams have clear progenitor systems: NGC 5466 and Palomar 5 \citep{TODO,TODO}.
+Stellar streams are
+
+\citealt{Bonaca:2018}
+
+Dynamical modeling of stellar streams or models of mean stream tracks have been used
+
 Though it has been shown with many independent methods that modeling these thin streams can place strong constraints on the gravitational potential in which they form \citep{apw14,TODO}, many efforts to apply these models to real streams around the Milky Way have been hampered by the small number of known progenitor systems.
+
+We revisit...do not try to model or fit for Galactic potential. Focus on stream properties, data.
 

### 2018-04-28 paper/gd1.tex
abstract edits

@@
 of stars that are extremely valuable tracers of the large- and small-scale
-properties of mass around the Galaxy.
+distribution of matter in the Galaxy.
 Most of such streams discovered in the Milky Way reside in its halo, and are therefore primarily sensitive to the distribution of dark matter.
-% Many thin stellar streams have been discovered around the Milky Way, the
-% majority of which exist in the Galactic halo and are therefore primarily
-- cleanest sample, foreground removed w gaia parallaxes; members selected on combination of gaia pm and panstarrs colors
-- we thus got the crispest view of the stream to date, revealing in high resolution the rich structure which was only hinted at in the previous generation of data
-% We present astrometric data for one of the longest stellar streams --- the
-% GD-1 stream --- from the second data release (\DR) of the \gaia\ mission.
-% We re-discover significant density variations along the stream by selecting
-% candidate member stars using the exquisite \gaia\ data alone, including the
-% existence of a prominent, apparent gap in the stream.
-% <Using photometric data from the \pans, we refine our selection ...>
-These density variations --- and especially the gap --- are possible indications
-of past interactions with large, massive perturbers.
-<estimate the mass of interaction...>
-% 
-% Many thin stellar streams have been discovered around the Milky Way, the
-% majority of which exist in the Galactic halo and are therefore primarily
-% sensitive to the distribution of dark matter.
-% Here we present astrometric data for one of the longest stellar streams --- the
-% ``GD-1'' stream --- from data release 2 (\DR) of the \gaia\ mission.
-% We re-discover significant density variations along the stream by selecting
-% candidate member stars first using only the exquisite \gaia\ data.
-% Combined with filtering using Pan-STARRS photometry, we see clear evidence of
-% multiple prominent under-densities and haps in the stream.
-% These density variations --- and especially the gaps --- are possible
-% indications of past interactions with large, massive perturbers.
-% Given that the orbit of the stream crosses the Milky Way disk at large radius,
-% these perturbers are likely halo objects and would have to have masses XX,
-% suggestive of dark matter sub-halos.
+We present a sample of highly-probable members of the longest cold stream, GD-1, selected on small parallaxes and retrograde proper motions provided by the \gaia\ mission, and \pans\ photometry consistent with an old and metal-poor population at a distance of $\sim10\,$kpc.
+This selection extended the stream by $10^\circ$(?), revealed a possible location of the progenitor and mapped significant density variations along the stream in unprecedented detail.
+In addition to several prominent gaps, for the first time we also find evidence of stream members off the main track.
+No(?) secular formation mechanisms predict the existence of such spurs, but they are a natural result of a stream interaction with a massive perturber.
+Based on the impulse approximation arguments, we estimate that the detected underdensity and the associated spur could have been induced by a perturbation of a $\sim10^7\rm M_\odot$(?) object.
+Our model of GD-1 indicates that the stream crosses the Milky Way disk at a large radius, which puts the likely origin of the perturber beyond the stellar disk.
 \end{abstract}

### 2018-04-28 paper/gd1.tex
intro

--- 2018-04-28 paper/gd1.tex
@@
 % - Reference for the dustmaps package? See software section.
+
+% Story:
+% - progenitor - show std_v, std_phi2
+% - 2 gaps, 1 under-density - epicycles/streakline? percentiles in polynomial
+% - spur - encounter?
 
@@
 \newcommand{\gaia}{\textsl{Gaia}}
-\newcommand{\pans}{\textsl{PanSTARRS}}
+\newcommand{\pans}{\textsl{Pan-STARRS}}
 \newcommand{\DR}{\acronym{DR2}}
 \newcommand{\msun}{\rm M_\odot}
+\newcommand{\article}{\textsl{Letter}}
 
@@
 \citealt{Johnston:1998}).
-Over 30 candidate stellar streams have been discovered throughout the halo of
-the Milky Way and have a wide range of properties such as length, velocity
+The phase-space density and mean track of stars in streams therefore encode
+information about the underlying distribution and shape of mass on galactic
+scales (\citealt{Bonaca:2018}).
+Most stellar streams are found in the halos of galaxies and are therefore
+important tracers of the large-scale distribution of dark matter in galactic
+halos.
+
+Streams formed from globular cluster systems are particularly useful as they can
+theoretically remain coherent for tens of orbital periods, depending in detail
+on the symmetries of the underlying galactic mass distribution
+(\citealt{Erkal:...}).
+This makes them excellent laboratories for studying small-scale structures in
+the mass distribution, as it is expected that density perturbations caused by
+interactions with massive perturbers can persist and remain visible for close to
+their survival time (e.g., \citealt{Yoon:2011}).
+In this sense, streams represent one of the most promising directions for
+studying or ruling out the existence of small-scale dark matter sub-halos, as
+would be predicted by standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
+(\citealt{Erkal, Sanders, Bovy} papers).
+
+While globular cluster streams are hard to detect around distant galaxies
+because of their low relative surface density, over 30 candidate stellar streams
+have been discovered throughout the halo of the Milky Way, largely thanks to
+large-area, multi-band photometric surveys (e.g., the Sloan Digital Sky Survey,
+SDSS; \citealt{York:2000, TODO}).
+The Milky Way streams have a wide range of properties such as length, velocity
 dispersion, and density (e.g., \citealt{SoManyPeople,Bonaca:2012}; see also the
-recent review \citealt{Grillmair:2006, Newberg:2016}).
-Of the thin, likely globular-cluster-origin streams, only two have clear
-progenitor systems: NGC 5466 (\citealt{TODO}) and Palomar 5 (\citealt{TODO}).
+recent review \citealt{Grillmair:2016, Newberg:2016}), and the majority have no
+obvious, bound progenitor systems; Of the thin, likely globular-cluster-origin
+streams, only two have clear progenitors (NGC 5466 and Palomar 5; \citealt{TODO,
+TODO}).
 
-Stellar streams are
+The most prominent and longest (in angle) thin stream is the GD-1 stream, named
+after its discoverers (\citealt{Grillmair:2006}).
+GD-1 was found by matched-filtering photometry from the SDSS and spans over
+$60^\circ$ in Heliocentric sky coordinates, partially because of its relative
+closeness to the sun ($\approx 7$--$10~\textrm{kpc}$).
+Despite its proximity, no remnant progenitor cluster or stellar system has been
+found associated with the stream.
+[Long, somewhat close: useful for constraining large-scale potential (\citealt{Koposov:2009})]
 
-\citealt{Bonaca:2018}
+[Small velocity dispersion (quote number?), likely old, retrograde orbit far
+from galactic disk, so prime place to search for relics of past subhalo
+interactions]
+[Known density wiggles and blah (\citealt{DeBoer:2017}), but color-magnitude only, bin stars]
 
-Dynamical modeling of stellar streams or models of mean stream tracks have been used
-
-Though it has been shown with many independent methods that modeling these thin streams can place strong constraints on the gravitational potential in which they form \citep{apw14,TODO}, many efforts to apply these models to real streams around the Milky Way have been hampered by the small number of known progenitor systems.
-
+In this \article, we present the cleanest view of the GD-1 stream using
+astrometric data from the \gaia\ mission data release 2 (\DR).
+Combined with photometry from \pans\ ...
 We revisit...do not try to model or fit for Galactic potential. Focus on stream properties, data.

### 2018-04-28 paper/gd1.tex

@@
 of stars that are extremely valuable tracers of the large- and small-scale
-properties of mass around the Galaxy.
+distribution of matter in the Galaxy.
 Most of such streams discovered in the Milky Way reside in its halo, and are therefore primarily sensitive to the distribution of dark matter.
-% Many thin stellar streams have been discovered around the Milky Way, the
-% majority of which exist in the Galactic halo and are therefore primarily
-- cleanest sample, foreground removed w gaia parallaxes; members selected on combination of gaia pm and panstarrs colors
-- we thus got the crispest view of the stream to date, revealing in high resolution the rich structure which was only hinted at in the previous generation of data
-% We present astrometric data for one of the longest stellar streams --- the
-% GD-1 stream --- from the second data release (\DR) of the \gaia\ mission.
-% We re-discover significant density variations along the stream by selecting
-% candidate member stars using the exquisite \gaia\ data alone, including the
-% existence of a prominent, apparent gap in the stream.
-% <Using photometric data from the \pans, we refine our selection ...>
-These density variations --- and especially the gap --- are possible indications
-of past interactions with large, massive perturbers.
-<estimate the mass of interaction...>
-%
-% Many thin stellar streams have been discovered around the Milky Way, the
-% majority of which exist in the Galactic halo and are therefore primarily
-% sensitive to the distribution of dark matter.
-% Here we present astrometric data for one of the longest stellar streams --- the
-% ``GD-1'' stream --- from data release 2 (\DR) of the \gaia\ mission.
-% We re-discover significant density variations along the stream by selecting
-% candidate member stars first using only the exquisite \gaia\ data.
-% Combined with filtering using Pan-STARRS photometry, we see clear evidence of
-% multiple prominent under-densities and haps in the stream.
-% These density variations --- and especially the gaps --- are possible
-% indications of past interactions with large, massive perturbers.
-% Given that the orbit of the stream crosses the Milky Way disk at large radius,
-% these perturbers are likely halo objects and would have to have masses XX,
-% suggestive of dark matter sub-halos.
+We present a sample of highly-probable members of the longest cold stream, GD-1, selected on small parallaxes and retrograde proper motions provided by the \gaia\ mission, and \pans\ photometry consistent with an old and metal-poor population at a distance of $\sim10\,$kpc.
+This selection extended the stream by $10^\circ$(?), revealed a possible location of the progenitor and mapped significant density variations along the stream in unprecedented detail.
+In addition to several prominent gaps, for the first time we also find evidence of stream members off the main track.
+No(?) secular formation mechanisms predict the existence of such spurs, but they are a natural result of a stream interaction with a massive perturber.
+Based on the impulse approximation arguments, we estimate that the detected underdensity and the associated spur could have been induced by a perturbation of a $\sim10^7\rm M_\odot$(?) object.
+Our model of GD-1 indicates that the stream crosses the Milky Way disk at a large radius, which puts the likely origin of the perturber beyond the stellar disk.
 \end{abstract}

### 2018-04-28 paper/gd1.tex
abstract tweaks

--- 2018-04-28 paper/gd1.tex
@@
 No(?) secular formation mechanisms predict the existence of such spurs, but they are a natural result of a stream interaction with a massive perturber.
-Based on the impulse approximation arguments, we estimate that the detected underdensity and the associated spur could have been induced by a perturbation of a $\sim10^7\rm M_\odot$(?) object.
-Our model of GD-1 indicates that the stream crosses the Milky Way disk at a large radius, which puts the likely origin of the perturber beyond the stellar disk.
+Based on the impulse approximation arguments, we estimate that the detected underdensity and the associated spur could have been induced by a perturbation of a $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
+Our model of GD-1 indicates that the stream crosses the Galactic plane at a large radius, which puts the likely origin of the perturber beyond the stellar disk.
 \end{abstract}

### 2018-04-28 paper/gd1.tex
Intro

--- 2018-04-28 paper/gd1.tex
@@
 \newcommand{\gaia}{\textsl{Gaia}}
-\newcommand{\pans}{\textsl{PanSTARRS}}
+\newcommand{\pans}{\textsl{Pan-STARRS}}
 \newcommand{\DR}{\acronym{DR2}}
-\newcommand{\msun}{\rm M_\odot}
+\newcommand{\msun}{\textrm{M}_\odot}
+\newcommand{\kpc}{\textrm{kpc}}
+\newcommand{\feh}{\ensuremath{[\textrm{Fe} / \textrm{H}]}}
+\newcommand{\article}{\textsl{Letter}}
 
@@
 Dynamically cold stellar streams are formed from the tidal disruption of stellar
-systems by the gravitational field of their host galaxy (e.g.,
-\citealt{Johnston:1998}).
+systems by the gravitational field of their host galaxy.
+The phase-space density and mean track of stars in streams therefore encode
+information about the underlying distribution and shape of mass on galactic
+scales (e.g., \citealt{Johnston:1999, Bonaca:2018}).
+Most stellar streams are found in the halos of galaxies and are therefore
+important tracers of the large-scale distribution of dark matter in galactic
+halos.
+
+Long, thin stellar streams are also excellent laboratories for studying
+small-scale structures in the mass distribution:
+thin streams can remain coherent for tens of orbital periods, depending in
+detail on the symmetries of the underlying galactic mass distribution (e.g.,
+\citealt{Erkal:2016}).
+Encounters between massive perturbers and stream stars imprint variations in the
+phase-space density that can persist and remain visible for close to their total
+survival time (e.g., \citealt{Yoon:2011}).
+In this sense, streams represent one of the most promising directions for
+studying or ruling out the existence of small-scale dark matter sub-halos, as
+would be predicted by standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
+(\citealt{Erkal:2015, Sanders:2016, Bovy:2017}).
+
 Over 30 candidate stellar streams have been discovered throughout the halo of
-the Milky Way and have a wide range of properties such as length, velocity
-dispersion, and density (e.g., \citealt{SoManyPeople,Bonaca:2012}; see also the
-recent review \citealt{Grillmair:2006, Newberg:2016}).
-Of the thin, likely globular-cluster-origin streams, only two have clear
-progenitor systems: NGC 5466 (\citealt{TODO}) and Palomar 5 (\citealt{TODO}).
+the Milky Way, largely thanks to large-area, multi-band photometric surveys
+(especially the Sloan Digital Sky Survey, SDSS; \citealt{York:2000}).
+The Milky Way streams have a wide range of properties such as length, velocity
+dispersion, and density (e.g., \citealt{Odenkirchen:2001, Grillmair:2006,
+Grillmair:2006b, Belokurov:2006, Belokurov:2007, Bonaca:2012, Shipp:2018}; see
+also the recent review \citealt{Grillmair:2016, Newberg:2016}), and the majority
+have no obvious, bound progenitor systems;
+of the thin, likely globular-cluster-origin streams, only two have clear
+progenitors (NGC 5466 and Palomar 5).
 
-Stellar streams are
+The most prominent and longest (in angle) thin stream is the GD-1 stream, named
+after its discoverers (\citealt{Grillmair:2006}).
+GD-1 was found using a matched filter on photometric data from the SDSS and
+spans over $60^\circ$ in Heliocentric sky coordinates, partially because of its
+relative closeness to the sun ($\approx 7$--$10~\textrm{kpc}$).
+Despite its proximity, no remnant progenitor cluster or stellar system has been
+found associated with the stream, but its metallicity ($\feh \approx -1.4$;
+\citealt{Koposov:2010}) and estimated stellar mass ($M \approx 2 \times
+10^4~\msun$; \citealt{Koposov:2010}) are consistent with being a disrupted
+globular cluster.
 
-\citealt{Bonaca:2018}
+The long physical length of the GD-1 stream ($\sim 15~\kpc$) and location in the
+Galactic halo (pericentric distance $r_\textrm{peri} \sim 13~\kpc$) make it an
+ideal object for constraining the gravitational potential of dark matter in the
+inner Milky Way halo.
+Indeed, both from fitting orbits to binned phase-space measurements along the
+GD-1 stream (\citealt{Koposov:2010}) and from modeling the stream track in
+phase-space coordinates (\citealt{Bovy:2016}), it has been shown that the dark
+matter distribution within its orbit is consistent with or nearly spherical.
 
-Dynamical modeling of stellar streams or models of mean stream tracks have been used
+Its long length --- and therefore its implied old dynamical age --- and its
+particular orbit also make it a prime stream to search for interactions with
+dark matter subhalos:
+because of its large pericentric distance and retrograde orbit with respect to
+the Galactic bar, density variations in the stream are not expected from either
+interactions with giant molecular clouds (\citealt{TODO}) or resonant
+interactions with the bar (\citealt{Pearson:2017}).
+It is particularly exciting, therefore, that deep photometry of the stream has
+revealed interesting small-scale ``wiggles'' and density variations along a
+$\approx 45^\citc$ span of the stream (\citealt{DeBoer:2017}).
+, but color-magnitude only, bin stars]
 
-Though it has been shown with many independent methods that modeling these thin streams can place strong constraints on the gravitational potential in which they form \citep{apw14,TODO}, many efforts to apply these models to real streams around the Milky Way have been hampered by the small number of known progenitor systems.
-
+In this \article, we present a the cleanest view of the GD-1 stream using
+astrometric data from the \gaia\ mission data release 2 (\DR).
+Combined with photometry from \pans\ ...
 We revisit...do not try to model or fit for Galactic potential. Focus on stream properties, data.
@@
 - Color-magnitude filtering
-
-Figure 1:
-- PM cut (panel a)
-- PM + CMD cut (panel b)
-- SFD (panel c)??
 

### 2018-04-28 paper/gd1.tex
intro refs

@@
 \shorttitle{GD-1 in Gaia DR2}
-\shortauthors{price-whelan \& bonaca}
+\shortauthors{Price-Whelan \& Bonaca}
 
@@
 
-\title{A first look at the GD-1 stellar stream with Gaia DR2}
+\title{Off the beaten path: \\
+       ...}
+% Gaia DR2 reveals stream stars...
+% A first look at the GD-1 stellar stream with Gaia DR2
 
@@
 \begin{abstract}\noindent % trust me
-Tidally-disrupted globular clusters are transformed into thin, dynamically-cold streams
-of stars that are extremely valuable tracers of the large- and small-scale
-distribution of matter in the Galaxy.
-Most of such streams discovered in the Milky Way reside in its halo, and are therefore primarily sensitive to the distribution of dark matter.
-We present a sample of highly-probable members of the longest cold stream, GD-1, selected on small parallaxes and retrograde proper motions provided by the \gaia\ mission, and \pans\ photometry consistent with an old and metal-poor population at a distance of $\sim10\,$kpc.
+Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
+streams of stars that are extremely valuable tracers of the large- and
+small-scale distribution of matter in the Galaxy.
+Most stellar streams discovered around the Milky Way reside in the Galactic halo
+and are therefore primarily sensitive to the distribution of dark matter.
+We present a sample of highly-probable members of the longest cold stream, GD-1,
+selected to have small parallaxes and retrograde proper motions using data from
+the \gaia\ mission, and \pans\ photometry consistent with an old and metal-poor
+population at a distance of $\sim10~\kpc$.
 This selection extended the stream by $10^\circ$(?), revealed a possible location of the progenitor and mapped significant density variations along the stream in unprecedented detail.
@@
 
-Its long length --- and therefore its implied old dynamical age --- and its
-particular orbit also make it a prime stream to search for interactions with
-dark matter subhalos:
+Its long length --- and therefore its implied old dynamical age --- combined
+with its particular orbit also make it a prime stream to search for interactions
+with dark matter subhalos:
 because of its large pericentric distance and retrograde orbit with respect to
 the Galactic bar, density variations in the stream are not expected from either
-interactions with giant molecular clouds (\citealt{TODO}) or resonant
-interactions with the bar (\citealt{Pearson:2017}).
-It is particularly exciting, therefore, that deep photometry of the stream has
-revealed interesting small-scale ``wiggles'' and density variations along a
-$\approx 45^\citc$ span of the stream (\citealt{DeBoer:2017}).
-, but color-magnitude only, bin stars]
+interactions with giant molecular clouds (\citealt{Amorisco:2016}) or resonant
+encounters with the bar (\citealt{Pearson:2017}).
+[Erkal 2016 predict ~1 gap]
 
-In this \article, we present a the cleanest view of the GD-1 stream using
-astrometric data from the \gaia\ mission data release 2 (\DR).
-Combined with photometry from \pans\ ...
-We revisit...do not try to model or fit for Galactic potential. Focus on stream properties, data.
+From SDSS photometry, hints of gaps and under-densities have been identified in
+the GD-1 stream, but with low significance (\citealt{Carlberg:2013}).
+More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
+interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\citc$ span of the stream (\citealt{DeBoer:2017}):
+[summarize their conclusions].
+
+In this \article, we present clearest view of the GD-1 stream to date using
+astrometric data from the \gaia\ mission data release 2 (\DR) combined with
+precise photometry from the \pans\ survey.
+We find XX under-densities gaps, some consistent with ...
+We revisit...do not try to model or fit for Galactic potential.
+Focus on stream properties, data.
 

### 2018-04-28 paper/gd1.tex
intro

@@
 Over 30 candidate stellar streams have been discovered throughout the halo of
-the Milky Way, largely thanks to large-area, multi-band photometric surveys
-(especially the Sloan Digital Sky Survey, SDSS; \citealt{York:2000}).
+the Milky Way, thanks to large-area, multi-band photometric surveys (especially
+the Sloan Digital Sky Survey, SDSS; \citealt{York:2000}).
 The Milky Way streams have a wide range of properties such as length, velocity
@@
 
-The most prominent and longest (in angle) thin stream is the GD-1 stream, named
-after its discoverers (\citealt{Grillmair:2006}).
+The most prominent and longest (in angular extent) thin stream is the GD-1
+stream, named after its discoverers (\citealt{Grillmair:2006}).
 GD-1 was found using a matched filter on photometric data from the SDSS and
@@
 ideal object for constraining the gravitational potential of dark matter in the
-inner Milky Way halo.
+inner Milky Way.
 Indeed, both from fitting orbits to binned phase-space measurements along the
@@
 encounters with the bar (\citealt{Pearson:2017}).
-[Erkal 2016 predict ~1 gap]
+By assuming an age [...], \citet{Erkal:2016} predict that XX (~1 gap??).
 
-From SDSS photometry, hints of gaps and under-densities have been identified in
-the GD-1 stream, but with low significance (\citealt{Carlberg:2013}).
+From SDSS photometry, hints of further gaps and under-densities have been
+identified in the GD-1 stream, but with low significance
+(\citealt{Carlberg:2013}).
 More recently, deep photometry of GD-1 from CFHT/Megacam has revealed

### 2018-04-28 paper/gd1.tex
Tweaks

@@
 The most prominent and longest (in angular extent) thin stream is the GD-1
-stream, named after its discoverers (\citealt{Grillmair:2006}).
+stream (named after its discoverers; \citealt{Grillmair:2006}).
 GD-1 was found using a matched filter on photometric data from the SDSS and
-spans over $60^\circ$ in Heliocentric sky coordinates, partially because of its
-relative closeness to the sun ($\approx 7$--$10~\textrm{kpc}$).
-Despite its proximity, no remnant progenitor cluster or stellar system has been
-found associated with the stream, but its metallicity ($\feh \approx -1.4$;
-\citealt{Koposov:2010}) and estimated stellar mass ($M \approx 2 \times
-10^4~\msun$; \citealt{Koposov:2010}) are consistent with being a disrupted
-globular cluster.
+was found to span $\approx 60^\circ$ in Heliocentric sky coordinates, partially
+because of its relative proximity to the sun ($\approx 7$--$10~\textrm{kpc}$).
+Despite being close and relatively high surface-density, no remnant progenitor
+cluster or stellar system has been found associated with the stream, but its
+width,  metallicity ($\feh \approx -1.4$; \citealt{Koposov:2010}) and estimated
+stellar mass ($M \approx 2 \times 10^4~\msun$; \citealt{Koposov:2010}) are
+consistent with being a disrupted globular cluster.
 
@@
 GD-1 stream (\citealt{Koposov:2010}) and from modeling the stream track in
-phase-space coordinates (\citealt{Bovy:2016}), it has been shown that the dark
-matter distribution within its orbit is consistent with or nearly spherical.
+phase-space coordinates (\citealt{Bovy:2016}), it has been used to show that the
+dark matter distribution within its orbit is consistent with being close to
+spherical.
 
@@
 encounters with the bar (\citealt{Pearson:2017}).
-By assuming an age [...], \citet{Erkal:2016} predict that XX (~1 gap??).
+By estimating a stream age of $\approx 4~\textrm{Gyr}$ assuming a progenitor
+mass of $10^5~\msun$, and by taking a subhalo mass function and density from
+XXX, \citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
+($\approx 5^\circ$--$7^\circ$) gap caused by an interaction with a
+$10^6$--$10^7~\msun$ subhalo.
 
-From SDSS photometry, hints of further gaps and under-densities have been
-identified in the GD-1 stream, but with low significance
+TODO: hints of density variations from initial work Koposov:2010, later confirmed by density modeling, but all low significance
 (\citealt{Carlberg:2013}).

### 2018-04-28 paper/gd1.tex
Altaffiliation

@@
 \title{Off the beaten path: \\
-       ...}
+       Clear density variations in the GD-1 stream and star perturbed off the
+       stream track}
 % Gaia DR2 reveals stream stars...
@@
 \author[0000-0003-0872-7098]{Adrian~M.~Price-Whelan}
+\altaffiliation{These authors contributed equally to this work.}
 \affiliation{Department of Astrophysical Sciences,
@@
 \author[0000-0002-7846-9787]{Ana Bonaca}
+\altaffiliation{These authors contributed equally to this work.}
 \affil{Harvard--Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA}
@@
 detail on the symmetries of the underlying galactic mass distribution (e.g.,
-\citealt{Erkal:2016}).
+\citealt{Erkal:2016a}).
 Encounters between massive perturbers and stream stars imprint variations in the
@@
 \label{sec:data}
+
+We use astrometric data from the \gaia\ mission (\citealt{Prusti:2016}), data
+release 2 (\citealt{Gaia-Collaboration:2018, Lindegren:2018}), and photometry
+from the \pans\ survey, data release 1 (\citealt{Chambers:2016}).
+We retrieve data along the previously-identified track of the GD-1 stream and
+the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
+selecting all sources with small parallax, $\varpi < 1~\textrm{mas}$, and use
+the spherical polygon functionality to select batches of stars that have small
+latitude, $\phi_2$, in the GD-1 coordinate system defined in
+\cite{Koposov:2010}.\footnote{This transformation is implemented using the
+\package{Astropy} (\citealt{astropy}) coordinate transformation framework in the
+\package{Gala} \texttt{Python} package (\citealt{gala}).}
 

### 2018-04-29 paper/gd1.tex
crazy title
data section

--- 2018-04-28 paper/gd1.tex
@@
 \newcommand{\kpc}{\textrm{kpc}}
+\newcommand{\kms}{\ensuremath{\textrm{km}~\textrm{s}^{-1}}}
+\newcommand{\bs}[1]{\boldsymbol{#1}}
+\newcommand{\masyr}{\ensuremath{\textrm{mas}~\textrm{yr}^{-1}}}
 \newcommand{\feh}{\ensuremath{[\textrm{Fe} / \textrm{H}]}}
@@
 \title{Off the beaten path: \\
-       ...}
+       Gaia reveals GD-1 stars perturbed off of the main stream track}
 % Gaia DR2 reveals stream stars...
@@
 
-- Transform to GD-1 coordinates (Koposov et al.)
-    - Can see deviation from great circle
-- Cut to be close to latitude = 0
-- Cut to get proper motion over-density
+We use astrometric data from the \gaia\ mission (\citealt{Prusti:2016}), data
+release 2 (\citealt{Gaia-Collaboration:2018, Lindegren:2018}), and photometry
+from the \pans\ survey, data release 1 (\citealt{Chambers:2016}).
+
+We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
+from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
+selecting all sources with small parallax, $\varpi < 1~\textrm{mas}$, and use a
+set of spherical polygons to select batches of stars that have small latitude in
+the heliocentric GD-1 coordinate system, $(\phi_1, \phi_2)$, defined in
+\cite{Koposov:2010}.\footnote{This transformation is implemented using the
+\package{Astropy} (\citealt{astropy}) coordinate transformation framework in the
+\package{Gala} \texttt{Python} package (\citealt{gala}).}
+We convert the sky coordinates and proper motions of the parallax-selected stars
+to the GD-1 stream coordinate system, and correct the proper motions for solar
+reflex motion by assuming that stars at a given stream longitude, $\phi_1$, have
+a distance given by the linear relationship $d(\phi_1) = (0.05 \, \phi_1 +
+10)~\textrm{kpc}$.
+For the solar velocity in a Galactocentric rest frame, we use $\bs{v}_\odot =
+(11.1, 232.24, 7.25)~\kms$ (\citealt{Schonrich:2010, Bovy:2015}).
+\figurename~\ref{fig:XX} (top right) shows the distribution of stars with
+$|\phi_2| < XX^\circ$ in solar-reflex-corrected proper motion components in the
+GD-1 coordinate system, $\mu_{\phi_1, \odot}$ (which includes the $\cos{\phi_2}$
+term) and $\mu_{\phi_2, \odot}$.
+GD-1 member stars are visible as the over-density at approximately
+$(\mu_{\phi_1, \odot}, \mu_{\phi_2, \odot}) \approx (XX, YY)~\masyr$.
+
+As a first selection of GD-1 candidate member stars, we select stars with proper
+motions $XX < \mu_{\phi_1, \odot} < XX~\masyr$ and $XX < \mu_{\phi_2, \odot} <
+XX~\masyr$, as visualized by the rectangle in the top right panel of
+\figurename~\ref{fig:XX}.
+\figurename~\ref{fig:XX} (top left) shows all stars the pass the selection on
+proper motion, plotted in the GD-1 coordinate system.
+From kinematic selection alone, the stream is identifiable as the over-density
+of stars around $\phi_2 = 0$ between longitudes $XX^\circ \lesssim \phi_1
+\lesssim XX^\circ$.
+
+To improve the contrast of the stream over the background, we next cross-match
+the sample to the \pans\ photometric catalog and use the photometry to further
+clean the sample.
+\figurename~\ref{fig:XX} (bottom right) shows the distribution of
+proper-motion-selected candidate stream stars in a de-reddened \pans\
+color-magnitude diagram.
+The over-plotted isochrone is from the MESA Isochrones & Stellar Tracks (MIST;
+\citealt{XX}) and represents a $12~\textrm{Gyr}$ old population with $\feh =
+-1.XX$ at a distance of $XX~\kpc$.
+
 - Color-magnitude filtering

### 2018-04-29 paper/gd1.tex
Figure placeholder

@@
 \newcommand{\article}{\textsl{Letter}}
+
+\newcommand{\todo}[1]{{\color{red} TODO: #1}}
 
@@
 XXX, \citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
-($\approx 5^\circ$--$7^\circ$) gap caused by an interaction with a
+($\approx 5$--$7^\circ$) gap caused by an interaction with a
 $10^6$--$10^7~\msun$ subhalo.
@@
 release 2 (\citealt{Gaia-Collaboration:2018, Lindegren:2018}), and photometry
-from the \pans\ survey, data release 1 (\citealt{Chambers:2016}).
+from the \pans\ survey, data release 1 (\citealt{Chambers:2016}) to select
+high-confidence members of the GD-1 stream.
+
+\subselection{Stream member selection}
 
@@
 (11.1, 232.24, 7.25)~\kms$ (\citealt{Schonrich:2010, Bovy:2015}).
-\figurename~\ref{fig:XX} (top right) shows the distribution of stars with
+\figurename~\ref{fig:selection} (top right) shows the distribution of stars with
 $|\phi_2| < XX^\circ$ in solar-reflex-corrected proper motion components in the
@@
 XX~\masyr$, as visualized by the rectangle in the top right panel of
-\figurename~\ref{fig:XX}.
-\figurename~\ref{fig:XX} (top left) shows all stars the pass the selection on
-proper motion, plotted in the GD-1 coordinate system.
+\figurename~\ref{fig:selection}.
+\figurename~\ref{fig:selection} (top left) shows all stars the pass the
+selection on proper motion, plotted in the GD-1 coordinate system.
 From kinematic selection alone, the stream is identifiable as the over-density
@@
 
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=\textwidth]{stream-selection.pdf}
+\end{center}
+\caption{
+    \todo{AB}
+\label{fig:selection}
+}
+\end{figure}
+
 To improve the contrast of the stream over the background, we next cross-match
@@
 clean the sample.
-\figurename~\ref{fig:XX} (bottom right) shows the distribution of
+\figurename~\ref{fig:selection} (bottom right) shows the distribution of
 proper-motion-selected candidate stream stars in a de-reddened \pans\
 color-magnitude diagram.
+\todo{Did you also cut on $phi_2$?}.
 The over-plotted isochrone is from the MESA Isochrones & Stellar Tracks (MIST;
@@
 -1.XX$ at a distance of $XX~\kpc$.
+We use this isochrone to motivate a polygonal selection in de-reddened $g-i$
+color and apparent $g$-band magnitude, as visualized by the \todo{shaded} region
+in the bottom right panel of \figurename~\ref{fig:selection}.
 
-- Color-magnitude filtering
+\figurename~\ref{fig:selection} (bottom left) shows the final sample of GD-1
+stream member candidates after selecting on both proper motions and photometry.
+\todo{Comment on how this is the bomb ass diggity and the clearest view of the
+stream so far}
+
+\subselection{Stream member selection}
 

### 2018-04-29 paper/gd1.tex
mesa MIST refs

@@
 
-TODO: hints of density variations from initial work Koposov:2010, later confirmed by density modeling, but all low significance
+\todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
 (\citealt{Carlberg:2013}).
@@
 \todo{Did you also cut on $phi_2$?}.
-The over-plotted isochrone is from the MESA Isochrones & Stellar Tracks (MIST;
-\citealt{XX}) and represents a $12~\textrm{Gyr}$ old population with $\feh =
--1.XX$ at a distance of $XX~\kpc$.
+The over-plotted isochrone is from the MESA Isochrones \& Stellar Tracks (MIST;
+\citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
+$12~\textrm{Gyr}$ old population with $\feh = -1.XX$ at a distance of $XX~\kpc$.
 We use this isochrone to motivate a polygonal selection in de-reddened $g-i$

### 2018-04-29 paper/gd1.tex
sfd
Dust

@@
 \newcommand{\kpc}{\textrm{kpc}}
+\newcommand{\mag}{\textrm{mag}}
 \newcommand{\kms}{\ensuremath{\textrm{km}~\textrm{s}^{-1}}}
@@
 
-\subselection{Stream member selection}
-
 We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
@@
 stream so far}
+\todo{Mention the under-densities}
+\todo{Already mention the spur here}
 
-\subselection{Stream member selection}
-
-Comment on these here - add figure for both in Appendix?
-- Dust / no significant extinction - combine with other figure
-- Scanning pattern - just discuss, no figure
-
-Most of the clearly identifiable portions of the GD-1 stream are located at high
-Galactic latitudes ($b > 35^\circ$), and we therefore do not expect significant
+All of the clearly identifiable portions of the GD-1 stream are located at high
+Galactic latitudes ($b > 20^\circ$), and we therefore do not expect significant
 dust extinction or variations in extinction along the stream.
@@
 (\cite{Schlegel:1998}; hereafter SFD).
-Vertical lines indicate the two regions clearly identifiable as under-densities
-from the proper-motion-selected stream members.
-The maximum $A_V$ is $\approx$0.07 mag in both the under-density and gap
-regions.
-The dispersion in $A_V$ in the region $-60^\circ < \phi_1 < 0^\circ$ and
-$-1^\circ < \phi_1 < 1^\circ$ is $\sigma_{A_V} \approx 0.03~\textrm{mag}$.
+For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
+1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
+0.04~\mag$, and the dispersion in $A_V$ in this region is $\sigma_{A_V} \approx
+0.03~\textrm{mag}$.
+For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
+dust extinction becomes more appreciable.
 
-% Notebook: GD1-dust-and-completeness
+Another \todo{segue to this paragraph...}.
+However, we find that \todo{...}.
+We discuss the scanning pattern in more detail in
+Appendix~\ref{sec:completeness}.
+
 \begin{figure}[h]
@@
 \appendix
-\section{Completeness check and data validation}
-\label{sec:validate}
+\section{Completeness check and scanning pattern}
+\label{sec:completeness}
 

### 2018-04-29 paper/gd1.tex
Ack

@@
 proper-motion-selected candidate stream stars in a de-reddened \pans\
-color-magnitude diagram.
+color-magnitude diagram (CMD).
 \todo{Did you also cut on $phi_2$?}.
@@
 dust extinction or variations in extinction along the stream.
-\figurename~\ref{fig:sfd} shows the $V$-band extinction in the region around the
-GD-1 stream, computed from the Schlegel-Finkbeiner-Davis extinction map
-(\cite{Schlegel:1998}; hereafter SFD).
+\figurename~\ref{fig:sfd-cmd} (top panel) shows the $V$-band extinction in the
+region around the GD-1 stream, computed from the Schlegel-Finkbeiner-Davis
+extinction map (\cite{Schlegel:1998}; hereafter SFD).
 For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
@@
 
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=\textwidth]{sfd.pdf}
-\end{center}
-\caption{%
-Colored background shows the $V$-band extinction, $A_V$, in the GD-1
-coordinate system.
-Vertical lines roughly show the extents the identified under-density (UD) and
-gap.
-The maximum extinction in the UD or Gap regions is $\approx$0.07 mag.
-\label{fig:sfd}
-}
-\end{figure}
-
 \section{Results}
 \label{sec:results}
+
+\todo{segue sentences...}
 
@@
 
-\subsection{Gap}
+\figurename~\ref{fig:sfd-cmd} (middle panel) again shows the proper-motion and CMD selected stream stars, with interesting features
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=\textwidth]{sfd-cmd.pdf}
+\end{center}
+\caption{%
+\todo{AB}
+\label{fig:sfd-cmd}
+}
+\end{figure}
+
+\subsection{Gaps and under-densities}
 \label{sec:res_gap}
 
-\subsection{Underdensity}
-\label{sec:res_underdensity}
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=\textwidth]{track-and-model.pdf}
+\end{center}
+\caption{%
+\todo{AB}
+\label{fig:track-and-model}
+}
+\end{figure}
+
+\subsection{Spur}
+\label{sec:res_spur}
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=\textwidth]{spur.pdf}
+\end{center}
+\caption{%
+\todo{AB}
+\label{fig:spur}
+}
+\end{figure}
 
@@
 
+\section{Conclusions}
+\label{sec:conclusions}
+
 \acknowledgements{
-Gaia
-Belokurov, Casey, Geha, Hogg, Johnston, Koposov, Lisanti, Schlafly, Spergel
-This research was started at the NYC Gaia DR2 Workshop at the Center for Computational Astrophysics of the Flatiron Institute in 2018 April.
-AB acknowledges generous support from the Institute for Theory and Computation at Harvard University.
-All code used in this work and all results are available at \url{https://github.com/adrn/GD1-DR2}.
+It is a pleasure to thank
+Vasily Belokurov,
+Andrew R. Casey,
+Marla Geha,
+David W. Hogg,
+Kathryn V. Johnston,
+Sergey Koposov,
+Mariangela Lisanti,
+Edward Schlafly,
+and David N. Spergel for useful discussions and feedback.
+
+This work has made use of data from the European Space Agency (ESA) mission {\it
+Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia} Data
+Processing and Analysis Consortium (DPAC,
+\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
+has been provided by national institutions, in particular the institutions
+participating in the {\it Gaia} Multilateral Agreement.  This research was
+started at the NYC Gaia DR2 Workshop at the Center for Computational
+Astrophysics of the Flatiron Institute in 2018 April.
+
+AB acknowledges generous support from the Institute for Theory and Computation
+at Harvard University.
+All code used in this work and all results are available at
+\url{https://github.com/adrn/GD1-DR2}.
 }
@@
 \appendix
-\section{Completeness check and scanning pattern}
+\section{Completeness and the \gaia\ scanning pattern}
 \label{sec:completeness}

### 2018-04-29 paper/gd1.tex
ack adds

--- 2018-04-29 paper/gd1.tex
@@
 \newcommand{\DR}{\acronym{DR2}}
-\newcommand{\msun}{\textrm{M}_\odot}
-\newcommand{\kpc}{\textrm{kpc}}
-\newcommand{\kms}{\ensuremath{\textrm{km}~\textrm{s}^{-1}}}
-\newcommand{\bs}[1]{\boldsymbol{#1}}
-\newcommand{\masyr}{\ensuremath{\textrm{mas}~\textrm{yr}^{-1}}}
-\newcommand{\feh}{\ensuremath{[\textrm{Fe} / \textrm{H}]}}
+\newcommand{\msun}{\rm M_\odot}
 \newcommand{\article}{\textsl{Letter}}
-
-\newcommand{\todo}[1]{{\color{red} TODO: #1}}
 
@@
 \shorttitle{GD-1 in Gaia DR2}
-\shortauthors{Price-Whelan \& Bonaca}
+\shortauthors{price-whelan \& bonaca}
 
@@
 
-\title{Off the beaten path: \\
-       Gaia reveals GD-1 stars perturbed off of the main stream track}
-% Gaia DR2 reveals stream stars...
-% A first look at the GD-1 stellar stream with Gaia DR2
+\title{A first look at the GD-1 stellar stream with Gaia DR2}
 
 \author[0000-0003-0872-7098]{Adrian~M.~Price-Whelan}
-\altaffiliation{These authors contributed equally to this work.}
 \affiliation{Department of Astrophysical Sciences,
@@
 \author[0000-0002-7846-9787]{Ana Bonaca}
-\altaffiliation{These authors contributed equally to this work.}
 \affil{Harvard--Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA}
@@
 \begin{abstract}\noindent % trust me
-Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
-streams of stars that are extremely valuable tracers of the large- and
-small-scale distribution of matter in the Galaxy.
-Most stellar streams discovered around the Milky Way reside in the Galactic halo
-and are therefore primarily sensitive to the distribution of dark matter.
-We present a sample of highly-probable members of the longest cold stream, GD-1,
-selected to have small parallaxes and retrograde proper motions using data from
-the \gaia\ mission, and \pans\ photometry consistent with an old and metal-poor
-population at a distance of $\sim10~\kpc$.
-This selection extended the stream by $10^\circ$(?), revealed a possible location of the progenitor and mapped significant density variations along the stream in unprecedented detail.
+Tidally-disrupted globular clusters are transformed into thin, dynamically-cold streams
+of stars that are extremely valuable tracers of the large- and small-scale
+distribution of matter in the Galaxy.
+Most of such streams discovered in the Milky Way reside in its halo, and are therefore primarily sensitive to the distribution of dark matter.
+We present a sample of highly-probable members of the longest cold stream, GD-1, selected on small parallaxes and retrograde proper motions provided by the \gaia\ mission, and \pans\ photometry consistent with an old and metal-poor population at a distance of $\sim10\,$kpc.
+This selection extended the stream by $25^\circ$, revealed a possible location of the progenitor and mapped significant density variations along the stream in unprecedented detail.
 In addition to several prominent gaps, for the first time we also find evidence of stream members off the main track.
 No(?) secular formation mechanisms predict the existence of such spurs, but they are a natural result of a stream interaction with a massive perturber.
-Based on the impulse approximation arguments, we estimate that the detected underdensity and the associated spur could have been induced by a perturbation of a $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
+Using the impulse approximation arguments, we estimate that the detected underdensity and the associated spur could have been induced by a perturbation of a $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
 Our model of GD-1 indicates that the stream crosses the Galactic plane at a large radius, which puts the likely origin of the perturber beyond the stellar disk.
@@
 Dynamically cold stellar streams are formed from the tidal disruption of stellar
-systems by the gravitational field of their host galaxy.
+systems by the gravitational field of their host galaxy (e.g.,
+\citealt{Johnston:1998}).
 The phase-space density and mean track of stars in streams therefore encode
 information about the underlying distribution and shape of mass on galactic
-scales (e.g., \citealt{Johnston:1999, Bonaca:2018}).
+scales (\citealt{Bonaca:2018}).
 Most stellar streams are found in the halos of galaxies and are therefore
@@
 
-Long, thin stellar streams are also excellent laboratories for studying
-small-scale structures in the mass distribution:
-thin streams can remain coherent for tens of orbital periods, depending in
-detail on the symmetries of the underlying galactic mass distribution (e.g.,
-\citealt{Erkal:2016a}).
-Encounters between massive perturbers and stream stars imprint variations in the
-phase-space density that can persist and remain visible for close to their total
-survival time (e.g., \citealt{Yoon:2011}).
+Streams formed from globular cluster systems are particularly useful as they can
+theoretically remain coherent for tens of orbital periods, depending in detail
+on the symmetries of the underlying galactic mass distribution
+(\citealt{Erkal:...}).
+This makes them excellent laboratories for studying small-scale structures in
+the mass distribution, as it is expected that density perturbations caused by
+interactions with massive perturbers can persist and remain visible for close to
+their survival time (e.g., \citealt{Yoon:2011}).
 In this sense, streams represent one of the most promising directions for
@@
 would be predicted by standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
-(\citealt{Erkal:2015, Sanders:2016, Bovy:2017}).
+(\citealt{Erkal, Sanders, Bovy} papers).
 
-Over 30 candidate stellar streams have been discovered throughout the halo of
-the Milky Way, thanks to large-area, multi-band photometric surveys (especially
-the Sloan Digital Sky Survey, SDSS; \citealt{York:2000}).
+While globular cluster streams are hard to detect around distant galaxies
+because of their low relative surface density, over 30 candidate stellar streams
+have been discovered throughout the halo of the Milky Way, largely thanks to
+large-area, multi-band photometric surveys (e.g., the Sloan Digital Sky Survey,
+SDSS; \citealt{York:2000, TODO}).
 The Milky Way streams have a wide range of properties such as length, velocity
-dispersion, and density (e.g., \citealt{Odenkirchen:2001, Grillmair:2006,
-Grillmair:2006b, Belokurov:2006, Belokurov:2007, Bonaca:2012, Shipp:2018}; see
-also the recent review \citealt{Grillmair:2016, Newberg:2016}), and the majority
-have no obvious, bound progenitor systems;
-of the thin, likely globular-cluster-origin streams, only two have clear
-progenitors (NGC 5466 and Palomar 5).
+dispersion, and density (e.g., \citealt{SoManyPeople,Bonaca:2012}; see also the
+recent review \citealt{Grillmair:2016, Newberg:2016}), and the majority have no
+obvious, bound progenitor systems; Of the thin, likely globular-cluster-origin
+streams, only two have clear progenitors (NGC 5466 and Palomar 5; \citealt{TODO,
+TODO}).
 
-The most prominent and longest (in angular extent) thin stream is the GD-1
-stream (named after its discoverers; \citealt{Grillmair:2006}).
-GD-1 was found using a matched filter on photometric data from the SDSS and
-was found to span $\approx 60^\circ$ in Heliocentric sky coordinates, partially
-because of its relative proximity to the sun ($\approx 7$--$10~\textrm{kpc}$).
-Despite being close and relatively high surface-density, no remnant progenitor
-cluster or stellar system has been found associated with the stream, but its
-width,  metallicity ($\feh \approx -1.4$; \citealt{Koposov:2010}) and estimated
-stellar mass ($M \approx 2 \times 10^4~\msun$; \citealt{Koposov:2010}) are
-consistent with being a disrupted globular cluster.
+The most prominent and longest (in angle) thin stream is the GD-1 stream, named
+after its discoverers (\citealt{Grillmair:2006}).
+GD-1 was found by matched-filtering photometry from the SDSS and spans over
+$60^\circ$ in Heliocentric sky coordinates, partially because of its relative
+closeness to the sun ($\approx 7$--$10~\textrm{kpc}$).
+Despite its proximity, no remnant progenitor cluster or stellar system has been
+found associated with the stream.
+[Long, somewhat close: useful for constraining large-scale potential (\citealt{Koposov:2009})]
 
-The long physical length of the GD-1 stream ($\sim 15~\kpc$) and location in the
-Galactic halo (pericentric distance $r_\textrm{peri} \sim 13~\kpc$) make it an
-ideal object for constraining the gravitational potential of dark matter in the
-inner Milky Way.
-Indeed, both from fitting orbits to binned phase-space measurements along the
-GD-1 stream (\citealt{Koposov:2010}) and from modeling the stream track in
-phase-space coordinates (\citealt{Bovy:2016}), it has been used to show that the
-dark matter distribution within its orbit is consistent with being close to
-spherical.
+[Small velocity dispersion (quote number?), likely old, retrograde orbit far
+from galactic disk, so prime place to search for relics of past subhalo
+interactions]
+[Known density wiggles and blah (\citealt{DeBoer:2017}), but color-magnitude only, bin stars]
 
-Its long length --- and therefore its implied old dynamical age --- combined
-with its particular orbit also make it a prime stream to search for interactions
-with dark matter subhalos:
-because of its large pericentric distance and retrograde orbit with respect to
-the Galactic bar, density variations in the stream are not expected from either
-interactions with giant molecular clouds (\citealt{Amorisco:2016}) or resonant
-encounters with the bar (\citealt{Pearson:2017}).
-By estimating a stream age of $\approx 4~\textrm{Gyr}$ assuming a progenitor
-mass of $10^5~\msun$, and by taking a subhalo mass function and density from
-XXX, \citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
-($\approx 5$--$7^\circ$) gap caused by an interaction with a
-$10^6$--$10^7~\msun$ subhalo.
-
-\todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
-(\citealt{Carlberg:2013}).
-More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
-interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\citc$ span of the stream (\citealt{DeBoer:2017}):
-[summarize their conclusions].
-
-In this \article, we present clearest view of the GD-1 stream to date using
-astrometric data from the \gaia\ mission data release 2 (\DR) combined with
-precise photometry from the \pans\ survey.
-We find XX under-densities gaps, some consistent with ...
-We revisit...do not try to model or fit for Galactic potential.
-Focus on stream properties, data.
+In this \article, we present the cleanest view of the GD-1 stream using
+astrometric data from the \gaia\ mission data release 2 (\DR).
+Combined with photometry from \pans\ ...
+We revisit...do not try to model or fit for Galactic potential. Focus on stream properties, data.
 
@@
 
-We use astrometric data from the \gaia\ mission (\citealt{Prusti:2016}), data
-release 2 (\citealt{Gaia-Collaboration:2018, Lindegren:2018}), and photometry
-from the \pans\ survey, data release 1 (\citealt{Chambers:2016}) to select
-high-confidence members of the GD-1 stream.
+- Transform to GD-1 coordinates (Koposov et al.)
+    - Can see deviation from great circle
+- Cut to be close to latitude = 0
+- Cut to get proper motion over-density
+- Color-magnitude filtering
 
-\subselection{Stream member selection}
-
-We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
-from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
-selecting all sources with small parallax, $\varpi < 1~\textrm{mas}$, and use a
-set of spherical polygons to select batches of stars that have small latitude in
-the heliocentric GD-1 coordinate system, $(\phi_1, \phi_2)$, defined in
-\cite{Koposov:2010}.\footnote{This transformation is implemented using the
-\package{Astropy} (\citealt{astropy}) coordinate transformation framework in the
-\package{Gala} \texttt{Python} package (\citealt{gala}).}
-We convert the sky coordinates and proper motions of the parallax-selected stars
-to the GD-1 stream coordinate system, and correct the proper motions for solar
-reflex motion by assuming that stars at a given stream longitude, $\phi_1$, have
-a distance given by the linear relationship $d(\phi_1) = (0.05 \, \phi_1 +
-10)~\textrm{kpc}$.
-For the solar velocity in a Galactocentric rest frame, we use $\bs{v}_\odot =
-(11.1, 232.24, 7.25)~\kms$ (\citealt{Schonrich:2010, Bovy:2015}).
-\figurename~\ref{fig:selection} (top right) shows the distribution of stars with
-$|\phi_2| < XX^\circ$ in solar-reflex-corrected proper motion components in the
-GD-1 coordinate system, $\mu_{\phi_1, \odot}$ (which includes the $\cos{\phi_2}$
-term) and $\mu_{\phi_2, \odot}$.
-GD-1 member stars are visible as the over-density at approximately
-$(\mu_{\phi_1, \odot}, \mu_{\phi_2, \odot}) \approx (XX, YY)~\masyr$.
-
-As a first selection of GD-1 candidate member stars, we select stars with proper
-motions $XX < \mu_{\phi_1, \odot} < XX~\masyr$ and $XX < \mu_{\phi_2, \odot} <
-XX~\masyr$, as visualized by the rectangle in the top right panel of
-\figurename~\ref{fig:selection}.
-\figurename~\ref{fig:selection} (top left) shows all stars the pass the
-selection on proper motion, plotted in the GD-1 coordinate system.
-From kinematic selection alone, the stream is identifiable as the over-density
-of stars around $\phi_2 = 0$ between longitudes $XX^\circ \lesssim \phi_1
-\lesssim XX^\circ$.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=\textwidth]{stream-selection.pdf}
-\end{center}
-\caption{
-    \todo{AB}
-\label{fig:selection}
-}
-\end{figure}
-
-To improve the contrast of the stream over the background, we next cross-match
-the sample to the \pans\ photometric catalog and use the photometry to further
-clean the sample.
-\figurename~\ref{fig:selection} (bottom right) shows the distribution of
-proper-motion-selected candidate stream stars in a de-reddened \pans\
-color-magnitude diagram.
-\todo{Did you also cut on $phi_2$?}.
-The over-plotted isochrone is from the MESA Isochrones \& Stellar Tracks (MIST;
-\citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
-$12~\textrm{Gyr}$ old population with $\feh = -1.XX$ at a distance of $XX~\kpc$.
-We use this isochrone to motivate a polygonal selection in de-reddened $g-i$
-color and apparent $g$-band magnitude, as visualized by the \todo{shaded} region
-in the bottom right panel of \figurename~\ref{fig:selection}.
-
-\figurename~\ref{fig:selection} (bottom left) shows the final sample of GD-1
-stream member candidates after selecting on both proper motions and photometry.
-\todo{Comment on how this is the bomb ass diggity and the clearest view of the
-stream so far}
-
-\subselection{Stream member selection}
+Figure 1:
+- PM cut (panel a)
+- PM + CMD cut (panel b)
+- SFD (panel c)??
 
@@
 Gaia
-Belokurov, Casey, Geha, Hogg, Johnston, Koposov, Lisanti, Schlafly, Spergel
+Vasily Belokurov, Andy Casey, Marla Geha, David Hogg, Ben Johnson, Kathryn Johnston, Sergey Koposov, Mariangela Lisanti, Eddie Schlafly, David Spergel
 This research was started at the NYC Gaia DR2 Workshop at the Center for Computational Astrophysics of the Flatiron Institute in 2018 April.

### 2018-04-29 paper/gd1.tex

--- 2018-04-29 paper/gd1.tex
@@
 % Copyright 2017 the authors. All rights reserved.
-
-% # style notes
-% - it is Cram\'er--Rao not Cram\'er-Rao. And yet Fisher-matrix not Fisher--matrix.
 
@@
 \newcommand{\kpc}{\textrm{kpc}}
-\newcommand{\mag}{\textrm{mag}}
+% \newcommand{\mag}{\textrm{mag}}
 \newcommand{\kms}{\ensuremath{\textrm{km}~\textrm{s}^{-1}}}
@@
 \begin{abstract}\noindent % trust me
-Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
-streams of stars that are extremely valuable tracers of the large- and
-small-scale distribution of matter in the Galaxy.
+Tidally-disrupted globular clusters are transformed into thin, dynamically-cold streams
+of stars that are extremely valuable tracers of the large- and small-scale
+distribution of matter in the Galaxy.
 Most stellar streams discovered around the Milky Way reside in the Galactic halo
@@
 population at a distance of $\sim10~\kpc$.
-This selection extended the stream by $10^\circ$(?), revealed a possible location of the progenitor and mapped significant density variations along the stream in unprecedented detail.
+This selection extended the stream by $20^\circ$, revealed a possible location of the progenitor and mapped significant density variations along the stream in unprecedented detail.
 In addition to several prominent gaps, for the first time we also find evidence of stream members off the main track.
 No(?) secular formation mechanisms predict the existence of such spurs, but they are a natural result of a stream interaction with a massive perturber.
-Based on the impulse approximation arguments, we estimate that the detected underdensity and the associated spur could have been induced by a perturbation of a $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
+Using the impulse approximation arguments, we estimate that the detected underdensity and the associated spur could have been induced by a perturbation of a $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
 Our model of GD-1 indicates that the stream crosses the Galactic plane at a large radius, which puts the likely origin of the perturber beyond the stellar disk.
@@
 More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
-interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\citc$ span of the stream (\citealt{DeBoer:2017}):
+interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\circ$ span of the stream (\citealt{DeBoer:2017}):
 [summarize their conclusions].
@@
 \begin{center}
-\includegraphics[width=\textwidth]{stream-selection.pdf}
+\includegraphics[width=\textwidth]{gd1_sample.pdf}
 \end{center}
@@
 1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
-0.04~\mag$, and the dispersion in $A_V$ in this region is $\sigma_{A_V} \approx
+0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V} \approx
 0.03~\textrm{mag}$.
@@
 \begin{center}
-\includegraphics[width=\textwidth]{sfd-cmd.pdf}
+% \includegraphics[width=\textwidth]{sfd-cmd.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{track-and-model.pdf}
+% \includegraphics[width=\textwidth]{track-and-model.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{spur.pdf}
+% \includegraphics[width=\textwidth]{spur.pdf}
 \end{center}
@@
 David W. Hogg,
+Benjamin Johnson,
 Kathryn V. Johnston,

### 2018-04-29 paper/gd1.tex
A few tweaks

--- 2018-04-29 paper/gd1.tex
@@
 streams of stars that are extremely valuable tracers of the large- and
-small-scale distribution of matter in the Galaxy.
+small-scale distribution of mass in the Galaxy.
 Most stellar streams discovered around the Milky Way reside in the Galactic halo
-and are therefore primarily sensitive to the distribution of dark matter.
+and are therefore primarily sensitive to the dark matter distribution.
 We present a sample of highly-probable members of the longest cold stream, GD-1,
@@
 
-\figurename~\ref{fig:sfd-cmd} (middle panel) again shows the proper-motion and CMD selected stream stars, with interesting features
+From the combined proper motion and CMD selection of stream members
+(\sectionname~\ref{sec:data}), it is now clear that the GD-1 stream extends
+at least $90^\circ$ in apparent length.
+Several clear under-densities and gaps are also
+\figurename~\ref{fig:sfd-cmd} (middle panel) again shows the proper-motion and CMD selected stream stars, with several features
 

### 2018-04-29 paper/gd1.tex
feature analysis

--- 2018-04-29 paper/gd1.tex
@@
 dust extinction or variations in extinction along the stream.
-\figurename~\ref{fig:sfd-cmd} (top panel) shows the $V$-band extinction in the
+\figurename~\ref{fig:track-and-model} (top panel) shows the $V$-band extinction in the
 region around the GD-1 stream, computed from the Schlegel-Finkbeiner-Davis
@@
 
-\figurename~\ref{fig:sfd-cmd} (middle panel) again shows the proper-motion and CMD selected stream stars, with interesting features
+\figurename~\ref{fig:features} (middle panel) again shows the proper-motion and CMD selected stream stars, with interesting features
 
@@
 \begin{center}
-% \includegraphics[width=\textwidth]{sfd-cmd.pdf}
+\includegraphics[width=\textwidth]{features.pdf}
 \end{center}
@@
 \todo{AB}
-\label{fig:sfd-cmd}
+\label{fig:features}
 }
@@
 David W. Hogg,
-Benjamin Johnson,
+Benjamin D. Johnson,
 Kathryn V. Johnston,

### 2018-04-29 paper/gd1.tex
text thoughts

@@
 In this sense, streams represent one of the most promising directions for
-studying or ruling out the existence of small-scale dark matter sub-halos, as
-would be predicted by standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
+testing the existence of small-scale dark matter sub-halos, predicted by 
+standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
 (\citealt{Erkal:2015, Sanders:2016, Bovy:2017}).
@@
 was found to span $\approx 60^\circ$ in Heliocentric sky coordinates, partially
-because of its relative proximity to the sun ($\approx 7$--$10~\textrm{kpc}$).
+because of its relative proximity to the Sun ($\approx 7$--$10~\textrm{kpc}$).
 Despite being close and relatively high surface-density, no remnant progenitor
@@
 GD-1 stream (\citealt{Koposov:2010}) and from modeling the stream track in
-phase-space coordinates (\citealt{Bovy:2016}), it has been used to show that the
+phase-space coordinates \citep{Bowden:2015, Bovy:2016}, it has been used to show that the
 dark matter distribution within its orbit is consistent with being close to
@@
 $10^6$--$10^7~\msun$ subhalo.
+\todo{implications of subhalo detection (cite bullock review along the lines: on small scales, lcdm has some issues, so alternative classes of models proposed, some that lack structure on small scales; so ruling on the existence of subhalos expected to be detectable with GD-1 directly informs the nature of dark matter.)}
 
@@
 More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
-interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\circ$ span of the stream (\citealt{DeBoer:2017}):
+interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\circ$ span of the stream (\citealt{DeBoer:2018}):
 [summarize their conclusions].
+\todo{these analyses relied on binned data, still contaminated by MW stars}
 
@@
 
+\section{Results}
+\label{sec:results}
+
+- properties of GD-1 along the track, and beyond
+
+\subsection{Global properties}
+\label{sec:res_global}
+
+- for study of global properties, we first find stream ridge points as a median $\phi_2$ every $x^\circ$ along $\phi_1$
+- then we fit a second order polynomial to these ridge points, and define a stream track as the region $0.6^\circ$ above and beyond the best-fit polynomial (shown with two blue lines in Figure~\ref{fig:track-and-model})
+
+- first measure density variations in background subtracted profile
+- find maximum -- morphology as expected from a progenitor in the final stages of dissolution (cit)
+- several prominent gaps: density contrast, width
+- some (all?) seen in earlier data
+- but not related to dust
 All of the clearly identifiable portions of the GD-1 stream are located at high
@@
 
-Another \todo{segue to this paragraph...}.
-However, we find that \todo{...}.
-We discuss the scanning pattern in more detail in
-Appendix~\ref{sec:completeness}.
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=\textwidth]{track_observables.pdf}
+\end{center}
+\caption{%
+\todo{APW}
+\label{fig:track-and-model}
+}
+\end{figure}
 
-\section{Results}
-\label{sec:results}
+- also show median properties along the stream 5deg window, 1deg spacings
+- coordinates and proper motion vary
+- these observations are reproduced with a simple model
 
-\todo{segue sentences...}
+% \todo{segue sentences...}
 
-\subsection{Global properties}
-\label{sec:res_global}
+% Another \todo{segue to this paragraph...}.
+% However, we find that \todo{...}.
+% We discuss the scanning pattern in more detail in
+% Appendix~\ref{sec:completeness}.
 
-\figurename~\ref{fig:features} (middle panel) again shows the proper-motion and CMD selected stream stars, with interesting features
+\subsection{Off-track features}
+\label{sec:res_gap}
 
@@
 
-\subsection{Gaps and under-densities}
-\label{sec:res_gap}
+\figurename~\ref{fig:features} (middle panel) again shows the proper-motion and CMD selected stream stars, with interesting features
 
-\begin{figure}[h]
-\begin{center}
-% \includegraphics[width=\textwidth]{track-and-model.pdf}
-\end{center}
-\caption{%
-\todo{AB}
-\label{fig:track-and-model}
-}
-\end{figure}
-
-\subsection{Spur}
-\label{sec:res_spur}
+\section{Discussion}
+\label{sec:discussion}
 
@@
 \caption{%
-\todo{AB}
-\label{fig:spur}
+\todo{APW}
+\label{fig:orbit}
 }
@@
 
-\section{Discussion}
-\label{sec:discussion}
+- novel way to identify members -- helpful that retrograde, but hopefully also applicable to other streams as kinematic important
+- gd-1 even longer -- useful for potential constraints --> future
 
-\section{Conclusions}
-\label{sec:conclusions}
+- progenitor? if confirmed, can create more realistic models (so far only modeled the whole stream as just a leading or trailing arm)
+- to confirm -- measure actions, which change direction (some bovy paper)?
+
+- no longer simple stream, discuss possible causes
+- evaporation?
+
+- smooth component of the potential?
+-- show orbit
+-- chaos?
+-- bar?
+
+- encounters?
+-- work out mass / time combinations in impulse approximation
+-- comment on GMC vs subhalo
+
+- future:
+-- radial velocities, measure 3d velocity structure (unique predictions)
+-- tests for other outstanding issues
+-- model facets of stream morphology to ascertain if consistent with encounter signatures
+
+% \section{Conclusions}
+% \label{sec:conclusions}
 

### 2018-04-29 paper/gd1.tex

@@
 \begin{abstract}\noindent % trust me
-Tidally-disrupted globular clusters are transformed into thin, dynamically-cold streams
-of stars that are extremely valuable tracers of the large- and small-scale
-distribution of matter in the Galaxy.
+Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
+streams of stars that are extremely valuable tracers of the large- and
+small-scale distribution of mass in the Galaxy.
 Most stellar streams discovered around the Milky Way reside in the Galactic halo
-and are therefore primarily sensitive to the distribution of dark matter.
+and are therefore primarily sensitive to the dark matter distribution.
 We present a sample of highly-probable members of the longest cold stream, GD-1,
@@
 
+Another \todo{segue to this paragraph...}.
+However, we find that \todo{...}.
+We discuss the scanning pattern in more detail in
+Appendix~\ref{sec:completeness}.
+
+\section{Results}
+\label{sec:results}
+
+\todo{segue sentences...}
+
+\subsection{Global properties}
+\label{sec:res_global}
+
+From the combined proper motion and CMD selection of stream members
+(\sectionname~\ref{sec:data}), it is now clear that the GD-1 stream extends
+at least $90^\circ$ in apparent length.
+Several clear under-densities and gaps are also
+\figurename~\ref{fig:sfd-cmd} (middle panel) again shows the proper-motion and CMD selected stream stars, with several features
+
 \begin{figure}[h]

### 2018-04-29 paper/gd1.tex
words and model

@@
 \newcommand{\article}{\textsl{Letter}}
+
+\newcommand{\sectionname}{Section}
+\newcommand{\equationname}{Equation}
+\renewcommand{\tablename}{Table}
 
@@
 term) and $\mu_{\phi_2, \odot}$.
-GD-1 member stars are visible as the over-density at approximately
-$(\mu_{\phi_1, \odot}, \mu_{\phi_2, \odot}) \approx (XX, YY)~\masyr$.
+GD-1 member stars are visible as the over-density around
+$(\mu_{\phi_1, \odot}, \mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.
 
@@
 
-To improve the contrast of the stream over the background, we next cross-match
-the sample to the \pans\ photometric catalog and use the photometry to further
-clean the sample.
+To improve the contrast of the stream over the background, we cross-match the
+sample to the \pans\ photometric catalog and use the photometry to further clean
+the sample.
 \figurename~\ref{fig:selection} (bottom right) shows the distribution of
@@
 color-magnitude diagram (CMD).
-\todo{Did you also cut on $phi_2$?}.
-The over-plotted isochrone is from the MESA Isochrones \& Stellar Tracks (MIST;
-\citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
+\todo{AB: Did you also cut on $phi_2$?}.
+The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
+Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
 $12~\textrm{Gyr}$ old population with $\feh = -1.XX$ at a distance of $XX~\kpc$.
 We use this isochrone to motivate a polygonal selection in de-reddened $g-i$
-color and apparent $g$-band magnitude, as visualized by the \todo{shaded} region
-in the bottom right panel of \figurename~\ref{fig:selection}.
+color and apparent $g$-band magnitude, as visualized by the shaded region in the
+bottom right panel of \figurename~\ref{fig:selection}.
 
 \figurename~\ref{fig:selection} (bottom left) shows the final sample of GD-1
-stream member candidates after selecting on both proper motions and photometry.
-\todo{Comment on how this is the bomb ass diggity and the clearest view of the
-stream so far}
-\todo{Mention the under-densities}
-\todo{Already mention the spur here}
+stream member candidates after selecting on both proper motion and photometry.
+The GD-1 stream is clearly visible in the point density of individual stars (the
+positions are not binned); this is the most pure view of the stream to date.
+The increased contrast shows that the stream extends at least another $20^\circ$
+to negative longitudes.
+Several of the under-densities and gaps hinted at from photometric selection
+alone (\citealt{Koposov:2010, Carlberg:2013}) appear as striking features in
+this significantly cleaner density map of the GD-1 stream.
+Finally, at least two new features are visible in this new view of the stream:
+(1) ``the spur,'' stars above the main stream track (in $\phi_2$) between
+$-40^\circ \lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars
+below the main stream track between $-20^\circ \lesssim \phi_1 \lesssim
+-10^\circ$.
+We discuss each of these in more detail in \sectionname
+s~\ref{sec:results}--\ref{sec:discussion} below.
 
@@
 
-- properties of GD-1 along the track, and beyond
+We extract high-confidence stream stars from the proper motion and CMD selected
+sample by further selecting stars close to the main stream track.
+We use these main stream track stars to study the global properties of the
+stream, and find interesting substructures off of the main stream track.
 
@@
 
-- for study of global properties, we first find stream ridge points as a median $\phi_2$ every $x^\circ$ along $\phi_1$
-- then we fit a second order polynomial to these ridge points, and define a stream track as the region $0.6^\circ$ above and beyond the best-fit polynomial (shown with two blue lines in Figure~\ref{fig:track-and-model})
+To study of global properties of the stream as a function of stream longitude,
+we extract stream stars around a low-order polynomial model for the median
+stream latitude, $\phi_2$, as a function of stream longitude, $\phi_1$.
+In detail, we find stream ridge points by computing a running median in $\phi_2$
+in steps of $XX^\circ$ along $\phi_1$, then fit a second-order polynomial to the
+ridge points to find $\phi_{2, \textrm{track}}(\phi_1)$.
+We define the stream as the region within $\left| \phi_2 - \phi_{2,
+\textrm{track}}(\phi_1) \right| < 0.6^\circ$.
+\figurename~\ref{fig:XX} (XX panel) again shows the high-confidence
+stream stars, and the two curved lines (blue) show the upper and lower boundary
+of the stream region.
+
+\begin{figure}[h]
+\begin{center}
+% \includegraphics[width=\textwidth]{???.pdf}
+\end{center}
+\caption{%
+\todo{???}
+\label{fig:XX}
+}
+\end{figure}
+
+The clearly identifiable portions of the GD-1 stream are located at high
+Galactic latitudes ($b > 20^\circ$), and we therefore do not expect significant
+dust extinction or variations in extinction within the stream region.
+\figurename~\ref{fig:XX} (XX panel) shows the $V$-band extinction
+in the region around the GD-1 stream, computed from the
+Schlegel-Finkbeiner-Davis extinction map (\cite{Schlegel:1998}; hereafter SFD).
+For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
+1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
+0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V}
+\approx 0.03~\textrm{mag}$.
+For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
+dust extinction becomes more appreciable, and the stream stars are harder to
+select apart from the background.
+
+We extract properties of the stream from the stream region defined above as a
+function of $\phi_1$ by computing properties in overlapping windows:
+we set the window size to $XX^\circ$ in $\phi_1$, and shift the window center by
+$1^\circ$ along the stream region.
+\figurename~\ref{fig:track-and-model} shows the median latitude, $\phi_2$,
+stream surface density, $XX$, stream width, $XX$, and median proper motions,
+$\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way.
+For the surface density, we subtract an estimate of the background surface
+density by \todo{AB}.
+To estimate the stream width, $w$, we compute the median absolute deviation,
+$\textrm{MAD}$, and use this to estimate a robust standard deviation, $w = 1.5
+\times MAD$.
+
+\todo{Comment on density variations, stream width variations. Do we want a
+$\Delta \phi_2$ plot (relative to mean stream track) with 15th/85th
+percentiles?}
 
@@
 - but not related to dust
-All of the clearly identifiable portions of the GD-1 stream are located at high
-Galactic latitudes ($b > 20^\circ$), and we therefore do not expect significant
-dust extinction or variations in extinction along the stream.
-\figurename~\ref{fig:track-and-model} (top panel) shows the $V$-band extinction in the
-region around the GD-1 stream, computed from the Schlegel-Finkbeiner-Davis
-extinction map (\cite{Schlegel:1998}; hereafter SFD).
-For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
-1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
-0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V} \approx
-0.03~\textrm{mag}$.
-For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
-dust extinction becomes more appreciable.
 
-Another \todo{segue to this paragraph...}.
-However, we find that \todo{...}.
-We discuss the scanning pattern in more detail in
-Appendix~\ref{sec:completeness}.
-
-\section{Results}
-\label{sec:results}
-
-\todo{segue sentences...}
-
-\subsection{Global properties}
-\label{sec:res_global}
-
-From the combined proper motion and CMD selection of stream members
-(\sectionname~\ref{sec:data}), it is now clear that the GD-1 stream extends
-at least $90^\circ$ in apparent length.
-Several clear under-densities and gaps are also
-\figurename~\ref{fig:sfd-cmd} (middle panel) again shows the proper-motion and CMD selected stream stars, with several features
+We compare the measured stream properties to a simple model for the phase-space
+density of the stream, generated by only simulating the orbital evolution of
+cluster stars once they are tidally stripped from the progenitor.
+In detail, during simulation, star particles are released from the Lagrange
+points of the cluster with a spread in position and velocity set by the mass of
+the satellite and the location within the host galaxy, with some tunable scale
+parameters.
+We use the parametrization and adopted scale parameters of \citet{Fardal:2015},
+who fit the release distribution parameters by comparing to full $N$-body
+simulations across a range of mass and orbital parameters.
+This method can accurately reproduce the mean track of the resulting stream, but
+the density of stars along the stream depends strongly on the mass-loss history
+and internal kinematics of the progenitor system.
+Here we assume a constant mass-loss rate, and release stream particles uniformly
+in time from each Lagrange point.
 
@@
 \end{figure}
-
-- also show median properties along the stream 5deg window, 1deg spacings
-- coordinates and proper motion vary
-- these observations are reproduced with a simple model
-
-% \todo{segue sentences...}
-
-% Another \todo{segue to this paragraph...}.
-% However, we find that \todo{...}.
-% We discuss the scanning pattern in more detail in
-% Appendix~\ref{sec:completeness}.
 
@@
 % \label{sec:conclusions}
+
+% From the combined proper motion and CMD selection of stream members
+% (\sectionname~\ref{sec:data}), it is now clear that the GD-1 stream extends
+% at least $90^\circ$ in apparent length.
+% Several clear under-densities and gaps are also
+% \figurename~\ref{fig:sfd-cmd} (middle panel) again shows the proper-motion and CMD selected stream stars, with several features
 

### 2018-04-29 paper/gd1.tex
Potential

@@
 
+To compute initial conditions for the stream model, we fit an orbit to the
+observed properties of the stream in a fixed model for the gravitational
+potential of the Milky Way.
+We use a three-component potential model to represent the Galaxy, consisting of
+a Miyamoto-Nagai disk (\citealt{Miyamoto:1975}), a Hernqust bulge
+(\citealt{Hernquist:1990}), and a spherical Navarro-Frenk-White dark matter halo
+(\citealt{Navarro:1996}).
+We set the disk scale-length parameters to closely match the disk model found in
+\citet{Bovy:2015}, and other parameters are set to reproduce an approximately
+flat rotation curve between $5 < R < 15~\kpc$ with a circular velocity at the
+solar circle, $v_{\textrm{circ}, \odot} \approx 230~\kms$.
+
+To initialize the model stream generation, we \todo{...}
+
+Stream properties computed from resulting model stream are plotted in \figurename~\ref{fig:track-and-model} as \todo{XX}.
+
 \begin{figure}[h]

### 2018-04-29 paper/gd1.tex
MAD
abstract

@@
 and are therefore primarily sensitive to the dark matter distribution.
-We present a sample of highly-probable members of the longest cold stream, GD-1,
-selected to have small parallaxes and retrograde proper motions using data from
-the \gaia\ mission, and \pans\ photometry consistent with an old and metal-poor
-population at a distance of $\sim10~\kpc$.
-This selection extended the stream by $20^\circ$, revealed a possible location of the progenitor and mapped significant density variations along the stream in unprecedented detail.
-In addition to several prominent gaps, for the first time we also find evidence of stream members off the main track.
-No(?) secular formation mechanisms predict the existence of such spurs, but they are a natural result of a stream interaction with a massive perturber.
-Using the impulse approximation arguments, we estimate that the detected underdensity and the associated spur could have been induced by a perturbation of a $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
-Our model of GD-1 indicates that the stream crosses the Galactic plane at a large radius, which puts the likely origin of the perturber beyond the stellar disk.
+The longest and most prominent thin stream --- the GD-1 stream --- has been used
+to constrain the shape of the inner halo, and is a promising place to search for
+signatures of dark matter subhalo interactions.
+% Aims
+Here we present a sample of highly-probable members of GD-1, selected to have
+small parallaxes and retrograde proper motions using data from the \gaia\
+mission, and \pans\ photometry consistent with an old and metal-poor population
+at a distance of $\sim10~\kpc$.
+This selection extends the stream by $20^\circ$, reveals a plausible location of
+the progenitor, and shows significant density variations along the stream in
+unprecedented detail.
+In addition to several prominent gaps, for the first time we also find evidence
+of stream members off the main track.
+Such features are a natural result of a stream interaction with a massive
+perturber, or an indication of complex internal kinematics within the
+stream progenitor.
+If caused by a subhalo interaction, we estimate that the detected underdensity
+and the associated spur could have been induced by a perturbation of a
+$\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
 \end{abstract}
@@
 $\textrm{MAD}$, and use this to estimate a robust standard deviation, $w = 1.5
-\times MAD$.
+\times \textrm{MAD}$.
 

### 2018-04-29 paper/gd1.tex
model stream

@@
 We use a three-component potential model to represent the Galaxy, consisting of
-a Miyamoto-Nagai disk (\citealt{Miyamoto:1975}), a Hernqust bulge
+a Miyamoto-Nagai disk (\citealt{Miyamoto:1975}), a Hernquist bulge
 (\citealt{Hernquist:1990}), and a spherical Navarro-Frenk-White dark matter halo
@@
 
-To initialize the model stream generation, we \todo{...}
-
-Stream properties computed from resulting model stream are plotted in \figurename~\ref{fig:track-and-model} as \todo{XX}.
+To initialize the model stream generation, we assume that the present day (i.e.
+final) position of the progenitor system is at $\phi_1 = XX^\circ$.
+We integrate backwards for $4~\textrm{Gyr}$ using the orbital parameters at this
+location using the Leapfrog integration scheme with a timestep of
+$0.5~\textrm{Myr}$, then begin the model stream generation using the Lagrange
+point stripping method discussed above.
+Stream particles are release at each timestep of the progenitor orbit
+integration, resulting in 16,000 stream particles by the end of the model stream
+generation.
+We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
+the mass of the progenitor until present day, where we assume $M = 0~\msun$
+(i.e. fully disrupted).
+Stream properties computed from this resulting model stream are plotted in
+\figurename~\ref{fig:track-and-model} as \todo{XX}.
 

### 2018-04-29 paper/gd1.tex
redundant

@@
 
-We extract properties of the stream from the stream region defined above as a
-function of $\phi_1$ by computing properties in overlapping windows:
+We extract properties of the stream from the region defined above as a function
+of $\phi_1$ by computing properties in overlapping windows:
 we set the window size to $XX^\circ$ in $\phi_1$, and shift the window center by

### 2018-04-29 paper/gd1.tex
long introduction

@@
 and are therefore primarily sensitive to the dark matter distribution.
-The longest and most prominent thin stream --- the GD-1 stream --- has been used
+The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
 to constrain the shape of the inner halo, and is a promising place to search for
@@
 Grillmair:2006b, Belokurov:2006, Belokurov:2007, Bonaca:2012, Shipp:2018}; see
-also the recent review \citealt{Grillmair:2016, Newberg:2016}), and the majority
-have no obvious, bound progenitor systems;
-of the thin, likely globular-cluster-origin streams, only two have clear
-progenitors (NGC 5466 and Palomar 5).
+also the recent review \citealt{Grillmair:2016, Newberg:2016}).
+% , and the majority
+% have no obvious, bound progenitor systems;
+% of the thin, likely globular-cluster-origin streams, only two have clear
+% progenitors (NGC 5466 and Palomar 5).
 
@@
 GD-1 was found using a matched filter on photometric data from the SDSS and
-was found to span $\approx 60^\circ$ in Heliocentric sky coordinates, partially
+was found to span $\approx 60^\circ$ on the sky, partially
 because of its relative proximity to the Sun ($\approx 7$--$10~\textrm{kpc}$).
@@
 cluster or stellar system has been found associated with the stream, but its
-width,  metallicity ($\feh \approx -1.4$; \citealt{Koposov:2010}) and estimated
-stellar mass ($M \approx 2 \times 10^4~\msun$; \citealt{Koposov:2010}) are
-consistent with being a disrupted globular cluster.
+width ($\sigma \approx 12'$),  metallicity ($\feh \approx -1.4$) and estimated
+stellar mass ($M \approx 2 \times 10^4~\msun$) are
+consistent with being a disrupted globular cluster \citep{Koposov:2010}.
 
@@
 GD-1 stream (\citealt{Koposov:2010}) and from modeling the stream track in
-phase-space coordinates \citep{Bowden:2015, Bovy:2016}, it has been used to show that the
-dark matter distribution within its orbit is consistent with being close to
+phase-space coordinates \citep{Bowden:2015, Bovy:2016}, we learned that the
+dark matter distribution in the inner $\approx20\,\rm kpc$ is consistent with being close to
 spherical.
+This is in line with findings from simulated Milky Way-like galaxies; while dark matter halos are in general expected to be triaxial \citep{Allgood:2006}, the baryonic processes forming galaxies tend to make their inner regions more spherical \citep{Zhu:2016}.
 
-Its long length --- and therefore its implied old dynamical age --- combined
-with its particular orbit also make it a prime stream to search for interactions
-with dark matter subhalos:
-because of its large pericentric distance and retrograde orbit with respect to
-the Galactic bar, density variations in the stream are not expected from either
-interactions with giant molecular clouds (\citealt{Amorisco:2016}) or resonant
-encounters with the bar (\citealt{Pearson:2017}).
-By estimating a stream age of $\approx 4~\textrm{Gyr}$ assuming a progenitor
-mass of $10^5~\msun$, and by taking a subhalo mass function and density from
-XXX, \citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
+An implied old dynamical age, combined with its orbital properties, makes GD-1 a prime stream to not only map the dark matter on large scales, but to also study its small scale structure.
+Density variations in the stream can be induced from interactions with dark matter subhalos \citep[e.g.,][]{Ngan:2014}, but also from interactions with giant molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar \citep{Pearson:2017}.
+However, these baryonic perturbations are unlikely in the case of GD-1 because of its large pericentric distance and retrograde orbit with respect to
+the Galactic bar.
+By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
+mass of $10^5~\msun$, and by assuming properties of the subhalo population from dark matter-only simulations \citep{Springel:2008, Diemand:2008} (scaled to the Milky Way halo mass of $10^{12}\,\msun$, and reduced in number by a factor of 3 to account for disruption by the Galactic disk), \citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
 ($\approx 5$--$7^\circ$) gap caused by an interaction with a
 $10^6$--$10^7~\msun$ subhalo.
-\todo{implications of subhalo detection (cite bullock review along the lines: on small scales, lcdm has some issues, so alternative classes of models proposed, some that lack structure on small scales; so ruling on the existence of subhalos expected to be detectable with GD-1 directly informs the nature of dark matter.)}
+A definitive ruling on the presence of dark matter subhalos in this mass regime would directly inform on the nature of dark matter \citep[e.g.][]{Bullock:2017}.
 
-\todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
-(\citealt{Carlberg:2013}).
-More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
-interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\circ$ span of the stream (\citealt{DeBoer:2018}):
-[summarize their conclusions].
-\todo{these analyses relied on binned data, still contaminated by MW stars}
-
-In this \article, we present clearest view of the GD-1 stream to date using
-astrometric data from the \gaia\ mission data release 2 (\DR) combined with
-precise photometry from the \pans\ survey.
-We find XX under-densities gaps, some consistent with ...
-We revisit...do not try to model or fit for Galactic potential.
-Focus on stream properties, data.
+Photometric studies of GD-1 revealed the presence of density variations and gaps along the stream, as well as wiggles in the main stream track \citep[][]{Carlberg:2013, DeBoer:2018}.
+Purely photometric stream studies suffer to some extent from contamination of the Milky Way foreground, whose small-scale spatial variations could partially account for density inhomogeneities reported along cold streams \citep[e.g.,][]{Ibata:2016}.
+In this \article, we improve upon the selection of likely GD-1 members by using astrometric data from the \gaia\ mission data release 2 in addition to precise photometry from the \pans\ survey (\S\,\ref{sec:data}).
+This data enabled us to measure global properties of GD-1 to an unprecedented precision (\S\,\ref{sec:res_global}), and detect clear signatures of stream stars beyond the main track (\S\,\ref{sec:res_gap}).
+Implications of this revolutionary remapping of GD-1 are discussed in \S\,\ref{sec:discussion}.
+% \todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
+% (\citealt{Carlberg:2013}).
+% More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
+% interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\circ$ span of the stream (\citealt{DeBoer:2018}):
+% [summarize their conclusions].
+% \todo{these analyses relied on binned data, still contaminated by MW stars}
+% 
+% In this \article, we present clearest view of the GD-1 stream to date using
+% astrometric data from the \gaia\ mission data release 2 (\DR) combined with
+% precise photometry from the \pans\ survey.
+% We find XX under-densities gaps, some consistent with ...
+% We revisit...do not try to model or fit for Galactic potential.
+% Focus on stream properties, data.
 
@@
 
+- Spur vs. blob
+- Estimate velocity offset and significance
+- No indication of these features in previous work
+
 \begin{figure}[h]
@@
 \label{sec:discussion}
-
-\begin{figure}[h]
-\begin{center}
-% \includegraphics[width=\textwidth]{spur.pdf}
-\end{center}
-\caption{%
-\todo{APW}
-\label{fig:orbit}
-}
-\end{figure}
 
@@
 -- model facets of stream morphology to ascertain if consistent with encounter signatures
+-- so far a lot of focus on gaps produced in encounters -- extra features we detected beside the GD-1 track suggest that further theoretical focus on scattered material is needed to fully understand its dynamical history
 

### 2018-04-29 paper/gd1.tex
intro condensation part 1

@@
 
-Dynamically cold stellar streams are formed from the tidal disruption of stellar
+Dynamically cold stellar streams are formed during the tidal disruption of stellar
 systems by the gravitational field of their host galaxy.
@@
 scales (e.g., \citealt{Johnston:1999, Bonaca:2018}).
-Most stellar streams are found in the halos of galaxies and are therefore
-important tracers of the large-scale distribution of dark matter in galactic
-halos.
-
-Long, thin stellar streams are also excellent laboratories for studying
-small-scale structures in the mass distribution:
-thin streams can remain coherent for tens of orbital periods, depending in
-detail on the symmetries of the underlying galactic mass distribution (e.g.,
-\citealt{Erkal:2016a}).
-Encounters between massive perturbers and stream stars imprint variations in the
-phase-space density that can persist and remain visible for close to their total
-survival time (e.g., \citealt{Yoon:2011}).
+% Most stellar streams are found in the halos of galaxies and are therefore
+% important tracers of the large-scale distribution of dark matter in galactic
+% halos.
+% Long, thin stellar streams are also excellent laboratories for studying
+% small-scale structures in the mass distribution:
+% thin streams can remain coherent for tens of orbital periods, depending in
+% detail on the symmetries of the underlying galactic mass distribution (e.g.,
+% \citealt{Erkal:2016a}).
+Being dynamically cold, stellar streams can retain any imprints on its phase-space density from encounters with a massive perturber for close to its total survival time \citep[e.g.,][]{Yoon:2011}.
+% Encounters between massive perturbers and stream stars imprint variations in the
+% phase-space density that can persist and remain visible for close to their total
+% survival time (e.g., \citealt{Yoon:2011}).
 In this sense, streams represent one of the most promising directions for
@@
 
-Over 30 candidate stellar streams have been discovered throughout the halo of
-the Milky Way, thanks to large-area, multi-band photometric surveys (especially
-the Sloan Digital Sky Survey, SDSS; \citealt{York:2000}).
-The Milky Way streams have a wide range of properties such as length, velocity
-dispersion, and density (e.g., \citealt{Odenkirchen:2001, Grillmair:2006,
-Grillmair:2006b, Belokurov:2006, Belokurov:2007, Bonaca:2012, Shipp:2018}; see
-also the recent review \citealt{Grillmair:2016, Newberg:2016}).
+Over 30 candidate stellar streams have been discovered in the Milky Way \citep[see][for a recent review]{Grillmair:2016, Newberg:2016}.
+% thanks to large-area, multi-band photometric surveys (especially
+% the Sloan Digital Sky Survey, SDSS; \citealt{York:2000}).
+% The Milky Way streams have a wide range of properties such as length, velocity
+% dispersion, and density (e.g., \citealt{Odenkirchen:2001, Grillmair:2006,
+% Grillmair:2006b, Belokurov:2006, Belokurov:2007, Bonaca:2012, Shipp:2018}; see
+% also the recent review \citealt{Grillmair:2016, Newberg:2016}).
 % , and the majority
@@
 % progenitors (NGC 5466 and Palomar 5).
+% Most streams were first detected as a linear overdensity in stellar counts filtered to preferentially select faint blue stars, most likely members of a former Galactic satellite.
+GD-1 is the most prominent among the thin streams, discovered as an arcing overdensity of faint blue stars in the SDSS photometry \citep{Grillmair:2006}.
+It spans $\sim 60^\circ$ on the sky, which at its distance between $\sim 7$--$10~\textrm{kpc}$ translates to a physical length of $\sim 15~\kpc$.
+No remnant progenitor has been associated with the stream, but its
+width ($\sigma \approx 12'$), metallicity ($\feh \approx -1.4$), and estimated
+stellar mass ($M \approx 2 \times 10^4~\msun$)
+imply that GD-1 is a disrupted globular cluster \citep{Koposov:2010}.
 
-The most prominent and longest (in angular extent) thin stream is the GD-1
-stream (named after its discoverers; \citealt{Grillmair:2006}).
-GD-1 was found using a matched filter on photometric data from the SDSS and
-was found to span $\approx 60^\circ$ on the sky, partially
-because of its relative proximity to the Sun ($\approx 7$--$10~\textrm{kpc}$).
-Despite being close and relatively high surface-density, no remnant progenitor
-cluster or stellar system has been found associated with the stream, but its
-width ($\sigma \approx 12'$),  metallicity ($\feh \approx -1.4$) and estimated
-stellar mass ($M \approx 2 \times 10^4~\msun$) are
-consistent with being a disrupted globular cluster \citep{Koposov:2010}.
-
-The long physical length of the GD-1 stream ($\sim 15~\kpc$) and location in the
-Galactic halo (pericentric distance $r_\textrm{peri} \sim 13~\kpc$) make it an
+Its length and location in the
+Galactic halo (pericentric distance $r_\textrm{peri} \sim 13~\kpc$) make GD-1 an
 ideal object for constraining the gravitational potential of dark matter in the
@@
 Density variations in the stream can be induced from interactions with dark matter subhalos \citep[e.g.,][]{Ngan:2014}, but also from interactions with giant molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar \citep{Pearson:2017}.
-However, these baryonic perturbations are unlikely in the case of GD-1 because of its large pericentric distance and retrograde orbit with respect to
+However, these baryonic effects are unlikely to have perturbed GD-1 because of its large pericentric distance and retrograde orbit with respect to
 the Galactic bar.
 By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
-mass of $10^5~\msun$, and by assuming properties of the subhalo population from dark matter-only simulations \citep{Springel:2008, Diemand:2008} (scaled to the Milky Way halo mass of $10^{12}\,\msun$, and reduced in number by a factor of 3 to account for disruption by the Galactic disk), \citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
+mass of $10^5~\msun$, and by assuming properties of the subhalo population from dark matter-only simulations \citep[scaled to the Milky Way halo mass of $10^{12}\,\msun$, and reduced in number by a factor of 3 to account for disruption by the Galactic disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
 ($\approx 5$--$7^\circ$) gap caused by an interaction with a
@@
 Purely photometric stream studies suffer to some extent from contamination of the Milky Way foreground, whose small-scale spatial variations could partially account for density inhomogeneities reported along cold streams \citep[e.g.,][]{Ibata:2016}.
-In this \article, we improve upon the selection of likely GD-1 members by using astrometric data from the \gaia\ mission data release 2 in addition to precise photometry from the \pans\ survey (\S\,\ref{sec:data}).
+In this \article, we improve upon the selection of likely GD-1 members by using astrometric data from the \gaia\ mission, in addition to precise photometry from the \pans\ survey (\S\,\ref{sec:data}).
 This data enabled us to measure global properties of GD-1 to an unprecedented precision (\S\,\ref{sec:res_global}), and detect clear signatures of stream stars beyond the main track (\S\,\ref{sec:res_gap}).

### 2018-04-29 paper/gd1.tex
add reminder
Density variations

@@
 % - spur - encounter?
+
+% Make the point that if we can measure distance to spur, stream, RVs - measure energy offset which doesn't depend on time, like gap size
+
+% Make sure to include simple equations so can rescale
+
+% Note that epicyclic density variations naturally in, but caveat uniform release times
 
@@
 
-To study of global properties of the stream as a function of stream longitude,
+To study the global properties of the stream as a function of stream longitude,
 we extract stream stars around a low-order polynomial model for the median
@@
 In detail, we find stream ridge points by computing a running median in $\phi_2$
-in steps of $XX^\circ$ along $\phi_1$, then fit a second-order polynomial to the
-ridge points to find $\phi_{2, \textrm{track}}(\phi_1)$.
+in steps of $2^\circ$ with a window size of $4^\circ$ along $\phi_1$, then fit a
+second-order polynomial to the ridge points to find $\phi_{2,
+\textrm{track}}(\phi_1)$.
 We define the stream as the region within $\left| \phi_2 - \phi_{2,
-\textrm{track}}(\phi_1) \right| < 0.6^\circ$.
-\figurename~\ref{fig:XX} (XX panel) again shows the high-confidence
-stream stars, and the two curved lines (blue) show the upper and lower boundary
-of the stream region.
+\textrm{track}}(\phi_1) \right| < 0.75^\circ$.
+\figurename~\ref{fig:track-and-model} (top panel) again shows the
+high-confidence stream stars, and the two curved lines (blue) show the adopted
+upper and lower boundary of the stream region.
 
-\begin{figure}[h]
-\begin{center}
-% \includegraphics[width=\textwidth]{???.pdf}
-\end{center}
-\caption{%
-\todo{???}
-\label{fig:XX}
-}
-\end{figure}
+We extract properties of the stream from the region defined above as a function
+of $\phi_1$ by computing stream properties in overlapping windows:
+we set the window size to $4^\circ$ in $\phi_1$, and shift the window center by
+$1^\circ$ in longitude for each measurement.
+\figurename~\ref{fig:track-and-model} shows the median latitude, $\phi_2$,
+stream surface density, $XX$, stream width, $XX$, and median proper motions,
+$\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way, and empirical scatters
+for each estimated from each window (blue lines and shaded regions).
+For the surface density estimate, we subtract an estimate of the background
+surface density computed from the area outside of the bounded stream region.
 
-The clearly identifiable portions of the GD-1 stream are located at high
-Galactic latitudes ($b > 20^\circ$), and we therefore do not expect significant
-dust extinction or variations in extinction within the stream region.
-\figurename~\ref{fig:XX} (XX panel) shows the $V$-band extinction
-in the region around the GD-1 stream, computed from the
-Schlegel-Finkbeiner-Davis extinction map (\cite{Schlegel:1998}; hereafter SFD).
+Clear density variations, also apparent in the 2D positions of the sources,
+manifest as sharp features in the surface density estimates along the stream;
+these are not a result of variable dust extinction or completeness variations
+from the \gaia\ scanning pattern.
+First, the visible portions of the GD-1 stream are located at high Galactic
+latitudes ($b > 20^\circ$), and we therefore do not expect significant dust
+extinction or variations in extinction within the stream region.
 For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
@@
 0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V}
-\approx 0.03~\textrm{mag}$.
+\approx 0.03~\textrm{mag}$, as computed from the Schlegel-Finkbeiner-Davis
+extinction map (\citealt{Schlegel:1998}).
 For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
@@
 select apart from the background.
+Second, \todo{what to say about the Gaia scanning law...}
 
-We extract properties of the stream from the region defined above as a function
-of $\phi_1$ by computing properties in overlapping windows:
-we set the window size to $XX^\circ$ in $\phi_1$, and shift the window center by
-$1^\circ$ along the stream region.
-\figurename~\ref{fig:track-and-model} shows the median latitude, $\phi_2$,
-stream surface density, $XX$, stream width, $XX$, and median proper motions,
-$\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way.
-For the surface density, we subtract an estimate of the background surface
-density by \todo{AB}.
-To estimate the stream width, $w$, we compute the median absolute deviation,
-$\textrm{MAD}$, and use this to estimate a robust standard deviation, $w = 1.5
-\times \textrm{MAD}$.
-
-\todo{Comment on density variations, stream width variations. Do we want a
-$\Delta \phi_2$ plot (relative to mean stream track) with 15th/85th
-percentiles?}
-
-- first measure density variations in background subtracted profile
-- find maximum -- morphology as expected from a progenitor in the final stages of dissolution (cit)
-- several prominent gaps: density contrast, width
-- some (all?) seen in earlier data
-- but not related to dust
+The density variations likely represent real morphological changes along the
+GD-1 stream, as have been suggested before from binned samples of
+color-magnitude-selected GD-1 stars (\citealt{Carlberg:2013, DeBoer:2018}).
+The deep under-densities we see from the kinematically-cleaned sample correspond
+to previously reported gaps in the stream.
+There is also a well-defined, sharp over-density in the surface density profile
+close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
+either side:
+this is suggestive of a progenitor system in the final stages of dissolution
+(\todo{cite who}).
+\todo{Table of locations, sizes, and density constrasts for the features?}
 
@@
 To initialize the model stream generation, we assume that the present day (i.e.
-final) position of the progenitor system is at $\phi_1 = XX^\circ$.
+final) position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak
+of the observed surface density profile.
 We integrate backwards for $4~\textrm{Gyr}$ using the orbital parameters at this
@@
 
-\appendix
-\section{Completeness and the \gaia\ scanning pattern}
-\label{sec:completeness}
+% \appendix
+% \section{Completeness and the \gaia\ scanning pattern}
+% \label{sec:completeness}
+
+% \figurename~\ref{fig:XX} (XX panel) shows the $V$-band extinction
+% in the region around the GD-1 stream, computed from the
+% Schlegel-Finkbeiner-Davis extinction map (\cite{Schlegel:1998}; hereafter SFD).
 

### 2018-04-29 paper/gd1.tex
Fill in XXs

@@
 \figurename~\ref{fig:selection} (top right) shows the distribution of stars with
-$|\phi_2| < XX^\circ$ in solar-reflex-corrected proper motion components in the
+$|\phi_2| < 1^\circ$ in solar-reflex-corrected proper motion components in the
 GD-1 coordinate system, $\mu_{\phi_1, \odot}$ (which includes the $\cos{\phi_2}$
@@
 As a first selection of GD-1 candidate member stars, we select stars with proper
-motions $XX < \mu_{\phi_1, \odot} < XX~\masyr$ and $XX < \mu_{\phi_2, \odot} <
-XX~\masyr$, as visualized by the rectangle in the top right panel of
+motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot}
+< 1~\masyr$, as visualized by the rectangle in the top right panel of
 \figurename~\ref{fig:selection}.
@@
 Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
-$12~\textrm{Gyr}$ old population with $\feh = -1.XX$ at a distance of $XX~\kpc$.
+$12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
+$7.8~\kpc$.
 We use this isochrone to motivate a polygonal selection in de-reddened $g-i$

### 2018-04-30 paper/gd1.tex

@@
 From kinematic selection alone, the stream is identifiable as the over-density
-of stars around $\phi_2 = 0$ between longitudes $XX^\circ \lesssim \phi_1
-\lesssim XX^\circ$.
+of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim \phi_1
+\lesssim 0^\circ$.
 
@@
 \figurename~\ref{fig:selection} (bottom right) shows the distribution of
-proper-motion-selected candidate stream stars in a de-reddened \pans\
-color-magnitude diagram (CMD).
-\todo{AB: Did you also cut on $phi_2$?}.
+proper-motion-selected candidate stream stars within $\left|\phi_2\right| <
+1^\circ$ in a de-reddened \pans\ color-magnitude diagram (CMD).
 The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar

### 2018-04-30 paper/gd1.tex
orbit figure
discussion points

@@
 
+- spur & blob as related to gaps? -- classify possible explanations in this way
+- if from encounter -- could get more info on the perturber than from the gaps alone, say from delta v
+
+- progenitor
+-- rotation -- could explain spurs, but not gaps?
+
 - smooth component of the potential?
@@
 -- chaos?
--- bar?
+-- \todo{APW} bar -- hard to make small feature, size should be ~angular size of the bar at stream (influence of bar should be a radian of the stream) + relative velocity (retrograde)
 
@@
 -- radial velocities, measure 3d velocity structure (unique predictions)
--- tests for other outstanding issues
+- delta energy, 
+- cleaner selection (any stars in the gap)
+
+- (should go at least to symmetric in positive), not captured with naive selection, as there are gradients
+- prediction: very negative rv, should be easy to map the full extent
+
+% -- tests for other outstanding issues
 -- model facets of stream morphology to ascertain if consistent with encounter signatures

### 2018-04-30 paper/gd1.tex
sec 3.2

@@
 unprecedented detail.
-In addition to several prominent gaps, for the first time we also find evidence
-of stream members off the main track.
+In addition to confirming several prominent gaps detected in photometric maps of GD-1, for the first time we find evidence
+of stream members off of the main track.
 Such features are a natural result of a stream interaction with a massive
@@
 stream progenitor.
-If caused by a subhalo interaction, we estimate that the detected underdensity
-and the associated spur could have been induced by a perturbation of a
-$\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
+Velocity offset between these features and the stream is an order of magnitude larger than the variance in velocity fields of known globular cluster, thus favoring the encounter origin.
+If this is confirmed, off-stream features will enable a robust estimate of the perturber's mass.
+% If caused by a subhalo interaction, we estimate that the detected underdensity
+% and the associated spur could have been induced by a perturbation of a
+% $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
 \end{abstract}
@@
 motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot}
-< 1~\masyr$, as visualized by the rectangle in the top right panel of
+< 1~\masyr$, as visualized by the orange rectangle in the top right panel of
 \figurename~\ref{fig:selection}.
@@
 selection on proper motion, plotted in the GD-1 coordinate system.
-From kinematic selection alone, the stream is identifiable as the over-density
+From kinematic selection alone, the stream is already identifiable as the over-density
 of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim \phi_1
@@
 
-\begin{figure}[h]
+\begin{figure}
 \begin{center}
@@
 proper-motion-selected candidate stream stars within $\left|\phi_2\right| <
-1^\circ$ in a de-reddened \pans\ color-magnitude diagram (CMD).
+1^\circ$ in a \pans\ color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
 The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
@@
 stream member candidates after selecting on both proper motion and photometry.
-The GD-1 stream is clearly visible in the point density of individual stars (the
+The GD-1 stream is clearly visible in the point density of \emph{individual} stars (the
 positions are not binned); this is the most pure view of the stream to date.
@@
 to negative longitudes.
+Furthermore, the stream reaches its highest surface density where it is the narrowest ($\phi_1\sim-13^\circ$), so this map might have uncovered the location of its elusive progenitor.
 Several of the under-densities and gaps hinted at from photometric selection
 alone (\citealt{Koposov:2010, Carlberg:2013}) appear as striking features in
-this significantly cleaner density map of the GD-1 stream.
+this significantly cleaner map of the GD-1 stream.
 Finally, at least two new features are visible in this new view of the stream:
@@
 We use these main stream track stars to study the global properties of the
-stream, and find interesting substructures off of the main stream track.
+stream (\S\,\ref{sec:res_global}), and find interesting substructures off of the main stream track (\S\,\ref{sec:res_gap}).
 
@@
 $1^\circ$ in longitude for each measurement.
-\figurename~\ref{fig:track-and-model} shows the median latitude, $\phi_2$,
-stream surface density, $XX$, stream width, $XX$, and median proper motions,
-$\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way, and empirical scatters
-for each estimated from each window (blue lines and shaded regions).
-For the surface density estimate, we subtract an estimate of the background
-surface density computed from the area outside of the bounded stream region.
+Rows 2--5 in \figurename~\ref{fig:track-and-model} show the background-subtracted
+stream surface density, \todo{$\Sigma$?}, the median latitude, $\phi_2$, and median proper motions,
+$\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way, and the associated empirical scatters
+(blue lines and shaded regions).
+We estimate the local background surface density as an average of areas north and south from the bounded stream region.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=\textwidth]{track_observables.pdf}
+\end{center}
+\caption{%
+\todo{APW}
+\label{fig:track-and-model}
+}
+\end{figure}
 
 Clear density variations, also apparent in the 2D positions of the sources,
-manifest as sharp features in the surface density estimates along the stream;
-these are not a result of variable dust extinction or completeness variations
+manifest as sharp features in the surface density estimates along the stream.
+These are not a result of variable dust extinction or completeness variations
 from the \gaia\ scanning pattern.
@@
 0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V}
-\approx 0.03~\textrm{mag}$, as computed from the Schlegel-Finkbeiner-Davis
+\approx 0.03~\textrm{mag}$, as computed from the SFD
 extinction map (\citealt{Schlegel:1998}).
@@
 select apart from the background.
-Second, \todo{what to say about the Gaia scanning law...}
+Second, each star in our sample has at least XX visibility periods, while the sample median and variance are XX and XX, respectively, guaranteeing a robust estimate of proper motions.
+Variations in the number density of faint stars originating from the scanning pattern are XX times smaller than the density variations in the stream, and their spatial distribution does not map to the stream features.
+% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they 
+\todo{check above sentences +  fill in}
 
@@
 (\todo{cite who}).
-\todo{Table of locations, sizes, and density constrasts for the features?}
+% \todo{Table of locations, sizes, and density constrasts for the features?}
 
@@
 Here we assume a constant mass-loss rate, and release stream particles uniformly
-in time from each Lagrange point.
+in time from each Lagrange point; these choices match results from more detailed numerical experiments of globular cluster disruption \citep[e.g.,][]{Kupper:2012}.
 
@@
 We use a three-component potential model to represent the Galaxy, consisting of
-a Miyamoto-Nagai disk (\citealt{Miyamoto:1975}), a Hernquist bulge
-(\citealt{Hernquist:1990}), and a spherical Navarro-Frenk-White dark matter halo
+a disk (\citealt{Miyamoto:1975}), a bulge
+(\citealt{Hernquist:1990}), and a spherical dark matter halo
 (\citealt{Navarro:1996}).
@@
 solar circle, $v_{\textrm{circ}, \odot} \approx 230~\kms$.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=\textwidth]{orbitfit.pdf}
+\end{center}
+\caption{%
+\todo{APW}
+\label{fig:orbitfit}
+}
+\end{figure}
 
@@
 
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=\textwidth]{track_observables.pdf}
-\end{center}
-\caption{%
-\todo{APW}
-\label{fig:track-and-model}
-}
-\end{figure}
+\figurename~\ref{fig:orbitfit} shows \todo{APW: ...}
 
@@
 
-- Spur vs. blob
-- Estimate velocity offset and significance
-- No indication of these features in previous work
+- overdensities apparent beyond the main stream track used in previous section -- if truly associated with the stream, this would be a novel insight into disruption from the \gaia\ mission
+- since some contamination still remains present, it is possible that this is a variation in the MW population, just randomly coming into our selection boxes
+- we thus perform an experiment and compare cmd and proper motions of spur and blob fields to stream and background fields to see which they are associated with
 
-\begin{figure}[h]
+\begin{figure}
 \begin{center}
@@
 
-\figurename~\ref{fig:features} (middle panel) again shows the proper-motion and CMD selected stream stars, with interesting features
+- central panel of figure~\ref{fig:features} shows the map of GD-1, with fields of interest in shaded rectangles
+- we first analyze the spur, region, stream same phi1, phi2 below, and an off-stream field in phi2
+- for each of these fields, we show a pair of plots in the top row: cmd on the left and proper motions on the right. 
+- we first select on pm, and show cmd, and then select on cmd and show pm
+- in both of these spaces, selection regions used to make the main catalog are shown in shaded orange.
+- stream field (top left) has clear overdensity in the cmd selection box and clusters in proper motion
+- on the other hand, the off field has some stars entering both selection boxes, but without any overdensities in either
+- the spur field has fewer stars than the stream, but XX more than expected from the MW alone
+- furthermore, the distribution in selection boxes is distributed similarly as in the stream field
+- this confirms the association, but we note that in detail they differ
+- median in pm offset from the stream field just below is XX, YY mas yr
+- also dispersion different
+- don't have enough stars to measure distance from the turn-off, so defer to future studies to measure by some other means, e.g., from photometry
+
+- analogous paragraph for blob on the bottom
+
+% - Spur vs. blob
+% - Estimate velocity offset and significance
+% - No indication of these features in previous work
+
+% \figurename~\ref{fig:features} (middle panel) again shows the proper-motion and CMD selected stream stars, with interesting features
 
@@
 
-- spur & blob as related to gaps? -- classify possible explanations in this way
+- spur \& blob as related to gaps? -- classify possible explanations in this way
 - if from encounter -- could get more info on the perturber than from the gaps alone, say from delta v

### 2018-04-30 paper/gd1.tex
abstract tweaks, fig1 caption

@@
 and are therefore primarily sensitive to the dark matter distribution.
-The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
-to constrain the shape of the inner halo, and is a promising place to search for
-signatures of dark matter subhalo interactions.
-% Aims
-Here we present a sample of highly-probable members of GD-1, selected to have
+% The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
+% to constrain the shape of the inner halo, and is a promising place to search for
+% signatures of dark matter subhalo interactions.
+Here we present a sample of highly-probable members of GD-1, the longest and most prominent thin stream in the Milky Way, selected to have
 small parallaxes and retrograde proper motions using data from the \gaia\
@@
 of stream members off of the main track.
-Such features are a natural result of a stream interaction with a massive
-perturber, or an indication of complex internal kinematics within the
-stream progenitor.
-Velocity offset between these features and the stream is an order of magnitude larger than the variance in velocity fields of known globular cluster, thus favoring the encounter origin.
-If this is confirmed, off-stream features will enable a robust estimate of the perturber's mass.
+These features might be an indication of complex internal kinematics within a progenitor unlike any of the known globular clusters, or a result of a stream interaction with a massive perturber.
+% Detailed modeling of the positional and velocity offsets
+% or an indication of complex internal kinematics within the
+% stream progenitor.
+% Velocity offset between these features and the stream is an order of magnitude larger than the variance in velocity fields of known globular cluster, thus favoring the encounter origin.
+% If this is confirmed, off-stream features will enable a robust estimate of the perturber's mass.
 % If caused by a subhalo interaction, we estimate that the detected underdensity
@@
 \caption{
-    \todo{AB}
+On-sky positions of likely GD-1 members, in the coordinate system aligned with the stream.
+GD-1 is apparent as an overdensity in retrograde proper motions (orange box on the top right panel), so a proper motion selection already reveals the stream in positions of individual stars (top left).
+The stream population is older and more metal poor than the field Milky Way population, so the stream also stands out in the color-magnitude diagram (bottom right).
+Additional photometric selection of the stream's main sequence (orange region in the CMD) unveils the stream in unprecedented detail (bottom left).
+This map extends the stream by $20^\circ$, identifies a plausible location of the progenitor at $\phi_1\sim13^\circ$, confirms the existence of a significant gap at $\phi_1\sim-20^\circ$, shows additional under-densities at $\phi_1\sim-40^\circ,-5^\circ$ and for the first time finds stream overdensities off from the main track of a cold stream at $(\phi_1,\phi_2)\sim(-35^\circ,-1^\circ)$ and $(-15^\circ,-2^\circ)$.
 \label{fig:selection}

### 2018-04-30 paper/gd1.tex
fig4 caption

@@
 % signatures of dark matter subhalo interactions.
-Here we present a sample of highly-probable members of GD-1, the longest and most prominent thin stream in the Milky Way, selected to have
+Here we present a sample of highly-probable members of the longest cold stream, GD-1, selected to have
 small parallaxes and retrograde proper motions using data from the \gaia\
@@
 This map extends the stream by $20^\circ$, identifies a plausible location of the progenitor at $\phi_1\sim13^\circ$, confirms the existence of a significant gap at $\phi_1\sim-20^\circ$, shows additional under-densities at $\phi_1\sim-40^\circ,-5^\circ$ and for the first time finds stream overdensities off from the main track of a cold stream at $(\phi_1,\phi_2)\sim(-35^\circ,-1^\circ)$ and $(-15^\circ,-2^\circ)$.
+}
 \label{fig:selection}
-}
 \end{figure}
@@
 \todo{APW}
+}
 \label{fig:track-and-model}
-}
 \end{figure}
@@
 \todo{APW}
+}
 \label{fig:orbitfit}
-}
 \end{figure}
@@
 \end{center}
-\caption{%
-\todo{AB}
+\caption{
+Color-magnitude and proper-motion diagrams of fields associated with a spur above GD-1 (top), and fields related to the blob below GD-1 (bottom).
+The left-most pair of panels show properties in stream fields (darkest rectangles in the middle panel), the middle pairs present spur and blob (medium rectangles), while the right-most panels have fields on the opposite side from spur and blob (lightest rectangles).
+Selection boxes used to select stars shown in the middle panel are shown in orange in all CMD and proper-motion panels.
+Both the spur and the blob fields have more stars in the CMD selection box than the corresponding control fields.
+Proper motions of these stars follow the distribution of stars in adjacent stream fields, thus confirming the association of these off-track features with GD-1.
+}
 \label{fig:features}
-}
 \end{figure}

### 2018-04-30 paper/gd1.tex
Caption

@@
 stream latitude, $\phi_2$, as a function of stream longitude, $\phi_1$.
-In detail, we find stream ridge points by computing a running median in $\phi_2$
-in steps of $2^\circ$ with a window size of $4^\circ$ along $\phi_1$, then fit a
+We find stream ridge points by computing a running median in $\phi_2$ in steps
+of $2^\circ$ with a window size of $4^\circ$ along $\phi_1$, then fit a
 second-order polynomial to the ridge points to find $\phi_{2,
 \textrm{track}}(\phi_1)$.
-We define the stream as the region within $\left| \phi_2 - \phi_{2,
+We then define the stream as the region within $\left| \phi_2 - \phi_{2,
 \textrm{track}}(\phi_1) \right| < 0.75^\circ$.
@@
 We extract properties of the stream from the region defined above as a function
-of $\phi_1$ by computing stream properties in overlapping windows:
-we set the window size to $4^\circ$ in $\phi_1$, and shift the window center by
-$1^\circ$ in longitude for each measurement.
-Rows 2--5 in \figurename~\ref{fig:track-and-model} show the background-subtracted
-stream surface density, \todo{$\Sigma$?}, the median latitude, $\phi_2$, and median proper motions,
-$\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way, and the associated empirical scatters
-(blue lines and shaded regions).
-We estimate the local background surface density as an average of areas north and south from the bounded stream region.
+of $\phi_1$ by computing median stream properties in overlapping windows:
+we set the window size to $3^\circ$ in $\phi_1$, and shift the window center by
+$1^\circ$ in longitude for each subsequent measurement.
+Panels 3--5 in \figurename~\ref{fig:track-and-model} show the
+background-subtracted stream surface density and median proper motion
+components, $\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way (dark
+lines), with the associated empirical scatters (shaded regions).
+We estimate the local background surface density as an average of areas north
+and south of the bounded stream region.
 
@@
 \caption{%
-\todo{APW}
+Median stream properties computed from the proper motion and CMD filtered GD-1
+stars, and the same for a toy model for the stream track.
+\textit{Panel (a)}: Sky positions of candidate GD-1 stars selected using \gaia\
+proper motions and \pans\ photometry.
+Lines show a 2nd-order polynomial fit to the median stream track offset by $\pm
+0.75^\circ$; the region between these lines is adopted as the nomial stream
+region.
+\textit{Panel (b)}: Sky positions of simulated star particles from a toy mock
+stream model of the GD-1 stream track.
+\textit{Panel (c)}: Background-subtracted surface-density estimates computed in
+running, overlapping $3^\circ$ windows for both GD-1 data and toy model.
+\textit{Panel (d)}: Median proper motion along the stream longitude computed in
+the same bins as previous panel.
+\textit{Panel (e)}: Same as Panel (d), but for proper motion in stream latitude.
 }

### 2018-04-30 paper/gd1.tex
Orbit text

@@
 \textrm{track}}(\phi_1) \right| < 0.75^\circ$.
-\figurename~\ref{fig:track-and-model} (top panel) again shows the
-high-confidence stream stars, and the two curved lines (blue) show the adopted
-upper and lower boundary of the stream region.
+\figurename~\ref{fig:track-and-model}, panel (a), shows the high-confidence
+stream stars (same as in lower panel of \figurename~\ref{fig:selection}), and
+the two curved lines (blue) show the adopted upper and lower boundary of the
+stream region.
 
@@
 $1^\circ$ in longitude for each subsequent measurement.
-Panels 3--5 in \figurename~\ref{fig:track-and-model} show the
+Panels (c)--(e) in \figurename~\ref{fig:track-and-model} show the
 background-subtracted stream surface density and median proper motion
@@
 select apart from the background.
-Second, each star in our sample has at least XX visibility periods, while the sample median and variance are XX and XX, respectively, guaranteeing a robust estimate of proper motions.
-Variations in the number density of faint stars originating from the scanning pattern are XX times smaller than the density variations in the stream, and their spatial distribution does not map to the stream features.
+Second, 99\% of the stars in our sample have at least 8 visibility periods,
+while the sample median and root-variance in number of visibility periods are 14
+and 3, respectively, guaranteeing a robust estimate of proper motions.
+The variance in the number density of faint stars originating from the \gaia\
+scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
+with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
+\mu_{\phi_2, \odot} < 4~\masyr$ (i.e. above the stream selection box), far
+smaller than the density variations in the stream.
 % The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they
-\todo{check above sentences +  fill in}
 
@@
 (\todo{cite who}).
-% \todo{Table of locations, sizes, and density constrasts for the features?}
+This region is also the thinnest part of the stream, as would be expected for
+recently stripped material that has not yet begun to phase-mix in multiple
+dimensions.
+
+We also measured the median proper motion profile along the stream (panels
+(d)--(e)) and find that these are consistent with previously measured mean
+proper motions from SDSS-USNOB (\citealt{Koposov:2010}).
+Here, the proper motion gradient is resolved in striking detail for individual
+stars from \gaia\ (see, e.g., elongated over-density in upper right panel of
+\figurename~\ref{fig:selection}).
 
@@
 cluster stars once they are tidally stripped from the progenitor.
-In detail, during simulation, star particles are released from the Lagrange
-points of the cluster with a spread in position and velocity set by the mass of
-the satellite and the location within the host galaxy, with some tunable scale
-parameters.
-We use the parametrization and adopted scale parameters of \citet{Fardal:2015},
-who fit the release distribution parameters by comparing to full $N$-body
-simulations across a range of mass and orbital parameters.
+During simulation, star particles are released from the Lagrange points of the
+cluster with a spread in position and velocity set by the mass of the satellite
+and the location within the host galaxy, with some tunable scale parameters
+(using the parametrization and adopted scale parameters of \cite{Fardal:2015}).
 This method can accurately reproduce the mean track of the resulting stream, but
-the density of stars along the stream depends strongly on the mass-loss history
+the density of stars along the stream depends in detail on the mass-loss history
 and internal kinematics of the progenitor system.
 Here we assume a constant mass-loss rate, and release stream particles uniformly
-in time from each Lagrange point; these choices match results from more detailed numerical experiments of globular cluster disruption \citep[e.g.,][]{Kupper:2012}.
+in time from each Lagrange point; these choices match results from more detailed
+numerical experiments of globular cluster disruption
+\citep[e.g.,][]{Kupper:2012}.
 
@@
 solar circle, $v_{\textrm{circ}, \odot} \approx 230~\kms$.
+We use L-BFGS-B optimization as implemented in the \package{scipy}
+\texttt{Python} package to maximize the likelihood of the orbit given sky track
+and proper motions from this work with mean distance and giant star radial
+velocities from \cite{Koposov:2010}.
+We do not vary the parameters of the Milky Way model in the fit because we are
+only interested in producing a comparison model of the stream properties.
+
+\figurename~\ref{fig:orbitfit} shows the best-fitting orbit from the procedure
+outlined above.
+The GD-1 stream crosses the midplane at large radius, far from the bulk of the
+stellar density:
+the grey line in the left panel shows the radius that contains $\approx 90\%$ of
+the mass, assuming an exponential radial density profile with a scale radius
+$\R_d = 3~\kpc$ (an intermediate value of recent measurements, e.g.,
+\citealt{Bovy:2012}).
 
@@
 \caption{%
-\todo{APW}
+Best-fitting orbit in a simple (fixed) model for the gravitational potential of
+the Milky Way using phase-space information along the stream measured from this
+work combined with data from \citet{Koposov:2010}.
+Left panel shows the orbit in Galactocentric cylindrical radius, $R$, and height
+above the Galactic midplane, $z$;
+grey line at $z=0$ shows the approximate radius of the disk.
+Right panel shows the orbit projected onto the midplane in Galactocentric
+cartesian coordinates.
+The darker orbit section highlights the range of stream longitudes over which we
+clearly see GD-1 stream stars in \figurename~\ref{fig:selection}.
+Note that GD-1 is retrograde with respect to the disk (see approximate location
+and direction of motion of the sun in right panel, $\odot$).
 }
@@
 location using the Leapfrog integration scheme with a timestep of
-$0.5~\textrm{Myr}$, then begin the model stream generation using the Lagrange
-point stripping method discussed above.
-Stream particles are release at each timestep of the progenitor orbit
-integration, resulting in 16,000 stream particles by the end of the model stream
-generation.
+$0.5~\textrm{Myr}$, then forward-generate the model stream using the method
+discussed above.
+Stream particles are release at each timestep of the integrated progenitor
+orbit.
 We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
@@
 Stream properties computed from this resulting model stream are plotted in
-\figurename~\ref{fig:track-and-model} as \todo{XX}.
-
-\figurename~\ref{fig:orbitfit} shows \todo{APW: ...}
+\figurename~\ref{fig:track-and-model} as dashed (orange) lines.
 

### 2018-04-30 paper/gd1.tex
global props

@@
 select apart from the background.
-Second, 99\% of the stars in our sample have at least 8 visibility periods,
-while the sample median and root-variance in number of visibility periods are 14
-and 3, respectively, guaranteeing a robust estimate of proper motions.
+Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
+periods (median and root-variance of 14 and 3, respectively), guaranteeing a
+robust estimate of proper motions.
 The variance in the number density of faint stars originating from the \gaia\
@@
 This region is also the thinnest part of the stream, as would be expected for
-recently stripped material that has not yet begun to phase-mix in multiple
-dimensions.
+recently stripped material that has not yet phase-mixed in multiple dimensions.
 
@@
 cluster stars once they are tidally stripped from the progenitor.
-During simulation, star particles are released from the Lagrange points of the
-cluster with a spread in position and velocity set by the mass of the satellite
-and the location within the host galaxy, with some tunable scale parameters
-(using the parametrization and adopted scale parameters of \cite{Fardal:2015}).
-This method can accurately reproduce the mean track of the resulting stream, but
-the density of stars along the stream depends in detail on the mass-loss history
-and internal kinematics of the progenitor system.
-Here we assume a constant mass-loss rate, and release stream particles uniformly
-in time from each Lagrange point; these choices match results from more detailed
-numerical experiments of globular cluster disruption
-\citep[e.g.,][]{Kupper:2012}.
+We use the ``particle-spray'' stream generation method (\citealt{Fardal:2015})
+and assume a uniform mass-loss history to generate the model stream, which
+choices matches results from more detailed ($N$-body) numerical experiments of
+globular cluster disruption \citep[e.g.,][]{Kupper:2012}.
+% During simulation, star particles are released from the Lagrange points of the
+% cluster with a spread in position and velocity set by the mass of the
+% satellite and the location within the host galaxy, with some tunable scale
+% parameters (using the parametrization and adopted scale parameters of
+% \cite{Fardal:2015}).  This method can accurately reproduce the mean track of
+% the resulting stream, but the density of stars along the stream depends in
+% detail on the mass-loss history and internal kinematics of the progenitor
+% system.  Here we assume a constant mass-loss rate, and release stream
+% particles uniformly in time from each Lagrange point; these choices match
+% results from more detailed numerical experiments of globular cluster
+% disruption \citep[e.g.,][]{Kupper:2012}.
 
@@
 (\citealt{Navarro:1996}).
-We set the disk scale-length parameters to closely match the disk model found in
-\citet{Bovy:2015}, and other parameters are set to reproduce an approximately
-flat rotation curve between $5 < R < 15~\kpc$ with a circular velocity at the
-solar circle, $v_{\textrm{circ}, \odot} \approx 230~\kms$.
+We set the disk scale-length parameters to match the disk model found in
+\citet{Bovy:2015}; other parameters are set to produce an approximately flat
+disk rotation curve between with a circular velocity at the solar circle,
+$v_{\textrm{circ}, \odot} \approx 230~\kms$.
 We use L-BFGS-B optimization as implemented in the \package{scipy}
@@
 of the observed surface density profile.
-We integrate backwards for $4~\textrm{Gyr}$ using the orbital parameters at this
-location using the Leapfrog integration scheme with a timestep of
-$0.5~\textrm{Myr}$, then forward-generate the model stream using the method
-discussed above.
-Stream particles are release at each timestep of the integrated progenitor
-orbit.
+We numerically integrate the backwards from this location for $4~\textrm{Gyr}$
+then forward-generate the model stream using the method discussed above, and
+release stream particles at each timestep.
 We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
@@
 \figurename~\ref{fig:track-and-model} as dashed (orange) lines.
+The model stream qualitatively matches the stream track in all phase-space
+dimensions, but has a much smoother surface density profile along the extent of
+the stream.
+We therefore don't expect that the observed surface density variations could be
+caused by, e.g., epicyclic over-densities (\citealt{Kupper:2012}).
 

### 2018-04-30 paper/gd1.tex
sec 3.2

--- 2018-04-30 paper/gd1.tex
@@
 stream latitude, $\phi_2$, as a function of stream longitude, $\phi_1$.
-We find stream ridge points by computing a running median in $\phi_2$ in steps
-of $2^\circ$ with a window size of $4^\circ$ along $\phi_1$, then fit a
+In detail, we find stream ridge points by computing a running median in $\phi_2$
+in steps of $2^\circ$ with a window size of $4^\circ$ along $\phi_1$, then fit a
 second-order polynomial to the ridge points to find $\phi_{2,
 \textrm{track}}(\phi_1)$.
-We then define the stream as the region within $\left| \phi_2 - \phi_{2,
+We define the stream as the region within $\left| \phi_2 - \phi_{2,
 \textrm{track}}(\phi_1) \right| < 0.75^\circ$.
@@
 We extract properties of the stream from the region defined above as a function
-of $\phi_1$ by computing median stream properties in overlapping windows:
-we set the window size to $3^\circ$ in $\phi_1$, and shift the window center by
-$1^\circ$ in longitude for each subsequent measurement.
-Panels 3--5 in \figurename~\ref{fig:track-and-model} show the
-background-subtracted stream surface density and median proper motion
-components, $\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way (dark
-lines), with the associated empirical scatters (shaded regions).
-We estimate the local background surface density as an average of areas north
-and south of the bounded stream region.
+of $\phi_1$ by computing stream properties in overlapping windows:
+we set the window size to $4^\circ$ in $\phi_1$, and shift the window center by
+$1^\circ$ in longitude for each measurement.
+Rows 2--5 in \figurename~\ref{fig:track-and-model} show the background-subtracted
+stream surface density, \todo{$\Sigma$?}, the median latitude, $\phi_2$, and median proper motions,
+$\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way, and the associated empirical scatters
+(blue lines and shaded regions).
+We estimate the local background surface density as an average of areas north and south from the bounded stream region.
 
@@
 \caption{%
-Median stream properties computed from the proper motion and CMD filtered GD-1
-stars, and the same for a toy model for the stream track.
-\textit{Panel (a)}: Sky positions of candidate GD-1 stars selected using \gaia\
-proper motions and \pans\ photometry.
-Lines show a 2nd-order polynomial fit to the median stream track offset by $\pm
-0.75^\circ$; the region between these lines is adopted as the nomial stream
-region.
-\textit{Panel (b)}: Sky positions of simulated star particles from a toy mock
-stream model of the GD-1 stream track.
-\textit{Panel (c)}: Background-subtracted surface-density estimates computed in
-running, overlapping $3^\circ$ windows for both GD-1 data and toy model.
-\textit{Panel (d)}: Median proper motion along the stream longitude computed in
-the same bins as previous panel.
-\textit{Panel (e)}: Same as Panel (d), but for proper motion in stream latitude.
+\todo{APW}
 }
@@
 \label{sec:res_gap}
-
-- overdensities apparent beyond the main stream track used in previous section -- if truly associated with the stream, this would be a novel insight into disruption from the \gaia\ mission
-- since some contamination still remains present, it is possible that this is a variation in the MW population, just randomly coming into our selection boxes
-- we thus perform an experiment and compare cmd and proper motions of spur and blob fields to stream and background fields to see which they are associated with
+In this section we test whether the overdensities apparent beyond the main stream track (Figure~\ref{fig:selection}) are truly associated with the GD-1 stream.
+% this would be a novel insight into disruption from the \gaia\ mission
+Some contamination still remains present in our selection, so it is possible that these features are due to the Milky Way population randomly over-populating our selection boxes.
+To assess the origin of these features, we compare color-magnitude diagrams and proper motions of the spur and blob fields to the stream and background fields.
 
@@
 
-- central panel of figure~\ref{fig:features} shows the map of GD-1, with fields of interest in shaded rectangles
-- we first analyze the spur, region, stream same phi1, phi2 below, and an off-stream field in phi2
-- for each of these fields, we show a pair of plots in the top row: cmd on the left and proper motions on the right.
-- we first select on pm, and show cmd, and then select on cmd and show pm
-- in both of these spaces, selection regions used to make the main catalog are shown in shaded orange.
-- stream field (top left) has clear overdensity in the cmd selection box and clusters in proper motion
-- on the other hand, the off field has some stars entering both selection boxes, but without any overdensities in either
-- the spur field has fewer stars than the stream, but XX more than expected from the MW alone
-- furthermore, the distribution in selection boxes is distributed similarly as in the stream field
-- this confirms the association, but we note that in detail they differ
-- median in pm offset from the stream field just below is XX, YY mas yr
-- also dispersion different
-- don't have enough stars to measure distance from the turn-off, so defer to future studies to measure by some other means, e.g., from photometry
+The central panel of Figure~\ref{fig:features} shows the map of GD-1, with fields of interest marked in shades of purple for spur-related fields, and shades of blue for blob-related fields.
+In both cases, we analyze the stream region (darkest shade), overdensity off from the stream (medium shade) and use a field on the opposite side of the stream from the overdensity as a control (lightest shade).
+The $\phi_1$ range is the same for all three fields associated with a given feature, and the control and feature field have the same area, while the stream field is smaller to avoid cross-contamination.
 
-- analogous paragraph for blob on the bottom
+For each of the examined fields, we present the color-magnitude diagram of stars selected on proper motions (left) and proper motions of photometrically selected stars (right), with spur fields on the top and blob fields on the bottom of Figure~\ref{fig:features}.
+In both rows, arranged from left to right are the stream, the feature, and the control field.
+Stars selected from the stream fields show a clear overdensity in the CMD selection box and they cluster in proper motions.
+In detail, the two stream fields have a clear relative offset in proper motions, as expected from the velocity gradients shown in Figure~\ref{fig:track-and-model}.
+Furthermore, the main-sequence turn-off of the $\phi_1\sim-15^\circ$ field appears to be fainter than the $\phi_1\sim-35^\circ$, consistent with the distance gradient across the stream measured by \citet{Koposov:2010}.
+On the other hand, the control fields have some stars entering both the CMD and proper-motion selection boxes, but without any significant overdensities in either.
+CMDs of the spur and blob fields have more stars in the selection box than their respective control fields, and the distribution within the box is qualitatively similar to that of the corresponding stream field, suggesting a similar distance.
+In proper motions, the spur and blob fields also display properties similar to the adjacent stream fields, with clear overdensities of stars in the selection box.
 
-% - Spur vs. blob
-% - Estimate velocity offset and significance
-% - No indication of these features in previous work
-
-% \figurename~\ref{fig:features} (middle panel) again shows the proper-motion and CMD selected stream stars, with interesting features
+Stars from the spur and the blob are similar to those along the main stream track in the color-magnitude diagram and in proper motions, which confirms their association to GD-1.
+The proximity of these off-stream features to large gaps in the stream is suggestive of the gaps being the origin site, further implied by hints of relative velocity differences; for example the median proper motion of the blob is offset from the adjacent stream field by $\approx0.3$\,mas\,yr$^{-1}$ in the positive $\mu_{\phi_1}$ direction.
+We defer detailed modeling of these features to infer their co-evolution with the stream for a future study.
 

### 2018-04-30 paper/gd1.tex
Trim intro

--- 2018-04-30 paper/gd1.tex
@@
 
-Dynamically cold stellar streams are formed during the tidal disruption of stellar
-systems by the gravitational field of their host galaxy.
+Stellar streams are formed during the tidal disruption of stellar systems by the
+gravitational field of their host galaxy.
 The phase-space density and mean track of stars in streams therefore encode
@@
 % \citealt{Erkal:2016a}).
-Being dynamically cold, stellar streams can retain any imprints on its phase-space density from encounters with a massive perturber for close to its total survival time \citep[e.g.,][]{Yoon:2011}.
+Dynamically-cold stellar streams, formed from disrupted globular clusters, can
+also retain imprints on their phase-space density from encounters with a massive
+perturber for close to their total survival time \citep[e.g.,][]{Yoon:2011}.
 % Encounters between massive perturbers and stream stars imprint variations in the
@@
 % Most streams were first detected as a linear overdensity in stellar counts filtered to preferentially select faint blue stars, most likely members of a former Galactic satellite.
-GD-1 is the most prominent among the thin streams, discovered as an arcing overdensity of faint blue stars in the SDSS photometry \citep{Grillmair:2006}.
-It spans $\sim 60^\circ$ on the sky, which at its distance between $\sim 7$--$10~\textrm{kpc}$ translates to a physical length of $\sim 15~\kpc$.
-No remnant progenitor has been associated with the stream, but its
-width ($\sigma \approx 12'$), metallicity ($\feh \approx -1.4$), and estimated
-stellar mass ($M \approx 2 \times 10^4~\msun$)
-imply that GD-1 is a disrupted globular cluster \citep{Koposov:2010}.
+The GD-1 stream is the most prominent among the thin streams, discovered as an
+overdensity of faint blue stars in the SDSS photometry \citep{Grillmair:2006}.
+GD-1 spans $\sim 60^\circ$ on the sky, which, at its distance ($\sim
+7$--$10~\textrm{kpc}$) translates to a physical length of $\sim 15~\kpc$.
+No remnant progenitor for the stream has been found, but its width ($\sigma
+\approx 12'$), metallicity ($\feh \approx -1.4$), and estimated stellar mass ($M
+\approx 2 \times 10^4~\msun$) imply that GD-1 is a disrupted globular cluster
+\citep{Koposov:2010}.
 
-Its length and location in the
-Galactic halo (pericentric distance $r_\textrm{peri} \sim 13~\kpc$) make GD-1 an
-ideal object for constraining the gravitational potential of dark matter in the
-inner Milky Way.
-Indeed, both from fitting orbits to binned phase-space measurements along the
-GD-1 stream (\citealt{Koposov:2010}) and from modeling the stream track in
-phase-space coordinates \citep{Bowden:2015, Bovy:2016}, we learned that the
-dark matter distribution in the inner $\approx20\,\rm kpc$ is consistent with being close to
-spherical.
-This is in line with findings from simulated Milky Way-like galaxies; while dark matter halos are in general expected to be triaxial \citep{Allgood:2006}, the baryonic processes forming galaxies tend to make their inner regions more spherical \citep{Zhu:2016}.
+Its length and location in the Galactic halo (pericentric distance
+$r_\textrm{peri} \sim 13~\kpc$) make GD-1 an ideal object for constraining dark
+matter properties in the inner Milky Way.
+Both from fitting orbits to binned phase-space measurements along the GD-1
+stream (\citealt{Koposov:2010}) and from modeling the stream track in
+phase-space coordinates \citep{Bowden:2015, Bovy:2016}, it appears that the dark
+matter distribution within $\approx20\,\rm kpc$ is close to spherical.
+This is consistent with findings from simulated Milky Way-like galaxies:
+simulated dark-matter-only halos tend to be triaxial \citep{Allgood:2006}, but
+baryonic processes tend to make their inner regions more spherical
+\citep{Zhu:2016}.
 
-An implied old dynamical age, combined with its orbital properties, makes GD-1 a prime stream to not only map the dark matter on large scales, but to also study its small scale structure.
-Density variations in the stream can be induced from interactions with dark matter subhalos \citep[e.g.,][]{Ngan:2014}, but also from interactions with giant molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar \citep{Pearson:2017}.
-However, these baryonic effects are unlikely to have perturbed GD-1 because of its large pericentric distance and retrograde orbit with respect to
-the Galactic bar.
+Its implied old dynamical age, combined with its orbital properties, makes GD-1
+a prime stream to also study the small scale structure of dark matter.
+While density variations in streams can be induced from interactions with giant
+molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar
+\citep{Pearson:2017}, these baryonic effects are unlikely to affect GD-1 because
+of its large pericentric distance and retrograde orbit with respect to the
+Galactic bar.
+Therefore, observed density variations in the GD-1 stream could instead be an
+indication of past interactions with dark matter subhalos
+\citep[e.g.,][]{Ngan:2014}.
 By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
-mass of $10^5~\msun$, and by assuming properties of the subhalo population from dark matter-only simulations \citep[scaled to the Milky Way halo mass of $10^{12}\,\msun$, and reduced in number by a factor of 3 to account for disruption by the Galactic disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
-($\approx 5$--$7^\circ$) gap caused by an interaction with a
-$10^6$--$10^7~\msun$ subhalo.
-A definitive ruling on the presence of dark matter subhalos in this mass regime would directly inform on the nature of dark matter \citep[e.g.][]{Bullock:2017}.
+mass of $10^5~\msun$, and by assuming properties of the subhalo population from
+dark matter-only simulations \citep[scaled to the Milky Way halo mass and
+reduced in number by a factor of 3 to account for disruption by the Galactic
+disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could
+have up to one significant, wide ($\approx 5$--$7^\circ$) gap caused by an
+interaction with a $10^6$--$10^7~\msun$ subhalo.
+A definitive ruling on the presence of dark matter subhalos in this mass regime
+would directly inform on the nature of dark matter \citep[e.g.][]{Bullock:2017}.
 
-Photometric studies of GD-1 revealed the presence of density variations and gaps along the stream, as well as wiggles in the main stream track \citep[][]{Carlberg:2013, DeBoer:2018}.
-Purely photometric stream studies suffer to some extent from contamination of the Milky Way foreground, whose small-scale spatial variations could partially account for density inhomogeneities reported along cold streams \citep[e.g.,][]{Ibata:2016}.
-In this \article, we improve upon the selection of likely GD-1 members by using astrometric data from the \gaia\ mission, in addition to precise photometry from the \pans\ survey (\S\,\ref{sec:data}).
-This data enabled us to measure global properties of GD-1 to an unprecedented precision (\S\,\ref{sec:res_global}), and detect clear signatures of stream stars beyond the main track (\S\,\ref{sec:res_gap}).
-Implications of this revolutionary remapping of GD-1 are discussed in \S\,\ref{sec:discussion}.
+Photometric studies of GD-1 revealed the presence of density variations and gaps
+along the stream, as well as wiggles in the main stream track
+\citep[][]{Carlberg:2013, DeBoer:2018}.
+However, purely photometric studies of streams suffer from contamination of the
+Milky Way foreground, whose small-scale spatial variations could partially
+account for density inhomogeneities reported along cold streams
+\citep[e.g.,][]{Ibata:2016}.
+In this \article, we improve upon the selection of likely GD-1 members by using
+astrometric data from the \gaia\ mission, in addition to precise photometry from
+the \pans\ survey (\S\,\ref{sec:data}).
+These data enable us to measure density variations along GD-1 to an
+unprecedented precision (\S\,\ref{sec:res_global}), and detect clear signatures
+of stream stars beyond the main track (\S\,\ref{sec:res_gap}).
+Implications of this revolutionary remapping of GD-1 are discussed in
+\S\,\ref{sec:discussion}.
 % \todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance

### 2018-04-30 paper/gd1.tex
data section trim

@@
 from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
-selecting all sources with small parallax, $\varpi < 1~\textrm{mas}$, and use a
+selecting sources with small parallax, $\varpi < 1~\textrm{mas}$, and use a
 set of spherical polygons to select batches of stars that have small latitude in
 the heliocentric GD-1 coordinate system, $(\phi_1, \phi_2)$, defined in
-\cite{Koposov:2010}.\footnote{This transformation is implemented using the
-\package{Astropy} (\citealt{astropy}) coordinate transformation framework in the
-\package{Gala} \texttt{Python} package (\citealt{gala}).}
+\cite{Koposov:2010}.
 We convert the sky coordinates and proper motions of the parallax-selected stars
-to the GD-1 stream coordinate system, and correct the proper motions for solar
-reflex motion by assuming that stars at a given stream longitude, $\phi_1$, have
-a distance given by the linear relationship $d(\phi_1) = (0.05 \, \phi_1 +
-10)~\textrm{kpc}$.
-For the solar velocity in a Galactocentric rest frame, we use $\bs{v}_\odot =
-(11.1, 232.24, 7.25)~\kms$ (\citealt{Schonrich:2010, Bovy:2015}).
-\figurename~\ref{fig:selection} (top right) shows the distribution of stars with
-$|\phi_2| < 1^\circ$ in solar-reflex-corrected proper motion components in the
-GD-1 coordinate system, $\mu_{\phi_1, \odot}$ (which includes the $\cos{\phi_2}$
-term) and $\mu_{\phi_2, \odot}$.
-GD-1 member stars are visible as the over-density around
-$(\mu_{\phi_1, \odot}, \mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.
+to the GD-1 stream system, and correct the proper motions for solar reflex
+motion by assuming that stars at a given stream longitude, $\phi_1$, have a
+distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
+we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
+(\citealt{Schonrich:2010, Bovy:2015}).
+\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
+1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
+$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
+GD-1 members are visible as the over-density around $(\mu_{\phi_1, \odot},
+\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.
 
-As a first selection of GD-1 candidate member stars, we select stars with proper
-motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot}
-< 1~\masyr$, as visualized by the orange rectangle in the top right panel of
+As a first selection of GD-1 member stars, we select stars with proper motions
+$-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot} <
+1~\masyr$, as visualized by the rectangle in the top right panel of
 \figurename~\ref{fig:selection}.
 \figurename~\ref{fig:selection} (top left) shows all stars the pass the
-selection on proper motion, plotted in the GD-1 coordinate system.
-From kinematic selection alone, the stream is already identifiable as the over-density
-of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim \phi_1
-\lesssim 0^\circ$.
+selection, plotted in the GD-1 coordinate system.
+From kinematic selection alone, the stream is already identifiable as the
+over-density of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim
+\phi_1 \lesssim 0^\circ$.
 
@@
 To improve the contrast of the stream over the background, we cross-match the
-sample to the \pans\ photometric catalog and use the photometry to further clean
-the sample.
-\figurename~\ref{fig:selection} (bottom right) shows the distribution of
-proper-motion-selected candidate stream stars within $\left|\phi_2\right| <
-1^\circ$ in a \pans\ color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
+sample to the \pans\ photometric catalog and use the photometry to clean the
+sample.
+\figurename~\ref{fig:selection} (bottom right) shows the proper-motion-selected
+candidate stream stars within $\left|\phi_2\right| < 1^\circ$ in a \pans\
+color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
 The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
@@
 $7.8~\kpc$.
-We use this isochrone to motivate a polygonal selection in de-reddened $g-i$
+We use this isochrone to generate a polygonal selection in de-reddened $g-i$
 color and apparent $g$-band magnitude, as visualized by the shaded region in the
@@
 
-\figurename~\ref{fig:selection} (bottom left) shows the final sample of GD-1
-stream member candidates after selecting on both proper motion and photometry.
-The GD-1 stream is clearly visible in the point density of \emph{individual} stars (the
-positions are not binned); this is the most pure view of the stream to date.
-The increased contrast shows that the stream extends at least another $20^\circ$
-to negative longitudes.
-Furthermore, the stream reaches its highest surface density where it is the narrowest ($\phi_1\sim-13^\circ$), so this map might have uncovered the location of its elusive progenitor.
+\figurename~\ref{fig:selection} (bottom left) shows the final sample of
+candidate stream members after selecting on both proper motion and photometry.
+The GD-1 stream is clearly visible in the point density of \emph{individual}
+stars (the positions are not binned); this is the most pure view of the stream
+to date.
+This increased contrast shows that the stream extends at least another
+$20^\circ$ to negative longitudes.
+Furthermore, the stream reaches its highest surface density where it is the
+narrowest ($\phi_1\sim-13^\circ$), which may be the location of its elusive and
+fully-disrupted progenitor.
 Several of the under-densities and gaps hinted at from photometric selection
-alone (\citealt{Koposov:2010, Carlberg:2013}) appear as striking features in
-this significantly cleaner map of the GD-1 stream.
-Finally, at least two new features are visible in this new view of the stream:
-(1) ``the spur,'' stars above the main stream track (in $\phi_2$) between
-$-40^\circ \lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars
-below the main stream track between $-20^\circ \lesssim \phi_1 \lesssim
--10^\circ$.
+(\citealt{Koposov:2010, Carlberg:2013}) appear as striking features in this
+significantly cleaner map of the GD-1 stream.
+Finally, at least two new features are visible near the stream: (1) ``the
+spur,'' stars above the main stream track (in $\phi_2$) between $-40^\circ
+\lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars below the main
+stream track between $-20^\circ \lesssim \phi_1 \lesssim -10^\circ$.
 We discuss each of these in more detail in \sectionname

### 2018-04-30 paper/gd1.tex
overdensity

@@
 $\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
-GD-1 members are visible as the over-density around $(\mu_{\phi_1, \odot},
+GD-1 members are visible as the overdensity around $(\mu_{\phi_1, \odot},
 \mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.
@@
 From kinematic selection alone, the stream is already identifiable as the
-over-density of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim
+overdensity of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim
 \phi_1 \lesssim 0^\circ$.
@@
 to previously reported gaps in the stream.
-There is also a well-defined, sharp over-density in the surface density profile
+There is also a well-defined, sharp overdensity in the surface density profile
 close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
@@
 Here, the proper motion gradient is resolved in striking detail for individual
-stars from \gaia\ (see, e.g., elongated over-density in upper right panel of
+stars from \gaia\ (see, e.g., elongated overdensity in upper right panel of
 \figurename~\ref{fig:selection}).
@@
 We therefore don't expect that the observed surface density variations could be
-caused by, e.g., epicyclic over-densities (\citealt{Kupper:2012}).
+caused by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).
 
@@
 
-The central panel of Figure~\ref{fig:features} shows the map of GD-1, with fields of interest marked in shades of purple for spur-related fields, and shades of blue for blob-related fields.
+The central panel of \figurename~\ref{fig:features} shows the map of GD-1, with fields of interest marked in shades of purple for spur-related fields, and shades of blue for blob-related fields.
 In both cases, we analyze the stream region (darkest shade), overdensity off from the stream (medium shade) and use a field on the opposite side of the stream from the overdensity as a control (lightest shade).
@@
 
-\clearpage
+% \clearpage
 

### 2018-04-30 paper/gd1.tex
trim trim

@@
 To study the global properties of the stream as a function of stream longitude,
-we extract stream stars around a low-order polynomial model for the median
-stream latitude, $\phi_2$, as a function of stream longitude, $\phi_1$.
-We find stream ridge points by computing a running median in $\phi_2$ in steps
-of $2^\circ$ with a window size of $4^\circ$ along $\phi_1$, then fit a
-second-order polynomial to the ridge points to find $\phi_{2,
-\textrm{track}}(\phi_1)$.
-We then define the stream as the region within $\left| \phi_2 - \phi_{2,
+we extract stream stars around a second-order polynomial model for the stream
+latitude, $\phi_2$, as a function of stream longitude, $\phi_1$.
+We find the stream midpoints by computing the median $\phi_2$ in $4^\circ$
+windows shifted by $2^\circ$ along $\phi_1$, then fit a parabola to the median
+positions to find $\phi_{2, \textrm{track}}(\phi_1)$.
+We then define the stream region as $\left| \phi_2 - \phi_{2,
 \textrm{track}}(\phi_1) \right| < 0.75^\circ$.
 \figurename~\ref{fig:track-and-model}, panel (a), shows the high-confidence
-stream stars (same as in lower panel of \figurename~\ref{fig:selection}), and
-the two curved lines (blue) show the adopted upper and lower boundary of the
-stream region.
+stream stars (same as \figurename~\ref{fig:selection}), and the two curved lines
+(blue) show the upper and lower boundary of the stream region.
 
@@
 Panels (c)--(e) in \figurename~\ref{fig:track-and-model} show the
-background-subtracted stream surface density and median proper motion
-components, $\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way (dark
-lines), with the associated empirical scatters (shaded regions).
+background-subtracted stream surface density and median proper motion components
+extracted in this way (dark lines), with the associated empirical scatters
+(shaded regions).
 We estimate the local background surface density as an average of areas north
 and south of the bounded stream region.
+The median proper motion profile along the stream (panels (d)--(e)) are
+consistent with previously measured mean proper motions from SDSS-USNOB
+(\citealt{Koposov:2010}), but here the proper motion gradient is resolved for
+individual stars (see, e.g., elongated overdensity in upper right panel of
+\figurename~\ref{fig:selection}).
 
@@
 extinction or variations in extinction within the stream region.
-For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
-1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
-0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V}
-\approx 0.03~\textrm{mag}$, as computed from the SFD
-extinction map (\citealt{Schlegel:1998}).
-For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
-dust extinction becomes more appreciable, and the stream stars are harder to
-select apart from the background.
+The median $V$-band extinction over the stream track is
+$\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
+$\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
+map (\citealt{Schlegel:1998}).
+% As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
+% dust extinction becomes appreciable, and the stream stars are harder to
+% select apart from the background.
 Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
-periods (median and root-variance of 14 and 3, respectively), guaranteeing a
-robust estimate of proper motions.
+periods (median and root-variance of 14 and 3, respectively), promising a robust
+estimate of the proper motions.
 The variance in the number density of faint stars originating from the \gaia\
@@
 with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
-\mu_{\phi_2, \odot} < 4~\masyr$ (i.e. above the stream selection box), far
-smaller than the density variations in the stream.
+\mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
+stream.
 % The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they
 
-The density variations likely represent real morphological changes along the
-GD-1 stream, as have been suggested before from binned samples of
-color-magnitude-selected GD-1 stars (\citealt{Carlberg:2013, DeBoer:2018}).
+The observed density variations are therefore likely real morphological changes
+along the GD-1 stream, as have been suggested before photometric selection alone
+(\citealt{Carlberg:2013, DeBoer:2018}).
 The deep under-densities we see from the kinematically-cleaned sample correspond
@@
 
-We also measured the median proper motion profile along the stream (panels
-(d)--(e)) and find that these are consistent with previously measured mean
-proper motions from SDSS-USNOB (\citealt{Koposov:2010}).
-Here, the proper motion gradient is resolved in striking detail for individual
-stars from \gaia\ (see, e.g., elongated overdensity in upper right panel of
-\figurename~\ref{fig:selection}).
-
 We compare the measured stream properties to a simple model for the phase-space
-density of the stream, generated by only simulating the orbital evolution of
-cluster stars once they are tidally stripped from the progenitor.
+density of the stream, generated by simulating the orbital evolution of cluster
+stars once they are tidally stripped from the progenitor.
 We use the ``particle-spray'' stream generation method (\citealt{Fardal:2015})
 and assume a uniform mass-loss history to generate the model stream, which
-choices matches results from more detailed ($N$-body) numerical experiments of
-globular cluster disruption \citep[e.g.,][]{Kupper:2012}.
+matches results from more detailed ($N$-body) numerical experiments of globular
+cluster disruption \citep[e.g.,][]{Kupper:2012}.
 % During simulation, star particles are released from the Lagrange points of the
@@
 To compute initial conditions for the stream model, we fit an orbit to the
-observed properties of the stream in a fixed model for the gravitational
-potential of the Milky Way.
-We use a three-component potential model to represent the Galaxy, consisting of
-a disk (\citealt{Miyamoto:1975}), a bulge
-(\citealt{Hernquist:1990}), and a spherical dark matter halo
+observed properties of the stream in a fixed, three-component potential model
+to represent the Milky Way, consisting of a disk (\citealt{Miyamoto:1975}), a
+bulge (\citealt{Hernquist:1990}), and a spherical dark matter halo
 (\citealt{Navarro:1996}).

### 2018-04-30 paper/gd1.tex
fix R

--- 2018-04-30 paper/gd1.tex
@@
 
-Stellar streams are formed during the tidal disruption of stellar systems by the
-gravitational field of their host galaxy.
+Dynamically cold stellar streams are formed during the tidal disruption of stellar
+systems by the gravitational field of their host galaxy.
 The phase-space density and mean track of stars in streams therefore encode
@@
 % \citealt{Erkal:2016a}).
-Dynamically-cold stellar streams, formed from disrupted globular clusters, can
-also retain imprints on their phase-space density from encounters with a massive
-perturber for close to their total survival time \citep[e.g.,][]{Yoon:2011}.
+Being dynamically cold, stellar streams can retain any imprints on its phase-space density from encounters with a massive perturber for close to its total survival time \citep[e.g.,][]{Yoon:2011}.
 % Encounters between massive perturbers and stream stars imprint variations in the
@@
 % Most streams were first detected as a linear overdensity in stellar counts filtered to preferentially select faint blue stars, most likely members of a former Galactic satellite.
-The GD-1 stream is the most prominent among the thin streams, discovered as an
-overdensity of faint blue stars in the SDSS photometry \citep{Grillmair:2006}.
-GD-1 spans $\sim 60^\circ$ on the sky, which, at its distance ($\sim
-7$--$10~\textrm{kpc}$) translates to a physical length of $\sim 15~\kpc$.
-No remnant progenitor for the stream has been found, but its width ($\sigma
-\approx 12'$), metallicity ($\feh \approx -1.4$), and estimated stellar mass ($M
-\approx 2 \times 10^4~\msun$) imply that GD-1 is a disrupted globular cluster
-\citep{Koposov:2010}.
+GD-1 is the most prominent among the thin streams, discovered as an arcing overdensity of faint blue stars in the SDSS photometry \citep{Grillmair:2006}.
+It spans $\sim 60^\circ$ on the sky, which at its distance between $\sim 7$--$10~\textrm{kpc}$ translates to a physical length of $\sim 15~\kpc$.
+No remnant progenitor has been associated with the stream, but its
+width ($\sigma \approx 12'$), metallicity ($\feh \approx -1.4$), and estimated
+stellar mass ($M \approx 2 \times 10^4~\msun$)
+imply that GD-1 is a disrupted globular cluster \citep{Koposov:2010}.
 
-Its length and location in the Galactic halo (pericentric distance
-$r_\textrm{peri} \sim 13~\kpc$) make GD-1 an ideal object for constraining dark
-matter properties in the inner Milky Way.
-Both from fitting orbits to binned phase-space measurements along the GD-1
-stream (\citealt{Koposov:2010}) and from modeling the stream track in
-phase-space coordinates \citep{Bowden:2015, Bovy:2016}, it appears that the dark
-matter distribution within $\approx20\,\rm kpc$ is close to spherical.
-This is consistent with findings from simulated Milky Way-like galaxies:
-simulated dark-matter-only halos tend to be triaxial \citep{Allgood:2006}, but
-baryonic processes tend to make their inner regions more spherical
-\citep{Zhu:2016}.
+Its length and location in the
+Galactic halo (pericentric distance $r_\textrm{peri} \sim 13~\kpc$) make GD-1 an
+ideal object for constraining the gravitational potential of dark matter in the
+inner Milky Way.
+Indeed, both from fitting orbits to binned phase-space measurements along the
+GD-1 stream (\citealt{Koposov:2010}) and from modeling the stream track in
+phase-space coordinates \citep{Bowden:2015, Bovy:2016}, we learned that the
+dark matter distribution in the inner $\approx20\,\rm kpc$ is consistent with being close to
+spherical.
+This is in line with findings from simulated Milky Way-like galaxies; while dark matter halos are in general expected to be triaxial \citep{Allgood:2006}, the baryonic processes forming galaxies tend to make their inner regions more spherical \citep{Zhu:2016}.
 
-Its implied old dynamical age, combined with its orbital properties, makes GD-1
-a prime stream to also study the small scale structure of dark matter.
-While density variations in streams can be induced from interactions with giant
-molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar
-\citep{Pearson:2017}, these baryonic effects are unlikely to affect GD-1 because
-of its large pericentric distance and retrograde orbit with respect to the
-Galactic bar.
-Therefore, observed density variations in the GD-1 stream could instead be an
-indication of past interactions with dark matter subhalos
-\citep[e.g.,][]{Ngan:2014}.
+An implied old dynamical age, combined with its orbital properties, makes GD-1 a prime stream to not only map the dark matter on large scales, but to also study its small scale structure.
+Density variations in the stream can be induced from interactions with dark matter subhalos \citep[e.g.,][]{Ngan:2014}, but also from interactions with giant molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar \citep{Pearson:2017}.
+However, these baryonic effects are unlikely to have perturbed GD-1 because of its large pericentric distance and retrograde orbit with respect to
+the Galactic bar.
 By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
-mass of $10^5~\msun$, and by assuming properties of the subhalo population from
-dark matter-only simulations \citep[scaled to the Milky Way halo mass and
-reduced in number by a factor of 3 to account for disruption by the Galactic
-disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could
-have up to one significant, wide ($\approx 5$--$7^\circ$) gap caused by an
-interaction with a $10^6$--$10^7~\msun$ subhalo.
-A definitive ruling on the presence of dark matter subhalos in this mass regime
-would directly inform on the nature of dark matter \citep[e.g.][]{Bullock:2017}.
+mass of $10^5~\msun$, and by assuming properties of the subhalo population from dark matter-only simulations \citep[scaled to the Milky Way halo mass of $10^{12}\,\msun$, and reduced in number by a factor of 3 to account for disruption by the Galactic disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
+($\approx 5$--$7^\circ$) gap caused by an interaction with a
+$10^6$--$10^7~\msun$ subhalo.
+A definitive ruling on the presence of dark matter subhalos in this mass regime would directly inform on the nature of dark matter \citep[e.g.][]{Bullock:2017}.
 
-Photometric studies of GD-1 revealed the presence of density variations and gaps
-along the stream, as well as wiggles in the main stream track
-\citep[][]{Carlberg:2013, DeBoer:2018}.
-However, purely photometric studies of streams suffer from contamination of the
-Milky Way foreground, whose small-scale spatial variations could partially
-account for density inhomogeneities reported along cold streams
-\citep[e.g.,][]{Ibata:2016}.
-In this \article, we improve upon the selection of likely GD-1 members by using
-astrometric data from the \gaia\ mission, in addition to precise photometry from
-the \pans\ survey (\S\,\ref{sec:data}).
-These data enable us to measure density variations along GD-1 to an
-unprecedented precision (\S\,\ref{sec:res_global}), and detect clear signatures
-of stream stars beyond the main track (\S\,\ref{sec:res_gap}).
-Implications of this revolutionary remapping of GD-1 are discussed in
-\S\,\ref{sec:discussion}.
+Photometric studies of GD-1 revealed the presence of density variations and gaps along the stream, as well as wiggles in the main stream track \citep[][]{Carlberg:2013, DeBoer:2018}.
+Purely photometric stream studies suffer to some extent from contamination of the Milky Way foreground, whose small-scale spatial variations could partially account for density inhomogeneities reported along cold streams \citep[e.g.,][]{Ibata:2016}.
+In this \article, we improve upon the selection of likely GD-1 members by using astrometric data from the \gaia\ mission, in addition to precise photometry from the \pans\ survey (\S\,\ref{sec:data}).
+This data enabled us to measure global properties of GD-1 to an unprecedented precision (\S\,\ref{sec:res_global}), and detect clear signatures of stream stars beyond the main track (\S\,\ref{sec:res_gap}).
+Implications of this revolutionary remapping of GD-1 are discussed in \S\,\ref{sec:discussion}.
 % \todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
@@
 from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
-selecting sources with small parallax, $\varpi < 1~\textrm{mas}$, and use a
+selecting all sources with small parallax, $\varpi < 1~\textrm{mas}$, and use a
 set of spherical polygons to select batches of stars that have small latitude in
 the heliocentric GD-1 coordinate system, $(\phi_1, \phi_2)$, defined in
-\cite{Koposov:2010}.
+\cite{Koposov:2010}.\footnote{This transformation is implemented using the
+\package{Astropy} (\citealt{astropy}) coordinate transformation framework in the
+\package{Gala} \texttt{Python} package (\citealt{gala}).}
 We convert the sky coordinates and proper motions of the parallax-selected stars
-to the GD-1 stream system, and correct the proper motions for solar reflex
-motion by assuming that stars at a given stream longitude, $\phi_1$, have a
-distance given by $d(\phi_1) = (0.05 \, \phi_1 + 10)~\textrm{kpc}$;
-we assume a solar velocity $\bs{v}_\odot = (11.1, 232.24, 7.25)~\kms$
-(\citealt{Schonrich:2010, Bovy:2015}).
-\figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
-1^\circ$ in solar-reflex-corrected proper motion components in the GD-1 system,
-$\mu_{\phi_1, \odot}$ and $\mu_{\phi_2, \odot}$.
-GD-1 members are visible as the over-density around $(\mu_{\phi_1, \odot},
-\mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.
+to the GD-1 stream coordinate system, and correct the proper motions for solar
+reflex motion by assuming that stars at a given stream longitude, $\phi_1$, have
+a distance given by the linear relationship $d(\phi_1) = (0.05 \, \phi_1 +
+10)~\textrm{kpc}$.
+For the solar velocity in a Galactocentric rest frame, we use $\bs{v}_\odot =
+(11.1, 232.24, 7.25)~\kms$ (\citealt{Schonrich:2010, Bovy:2015}).
+\figurename~\ref{fig:selection} (top right) shows the distribution of stars with
+$|\phi_2| < 1^\circ$ in solar-reflex-corrected proper motion components in the
+GD-1 coordinate system, $\mu_{\phi_1, \odot}$ (which includes the $\cos{\phi_2}$
+term) and $\mu_{\phi_2, \odot}$.
+GD-1 member stars are visible as the over-density around
+$(\mu_{\phi_1, \odot}, \mu_{\phi_2, \odot}) \approx (-8, 0)~\masyr$.
 
-As a first selection of GD-1 member stars, we select stars with proper motions
-$-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot} <
-1~\masyr$, as visualized by the rectangle in the top right panel of
+As a first selection of GD-1 candidate member stars, we select stars with proper
+motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot}
+< 1~\masyr$, as visualized by the orange rectangle in the top right panel of
 \figurename~\ref{fig:selection}.
 \figurename~\ref{fig:selection} (top left) shows all stars the pass the
-selection, plotted in the GD-1 coordinate system.
-From kinematic selection alone, the stream is already identifiable as the
-over-density of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim
-\phi_1 \lesssim 0^\circ$.
+selection on proper motion, plotted in the GD-1 coordinate system.
+From kinematic selection alone, the stream is already identifiable as the over-density
+of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim \phi_1
+\lesssim 0^\circ$.
 
@@
 To improve the contrast of the stream over the background, we cross-match the
-sample to the \pans\ photometric catalog and use the photometry to clean the
-sample.
-\figurename~\ref{fig:selection} (bottom right) shows the proper-motion-selected
-candidate stream stars within $\left|\phi_2\right| < 1^\circ$ in a \pans\
-color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
+sample to the \pans\ photometric catalog and use the photometry to further clean
+the sample.
+\figurename~\ref{fig:selection} (bottom right) shows the distribution of
+proper-motion-selected candidate stream stars within $\left|\phi_2\right| <
+1^\circ$ in a \pans\ color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
 The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
@@
 $7.8~\kpc$.
-We use this isochrone to generate a polygonal selection in de-reddened $g-i$
+We use this isochrone to motivate a polygonal selection in de-reddened $g-i$
 color and apparent $g$-band magnitude, as visualized by the shaded region in the
@@
 
-\figurename~\ref{fig:selection} (bottom left) shows the final sample of
-candidate stream members after selecting on both proper motion and photometry.
-The GD-1 stream is clearly visible in the point density of \emph{individual}
-stars (the positions are not binned); this is the most pure view of the stream
-to date.
-This increased contrast shows that the stream extends at least another
-$20^\circ$ to negative longitudes.
-Furthermore, the stream reaches its highest surface density where it is the
-narrowest ($\phi_1\sim-13^\circ$), which may be the location of its elusive and
-fully-disrupted progenitor.
+\figurename~\ref{fig:selection} (bottom left) shows the final sample of GD-1
+stream member candidates after selecting on both proper motion and photometry.
+The GD-1 stream is clearly visible in the point density of \emph{individual} stars (the
+positions are not binned); this is the most pure view of the stream to date.
+The increased contrast shows that the stream extends at least another $20^\circ$
+to negative longitudes.
+Furthermore, the stream reaches its highest surface density where it is the narrowest ($\phi_1\sim-13^\circ$), so this map might have uncovered the location of its elusive progenitor.
 Several of the under-densities and gaps hinted at from photometric selection
-(\citealt{Koposov:2010, Carlberg:2013}) appear as striking features in this
-significantly cleaner map of the GD-1 stream.
-Finally, at least two new features are visible near the stream: (1) ``the
-spur,'' stars above the main stream track (in $\phi_2$) between $-40^\circ
-\lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars below the main
-stream track between $-20^\circ \lesssim \phi_1 \lesssim -10^\circ$.
+alone (\citealt{Koposov:2010, Carlberg:2013}) appear as striking features in
+this significantly cleaner map of the GD-1 stream.
+Finally, at least two new features are visible in this new view of the stream:
+(1) ``the spur,'' stars above the main stream track (in $\phi_2$) between
+$-40^\circ \lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars
+below the main stream track between $-20^\circ \lesssim \phi_1 \lesssim
+-10^\circ$.
 We discuss each of these in more detail in \sectionname
@@
 the mass, assuming an exponential radial density profile with a scale radius
-$\R_d = 3~\kpc$ (an intermediate value of recent measurements, e.g.,
+$\rm R_d = 3~\kpc$ (an intermediate value of recent measurements, e.g.,
 \citealt{Bovy:2012}).

### 2018-04-30 paper/gd1.tex
begin discussion

@@
 
-- novel way to identify members -- helpful that retrograde, but hopefully also applicable to other streams as kinematic important
-- gd-1 even longer -- useful for potential constraints --> future
+In this work we mapped a cold stellar stream in individual tracers for the first time by combining precise \pans\ photometry with the revolutionary \gaia\ data set.
+\gaia\ proper motions were especially helpful for selecting members of the retrograde GD-1 stream, and this success signals a novel way to find members of other structures in the Galactic halo.
+In the rest of this section we discuss how the first view of GD-1 in the \gaia\ era is already transforming what we expect to learn about the Galaxy from streams in general, and GD-1 in particular.
 
-- progenitor? if confirmed, can create more realistic models (so far only modeled the whole stream as just a leading or trailing arm)
-- to confirm -- measure actions, which change direction (some bovy paper)?
-
-- no longer simple stream, discuss possible causes
-- evaporation?
+The absence of its progenitor has been a long-standing hurdle in dynamical modeling of GD-1, mostly due to the unknown true extent of the stream.
+Confirming the progenitor location suggested here, $(\phi_1, \phi_2) = (-13.5^\circ,-0.5^\circ)$, for example by measuring the (canonical) angle variations along the stream \citep{Bovy:2014}, would motivate further searches for stream members along the trailing arm (positive $\phi_1$) to symmetrically match the extent of the leading arm.
+The stream length is directly proportional to the information it provides on the underlying gravitational potential \citep{Bonaca:2018}, making GD-1 (even at its current length of $\sim90^\circ$) a top modeling priority to map the inner 30\,kpc of the Galaxy.
+Properly accounting for the progenitor would also unleash the full constraining power of detailed density structure within the tails, as advocated by \citet{Kupper:2015}.
 
@@
 - if from encounter -- could get more info on the perturber than from the gaps alone, say from delta v
+
+* no longer simple stream, discuss possible causes
 

### 2018-04-30 paper/gd1.tex
abstract tweak

@@
 unprecedented detail.
-In addition to confirming several prominent gaps detected in photometric maps of GD-1, for the first time we find evidence
-of stream members off of the main track.
-These features might be an indication of complex internal kinematics within a progenitor unlike any of the known globular clusters, or a result of a stream interaction with a massive perturber.
+In addition to confirming several prominent gaps detected in photometric maps of
+GD-1, for the first time we find evidence of stream members off of the main
+track.
+These features could be an indication of complex internal kinematics within the
+progenitor unlike any of the known globular clusters, or a result of a stream
+interaction with a massive perturber.
 % Detailed modeling of the positional and velocity offsets

### 2018-04-30 paper/gd1.tex
Captions

@@
 \caption{
-On-sky positions of likely GD-1 members, in the coordinate system aligned with the stream.
-GD-1 is apparent as an overdensity in retrograde proper motions (orange box on the top right panel), so a proper motion selection already reveals the stream in positions of individual stars (top left).
-The stream population is older and more metal poor than the field Milky Way population, so the stream also stands out in the color-magnitude diagram (bottom right).
-Additional photometric selection of the stream's main sequence (orange region in the CMD) unveils the stream in unprecedented detail (bottom left).
-This map extends the stream by $20^\circ$, identifies a plausible location of the progenitor at $\phi_1\sim13^\circ$, confirms the existence of a significant gap at $\phi_1\sim-20^\circ$, shows additional under-densities at $\phi_1\sim-40^\circ,-5^\circ$ and for the first time finds stream overdensities off from the main track of a cold stream at $(\phi_1,\phi_2)\sim(-35^\circ,-1^\circ)$ and $(-15^\circ,-2^\circ)$.
+On-sky positions of likely GD-1 members in the GD-1 coordinate system.
+GD-1 is apparent as an overdensity in negative proper motions (top right panel,
+orange box), so selecting on proper motion already reveals the stream in
+positions of individual stars (top left).
+The stream also stands out in the color-magnitude diagram (bottom right) as
+older and more metal poor than the background.
+Selecting the main sequence of GD-1 (orange, shaded region in bottom right)
+along with proper motion cuts unveils the stream in unprecedented detail (bottom
+left).
+% This map extends the stream by $20^\circ$, identifies a plausible location of the progenitor at $\phi_1\sim13^\circ$, confirms the existence of a significant gap at $\phi_1\sim-20^\circ$, shows additional under-densities at $\phi_1\sim-40^\circ,-5^\circ$ and for the first time finds stream overdensities off from the main track of a cold stream at $(\phi_1,\phi_2)\sim(-35^\circ,-1^\circ)$ and $(-15^\circ,-2^\circ)$.
 }
@@
 stars, and the same for a toy model for the stream track.
-\textit{Panel (a)}: Sky positions of candidate GD-1 stars selected using \gaia\
-proper motions and \pans\ photometry.
+\textit{Panel (a)}: Sky positions of candidate GD-1 stars.
 Lines show a 2nd-order polynomial fit to the median stream track offset by $\pm
-0.75^\circ$; the region between these lines is adopted as the nomial stream
-region.
-\textit{Panel (b)}: Sky positions of simulated star particles from a toy mock
-stream model of the GD-1 stream track.
+0.75^\circ$; the enclosed region is adopted as the nominal stream region.
+\textit{Panel (b)}: Sky positions of simulated star particles from a mock stream
+model of the GD-1 stream.
 \textit{Panel (c)}: Background-subtracted surface-density estimates computed in
-running, overlapping $3^\circ$ windows for both GD-1 data and toy model.
-\textit{Panel (d)}: Median proper motion along the stream longitude computed in
-the same bins as previous panel.
+successive, overlapping $3^\circ$ windows for both GD-1 data and the model
+stream.
+\textit{Panel (d)}: Median proper motion along the stream computed in the same
+bins as previous panel.
 \textit{Panel (e)}: Same as Panel (d), but for proper motion in stream latitude.
@@
 \caption{%
-Best-fitting orbit in a simple (fixed) model for the gravitational potential of
-the Milky Way using phase-space information along the stream measured from this
-work combined with data from \citet{Koposov:2010}.
-Left panel shows the orbit in Galactocentric cylindrical radius, $R$, and height
-above the Galactic midplane, $z$;
-grey line at $z=0$ shows the approximate radius of the disk.
+Best-fit orbit in a simple model for the gravitational potential of the Milky
+Way using \gaia\ proper motions and the filtered stream track.
+Left panel shows the orbit in Galactocentric cylindrical radius, $R$, and
+position above the Galactic midplane, $z$;
+gray line at $z=0$ shows the approximate radius of the disk.
 Right panel shows the orbit projected onto the midplane in Galactocentric
-cartesian coordinates.
-The darker orbit section highlights the range of stream longitudes over which we
-clearly see GD-1 stream stars in \figurename~\ref{fig:selection}.
-Note that GD-1 is retrograde with respect to the disk (see approximate location
-and direction of motion of the sun in right panel, $\odot$).
+cartesian coordinates, ($x$, $y$).
+Darker orbit section highlights the stream longitudes over which we clearly see
+GD-1 stream stars.
+Note that GD-1 is retrograde with respect to the disk (location and direction of
+motion of the sun are indicated in the right panel, $\odot$).
 }

### 2018-04-30 paper/gd1.tex
Trimming

@@
 at a distance of $\sim10~\kpc$.
-This selection extends the stream by $20^\circ$, reveals a plausible location of
-the progenitor, and shows significant density variations along the stream in
-unprecedented detail.
+This selection extends the stream by $20^\circ$ in apparent length, reveals a
+plausible location of the progenitor, and shows significant density variations
+along the stream in unprecedented detail.
 In addition to confirming several prominent gaps detected in photometric maps of
@@
 % \citealt{Erkal:2016a}).
-Dynamically-cold stellar streams, formed from disrupted globular clusters, can
+Dynamically-cold stellar streams formed from disrupted globular clusters can
 also retain imprints on their phase-space density from encounters with a massive
-perturber for close to their total survival time \citep[e.g.,][]{Yoon:2011}.
+perturber for long times \citep[e.g.,][]{Yoon:2011}.
 % Encounters between massive perturbers and stream stars imprint variations in the
@@
 mass of $10^5~\msun$, and by assuming properties of the subhalo population from
-dark matter-only simulations \citep[scaled to the Milky Way halo mass and
-reduced in number by a factor of 3 to account for disruption by the Galactic
-disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could
-have up to one significant, wide ($\approx 5$--$7^\circ$) gap caused by an
-interaction with a $10^6$--$10^7~\msun$ subhalo.
+% dark matter-only simulations \citep[scaled to the Milky Way halo mass and
+% reduced in number by a factor of 3 to account for disruption by the Galactic
+% disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could
+dark matter-only simulations \citep{Springel:2008, Diemand:2008},
+\citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
+($\approx 5$--$7^\circ$) gap caused by an interaction with a
+$10^6$--$10^7~\msun$ subhalo.
 A definitive ruling on the presence of dark matter subhalos in this mass regime
-would directly inform on the nature of dark matter \citep[e.g.][]{Bullock:2017}.
+would directly inform the nature of dark matter \citep[e.g.][]{Bullock:2017}.
 
@@
 However, purely photometric studies of streams suffer from contamination of the
-Milky Way foreground, whose small-scale spatial variations could partially
-account for density inhomogeneities reported along cold streams
-\citep[e.g.,][]{Ibata:2016}.
+Milky Way foreground, which, on small scales, could partially account for
+density inhomogeneities reported along cold streams \citep[e.g.,][]{Ibata:2016}.
 In this \article, we improve upon the selection of likely GD-1 members by using
-astrometric data from the \gaia\ mission, in addition to precise photometry from
+astrometric data from the \gaia\ mission, combined with precise photometry from
 the \pans\ survey (\S\,\ref{sec:data}).
@@
 from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
-selecting sources with small parallax, $\varpi < 1~\textrm{mas}$, and use a
-set of spherical polygons to select batches of stars that have small latitude in
-the heliocentric GD-1 coordinate system, $(\phi_1, \phi_2)$, defined in
-\cite{Koposov:2010}.
+selecting sources with small parallax, $\varpi < 1~\textrm{mas}$, and select
+batches of stars that have small latitude in the heliocentric GD-1 coordinate
+system, $(\phi_1, \phi_2)$, defined in \cite{Koposov:2010}.
 We convert the sky coordinates and proper motions of the parallax-selected stars
@@
 \figurename~\ref{fig:selection}.
-\figurename~\ref{fig:selection} (top left) shows all stars the pass the
+\figurename~\ref{fig:selection} (top left) shows all stars that pass the
 selection, plotted in the GD-1 coordinate system.
-From kinematic selection alone, the stream is already identifiable as the
-overdensity of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim
-\phi_1 \lesssim 0^\circ$.
+The stream is already identifiable as the overdensity of stars around $\phi_2 =
+0$ between longitudes $-60^\circ \lesssim \phi_1 \lesssim 0^\circ$.
 
@@
 To improve the contrast of the stream over the background, we cross-match the
-sample to the \pans\ photometric catalog and use the photometry to clean the
-sample.
+sample to the \pans\ photometric catalog.% and use the photometry to clean the sample.
 \figurename~\ref{fig:selection} (bottom right) shows the proper-motion-selected
@@
 Several of the under-densities and gaps hinted at from photometric selection
-(\citealt{Koposov:2010, Carlberg:2013}) appear as striking features in this
+(\citealt{Carlberg:2013, DeBoer:2018}) appear as striking features in this
 significantly cleaner map of the GD-1 stream.
+% Finally, at least two new features are visible near the stream: (1) ``the
+% spur,'' stars above the main stream track (in $\phi_2$) between $-40^\circ
+% \lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars below the main
+% stream track between $-20^\circ \lesssim \phi_1 \lesssim -10^\circ$.
+% We discuss each of these in more detail in \sectionname
+% s~\ref{sec:results}--\ref{sec:discussion} below.
 Finally, at least two new features are visible near the stream: (1) ``the
-spur,'' stars above the main stream track (in $\phi_2$) between $-40^\circ
-\lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars below the main
-stream track between $-20^\circ \lesssim \phi_1 \lesssim -10^\circ$.
+spur,'' stars above the main stream track (in $\phi_2$), and (2) ``the blob,''
+stars below the main stream track, both highlighted in
+\figurename~\ref{fig:selection}.
 We discuss each of these in more detail in \sectionname
@@
 and south of the bounded stream region.
-The median proper motion profile along the stream (panels (d)--(e)) are
-consistent with previously measured mean proper motions from SDSS-USNOB
+The median proper motion profiles along the stream (panels (d)--(e)) are
+consistent with previous mean proper motion measurements
 (\citealt{Koposov:2010}), but here the proper motion gradient is resolved for
@@
 First, the visible portions of the GD-1 stream are located at high Galactic
-latitudes ($b > 20^\circ$), and we therefore do not expect significant dust
-extinction or variations in extinction within the stream region.
-The median $V$-band extinction over the stream track is
-$\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
-$\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
-map (\citealt{Schlegel:1998}).
+latitudes ($b > 20^\circ$).
+%, and we therefore do not expect significant dust
+% extinction or variations in extinction within the stream region.
+% The median $V$-band extinction over the stream track is
+% $\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
+% $\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
+% map (\citealt{Schlegel:1998}).
 % As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
@@
 estimate of the proper motions.
-The variance in the number density of faint stars originating from the \gaia\
-scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
-with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
-\mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
-stream.
+% The variance in the number density of faint stars originating from the \gaia\
+% scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
+% with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
+% \mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
+% stream.
 % The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they
@@
 This region is also the thinnest part of the stream, as would be expected for
-recently stripped material that has not yet phase-mixed in multiple dimensions.
+recently stripped material.
+% that has not yet phase-mixed in multiple dimensions.
 
 We compare the measured stream properties to a simple model for the phase-space
-density of the stream, generated by simulating the orbital evolution of cluster
-stars once they are tidally stripped from the progenitor.
+density of the stream, generated by simulating the orbital evolution of globular
+cluster stars once they are tidally stripped from the progenitor.
 We use the ``particle-spray'' stream generation method (\citealt{Fardal:2015})
-and assume a uniform mass-loss history to generate the model stream, which
-matches results from more detailed ($N$-body) numerical experiments of globular
-cluster disruption \citep[e.g.,][]{Kupper:2012}.
+and assume a uniform mass-loss history to generate the model stream \citep[e.g.,][]{Kupper:2012}.
+%, which
+% matches results from more detailed ($N$-body) numerical experiments of globular
+% cluster disruption \citep[e.g.,][]{Kupper:2012}.
 % During simulation, star particles are released from the Lagrange points of the
@@
 To compute initial conditions for the stream model, we fit an orbit to the
-observed properties of the stream in a fixed, three-component potential model
-to represent the Milky Way, consisting of a disk (\citealt{Miyamoto:1975}), a
-bulge (\citealt{Hernquist:1990}), and a spherical dark matter halo
-(\citealt{Navarro:1996}).
+observed properties of the stream in a fixed Milky Way model consisting of a
+disk (\citealt{Miyamoto:1975}), a bulge (\citealt{Hernquist:1990}), and a
+spherical dark matter halo (\citealt{Navarro:1996}).
 We set the disk scale-length parameters to match the disk model found in
 \citet{Bovy:2015}; other parameters are set to produce an approximately flat
-disk rotation curve between with a circular velocity at the solar circle,
+disk rotation curve with a circular velocity at the solar circle,
 $v_{\textrm{circ}, \odot} \approx 230~\kms$.
-We use L-BFGS-B optimization as implemented in the \package{scipy}
-\texttt{Python} package to maximize the likelihood of the orbit given sky track
-and proper motions from this work with mean distance and giant star radial
-velocities from \cite{Koposov:2010}.
+%use L-BFGS-B optimization as implemented in the \package{scipy}
+% \texttt{Python} package to
+We maximize the likelihood of the orbit given sky track and proper motions from
+this work, with mean distance and giant star radial velocities from
+\cite{Koposov:2010}.
 We do not vary the parameters of the Milky Way model in the fit because we are
-only interested in producing a comparison model of the stream properties.
+only interested in producing a comparison stream model. % of the stream properties.
 
@@
 
-To initialize the model stream generation, we assume that the present day (i.e.
-final) position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak
-of the observed surface density profile.
-We numerically integrate the backwards from this location for $4~\textrm{Gyr}$
-then forward-generate the model stream using the method discussed above, and
-release stream particles at each timestep.
+To initialize the model stream generation, we assume that the present day
+position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak of the
+observed surface density profile.
+We numerically integrate the backwards from this location for $4~\textrm{Gyr}$,
+then forward-generate the model stream, and release stream particles at each
+timestep.
 We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
@@
 Some contamination still remains present in our selection, so it is possible that these features are due to the Milky Way population randomly over-populating our selection boxes.
-To assess the origin of these features, we compare color-magnitude diagrams and proper motions of the spur and blob fields to the stream and background fields.
+We compare color-magnitude diagrams and proper motions of the spur and blob fields to the stream and background fields.
 
@@
 For each of the examined fields, we present the color-magnitude diagram of stars selected on proper motions (left) and proper motions of photometrically selected stars (right), with spur fields on the top and blob fields on the bottom of Figure~\ref{fig:features}.
-In both rows, arranged from left to right are the stream, the feature, and the control field.
-Stars selected from the stream fields show a clear overdensity in the CMD selection box and they cluster in proper motions.
-In detail, the two stream fields have a clear relative offset in proper motions, as expected from the velocity gradients shown in Figure~\ref{fig:track-and-model}.
+% In both rows, arranged from left to right are the stream, the feature, and the control field.
+Stars selected from the stream fields show a clear overdensity in the CMD selection box and cluster in proper motions.
+The two stream fields have a clear relative offset in proper motions, as expected from the velocity gradients shown in Figure~\ref{fig:track-and-model}.
 Furthermore, the main-sequence turn-off of the $\phi_1\sim-15^\circ$ field appears to be fainter than the $\phi_1\sim-35^\circ$, consistent with the distance gradient across the stream measured by \citet{Koposov:2010}.

### 2018-04-30 paper/gd1.tex
discussion p2

--- 2018-04-30 paper/gd1.tex
@@
 unprecedented detail.
-In addition to confirming several prominent gaps detected in photometric maps of
-GD-1, for the first time we find evidence of stream members off of the main
-track.
-These features could be an indication of complex internal kinematics within the
-progenitor unlike any of the known globular clusters, or a result of a stream
-interaction with a massive perturber.
+In addition to confirming several prominent gaps detected in photometric maps of GD-1, for the first time we find evidence
+of stream members off of the main track.
+These features might be an indication of complex internal kinematics within a progenitor unlike any of the known globular clusters, or a result of a stream interaction with a massive perturber.
 % Detailed modeling of the positional and velocity offsets
@@
 Confirming the progenitor location suggested here, $(\phi_1, \phi_2) = (-13.5^\circ,-0.5^\circ)$, for example by measuring the (canonical) angle variations along the stream \citep{Bovy:2014}, would motivate further searches for stream members along the trailing arm (positive $\phi_1$) to symmetrically match the extent of the leading arm.
-The stream length is directly proportional to the information it provides on the underlying gravitational potential \citep{Bonaca:2018}, making GD-1 (even at its current length of $\sim90^\circ$) a top modeling priority to map the inner 30\,kpc of the Galaxy.
+The stream length is directly proportional to the information it provides on the underlying gravitational potential \citep{Bonaca:2018}, making GD-1 a top modeling priority to map the inner 30\,kpc of the Galaxy, even at the length of $\sim90^\circ$ inferred in this work.
 Properly accounting for the progenitor would also unleash the full constraining power of detailed density structure within the tails, as advocated by \citet{Kupper:2015}.
 
-- spur \& blob as related to gaps? -- classify possible explanations in this way
-- if from encounter -- could get more info on the perturber than from the gaps alone, say from delta v
+GD-1 seen by \gaia\ is no longer a simple, one-dimensional structure.
+Not only are the density variations, tentatively deduced in previous studies, confirmed as highly significant, but for the first time we have mapped tidal debris from a globular cluster beyond a thin stellar track: parts of GD-1 are at a distance of up to $\sim1^\circ$ from the main track.
+Separation of a part of a stellar stream from the main track could be caused by the different exit velocity from the cluster, but explaining the amount of separation and velocity offset in GD-1 would require the progenitor to rotate at XX km\,s$^{-1}$, much faster than measured for the known clusters (cit).
+Features off from the cold part of the stream are likely too small to be caused by chaotic dispersal from either triaxial matter distribution (cit apw) or the time-dependent influence of the rotating bar (cit pearson).
+However, encounters with a massive perturber are expected to open gaps in streams (cits), so we might be seeing the expelled material in the process of phase mixing away from the stream track.
+The encounter origin would be especially exciting -- since GD-1 resides in the halo, the only objects massive enough likely to cross its path are the long-searched-after dark-matter subhalos.
 
-* no longer simple stream, discuss possible causes
+Finding the origin of gaps and off-track features in GD-1 will require progress on two fronts: theoretically, we need to understand the co-evolution of the complex stream system under different origin scenarios, while observing radial velocities of candidate GD-1 members would facilitate data interpretation by remove the remaining contamination and enabling studies in the space of conserved quantities.
+-- sentence on getting more from off-track than gaps alone, but somehow more general and not tied specifically to the encounter scenario / or some other nice parting thought
 
-- progenitor
--- rotation -- could explain spurs, but not gaps?
+% -- radial velocities, measure 3d velocity structure (unique predictions)
+% - if from encounter -- could get more info on the perturber than from the gaps alone, say from delta v -> delta energy
+% - cleaner selection (any stars in the gap?)
+% - prediction: very negative rv, should be easy to map the full extent
 
-- smooth component of the potential?
--- show orbit
--- chaos?
--- \todo{APW} bar -- hard to make small feature, size should be ~angular size of the bar at stream (influence of bar should be a radian of the stream) + relative velocity (retrograde)
-
-- encounters?
--- work out mass / time combinations in impulse approximation
--- comment on GMC vs subhalo
-
-- future:
--- radial velocities, measure 3d velocity structure (unique predictions)
-- delta energy,
-- cleaner selection (any stars in the gap)
-
-- (should go at least to symmetric in positive), not captured with naive selection, as there are gradients
-- prediction: very negative rv, should be easy to map the full extent
-
-% -- tests for other outstanding issues
--- model facets of stream morphology to ascertain if consistent with encounter signatures
--- so far a lot of focus on gaps produced in encounters -- extra features we detected beside the GD-1 track suggest that further theoretical focus on scattered material is needed to fully understand its dynamical history
+% -- model facets of stream morphology to ascertain if consistent with encounter signatures
+% -- so far a lot of focus on gaps produced in encounters -- extra features we detected beside the GD-1 track suggest that further theoretical focus on scattered material is needed to fully understand its dynamical history
 

### 2018-04-30 paper/gd1.tex
title and abstract

--- 2018-04-30 paper/gd1.tex
@@
 \title{Off the beaten path: \\
-       Gaia reveals GD-1 stars perturbed off of the main stream track}
+       Gaia reveals GD-1 stars outside of the main stream}
 % Gaia DR2 reveals stream stars...
@@
 % signatures of dark matter subhalo interactions.
-Here we present a sample of highly-probable members of the longest cold stream, GD-1, selected to have
-small parallaxes and retrograde proper motions using data from the \gaia\
-mission, and \pans\ photometry consistent with an old and metal-poor population
-at a distance of $\sim10~\kpc$.
-This selection extends the stream by $20^\circ$ in apparent length, reveals a
-plausible location of the progenitor, and shows significant density variations
+Using the exquisite data from the \gaia\ second data release combined with
+\pans\ photometry,  we present a sample of highly-probable members of the
+longest cold stream, GD-1, that (1) extend the apparent length of the stream by
+$20^\circ$, (2) reveal a plausible location for the progenitor, and (3) indicate
+the existence of stream members off of the main stream track.
+% selected to have
+% small parallaxes and retrograde proper motions using data from the \gaia\
+% mission, and \pans\ photometry consistent with an old and metal-poor population
+% at a distance of $\sim10~\kpc$.
+% This selection extends the stream by $20^\circ$ in apparent length, reveals a
+% plausible location of the progenitor, and shows significant density variations
+% along the stream in unprecedented detail.
+This new view of the stream reveals significant density variations
 along the stream in unprecedented detail.
-In addition to confirming several prominent gaps detected in photometric maps of
-GD-1, for the first time we find evidence of stream members off of the main
-track.
+% In addition to confirming several prominent gaps detected in photometric maps of
+% GD-1, for the first time we find evidence of stream members off of the main
+% track.
 These features could be an indication of complex internal kinematics within the

### 2018-05-01 paper/gd1.tex
keyword

--- 2018-04-30 paper/gd1.tex
@@
 
-\keywords{Galaxy: halo --- dark matter}
+\keywords{Galaxy: halo --- dark matter --
+          Galaxy: kinematics and dynamics}
 

### 2018-05-01 paper/gd1.tex
Discussion and other edits

@@
 % - Reference for the dustmaps package? See software section.
-
-% Story:
-% - progenitor - show std_v, std_phi2
-% - 2 gaps, 1 under-density - epicycles/streakline? percentiles in polynomial
-% - spur - encounter?
-
-% Make the point that if we can measure distance to spur, stream, RVs - measure energy offset which doesn't depend on time, like gap size
-
-% Make sure to include simple equations so can rescale
-
-% Note that epicyclic density variations naturally in, but caveat uniform release times
 
@@
 this is suggestive of a progenitor system in the final stages of dissolution
-(\todo{cite who}).
+(\citealt{Balbinot:2018}).
 This region is also the thinnest part of the stream, as would be expected for
@@
 \label{sec:res_gap}
-In this section we test whether the overdensities apparent beyond the main stream track (Figure~\ref{fig:selection}) are truly associated with the GD-1 stream.
+
+Here we test whether the overdensities apparent beyond the main stream track
+(\figurename~\ref{fig:selection}) are truly associated with the GD-1 stream by
+comparing color-magnitude diagrams and proper motions of the spur and blob
+fields to the stream and background fields.
 % this would be a novel insight into disruption from the \gaia\ mission
-Some contamination still remains present in our selection, so it is possible that these features are due to the Milky Way population randomly over-populating our selection boxes.
-We compare color-magnitude diagrams and proper motions of the spur and blob fields to the stream and background fields.
+% Some contamination still remains present in our selection, so it is possible that these features are due to the Milky Way population randomly over-populating our selection boxes.
 
@@
 Color-magnitude and proper-motion diagrams of fields associated with a spur above GD-1 (top), and fields related to the blob below GD-1 (bottom).
-The left-most pair of panels show properties in stream fields (darkest rectangles in the middle panel), the middle pairs present spur and blob (medium rectangles), while the right-most panels have fields on the opposite side from spur and blob (lightest rectangles).
-Selection boxes used to select stars shown in the middle panel are shown in orange in all CMD and proper-motion panels.
+The top and bottom left panels show properties in stream fields (darkest rectangles in middle panel), both middle panels present spur and blob (medium rectangles), while the right-most top and bottom panels are selected opposite from spur and blob (lightest rectangles).
+Spatial selection boxes are shown in the middle panel as rectangles, and in all CMD and proper-motion panels as shaded regions.
 Both the spur and the blob fields have more stars in the CMD selection box than the corresponding control fields.
@@
 
-For each of the examined fields, we present the color-magnitude diagram of stars selected on proper motions (left) and proper motions of photometrically selected stars (right), with spur fields on the top and blob fields on the bottom of Figure~\ref{fig:features}.
+For each of the examined fields, we present the color-magnitude diagram of stars selected on proper motions (left subpanels) and proper motions of photometrically selected stars (right subpanels), with spur fields on the top and blob fields on the bottom of \figurename~\ref{fig:features}.
 % In both rows, arranged from left to right are the stream, the feature, and the control field.
-Stars selected from the stream fields show a clear overdensity in the CMD selection box and cluster in proper motions.
-The two stream fields have a clear relative offset in proper motions, as expected from the velocity gradients shown in Figure~\ref{fig:track-and-model}.
+Stars selected from the stream fields show a clear overdensity in the CMD selection box and are clustered in proper motions.
+The two stream fields have a clear relative offset in proper motions, as expected from the velocity gradients shown in \figurename~\ref{fig:track-and-model}.
 Furthermore, the main-sequence turn-off of the $\phi_1\sim-15^\circ$ field appears to be fainter than the $\phi_1\sim-35^\circ$, consistent with the distance gradient across the stream measured by \citet{Koposov:2010}.
+
 On the other hand, the control fields have some stars entering both the CMD and proper-motion selection boxes, but without any significant overdensities in either.
@@
 Stars from the spur and the blob are similar to those along the main stream track in the color-magnitude diagram and in proper motions, which confirms their association to GD-1.
-The proximity of these off-stream features to large gaps in the stream is suggestive of the gaps being the origin site, further implied by hints of relative velocity differences; for example the median proper motion of the blob is offset from the adjacent stream field by $\approx0.3$\,mas\,yr$^{-1}$ in the positive $\mu_{\phi_1}$ direction.
+The proximity of these off-stream features to large gaps in the stream is suggestive of the gaps being their origin, further implied by hints of relative velocity differences.
+For example, the median proper motion of the blob is offset from the adjacent stream field by $\approx0.3$\,mas\,yr$^{-1}$ in the positive $\mu_{\phi_1}$ direction.
 We defer detailed modeling of these features to infer their co-evolution with the stream for a future study.
@@
 
-In this work we mapped a cold stellar stream in individual tracers for the first time by combining precise \pans\ photometry with the revolutionary \gaia\ data set.
-\gaia\ proper motions were especially helpful for selecting members of the retrograde GD-1 stream, and this success signals a novel way to find members of other structures in the Galactic halo.
-In the rest of this section we discuss how the first view of GD-1 in the \gaia\ era is already transforming what we expect to learn about the Galaxy from streams in general, and GD-1 in particular.
+In this work, we map a cold stellar stream in individual stars by combining precise \pans\ photometry with the revolutionary \gaia\ astrometry.
+The \gaia\ proper motions were critical for selecting members of the retrograde GD-1 stream, and this success signals a novel way to find members of other structures in the Galactic halo.
+In what follows, we discuss how this first view of GD-1 in the \gaia\ era is already transforming what we expect to learn about the Galaxy from streams in general, and GD-1 in particular.
 
-The absence of its progenitor has been a long-standing hurdle in dynamical modeling of GD-1, mostly due to the unknown true extent of the stream.
+First, the absence of a clear progenitor has been a long-standing hurdle in dynamical modeling of GD-1, mostly due to the unknown true extent of the stream.
 Confirming the progenitor location suggested here, $(\phi_1, \phi_2) = (-13.5^\circ,-0.5^\circ)$, for example by measuring the (canonical) angle variations along the stream \citep{Bovy:2014}, would motivate further searches for stream members along the trailing arm (positive $\phi_1$) to symmetrically match the extent of the leading arm.
@@
 
-GD-1 seen by \gaia\ is no longer a simple, one-dimensional structure.
-Not only are the density variations, tentatively deduced in previous studies, confirmed as highly significant, but for the first time we have mapped tidal debris from a globular cluster beyond a thin stellar track: parts of GD-1 are at a distance of up to $\sim1^\circ$ from the main track.
-Separation of a part of a stellar stream from the main track could be caused by the different exit velocity from the cluster, but explaining the amount of separation and velocity offset in GD-1 would require the progenitor to rotate at XX km\,s$^{-1}$, much faster than measured for the known clusters (cit).
-Features off from the cold part of the stream are likely too small to be caused by chaotic dispersal from either triaxial matter distribution (cit apw) or the time-dependent influence of the rotating bar (cit pearson).
-However, encounters with a massive perturber are expected to open gaps in streams (cits), so we might be seeing the expelled material in the process of phase mixing away from the stream track.
-The encounter origin would be especially exciting -- since GD-1 resides in the halo, the only objects massive enough likely to cross its path are the long-searched-after dark-matter subhalos.
+Next, GD-1 is no longer a simple, one-dimensional structure (see also \citealt{DeBoer:2018}).
+Not only are the density variations (gaps) confirmed as highly significant, but for the first time we have mapped tidal debris from a globular cluster beyond the thin stellar track: stars from the GD-1 stream have been found up to $\sim1^\circ$ away from the main stream track.
+One explanation for this separated debris could be from velocity substructure within the cluster, but would likely require a fine-tuned orientation to simultaneously explain the sharpness of the neighboring gaps.
+% explaining the amount of separation and velocity offset in GD-1 would require the progenitor to rotate at $\gtrsim 10~\kms$, faster than the measured rotation velocity for any known Milky Way cluster (\citealt{}).
+% APW note: The more I thought about this, I wasn't sure we could make a very strong claim without running or pointing to some stream simulations with rotating progenitors...
+These features are also likely too localized to be explained by chaotic
+dispersal from either triaxialilty (\citealt{Price-Whelan:2016}) or the
+time-dependent influence of the rotating bar (\citealt{Pearson:2017}).
 
-Finding the origin of gaps and off-track features in GD-1 will require progress on two fronts: theoretically, we need to understand the co-evolution of the complex stream system under different origin scenarios, while observing radial velocities of candidate GD-1 members would facilitate data interpretation by remove the remaining contamination and enabling studies in the space of conserved quantities.
--- sentence on getting more from off-track than gaps alone, but somehow more general and not tied specifically to the encounter scenario / or some other nice parting thought
+Encounters between a massive perturber and a thin stream are expected to open gaps in the observed density, and can expel debris into unusual locations and morphologies (e.g., \citealt{Yoon:2011}, see especially \figurename~9).
+The encounter origin would be especially exciting: since GD-1 resides in the halo, the only objects massive enough to cross its path are the long-sought-after dark-matter subhalos.
+
+Confirming the origin of gaps and the off-track features in GD-1 will require
+both detailed modeling of the stream formation in the presence of massive
+perturbers and follow-up radial velocity measurements of candidate GD-1 members.
+Radial velocities will help remove the remaining contamination, further
+improving the contrast of the stream, and will enable measurements of energy and
+angular momentum along the stream and off-track features.
+While constraints on subhalo interactions from individual gaps leave serious
+degeneracies between, e.g., the time of the encounter and the mass of the
+perturber, energy offsets are time-independent and are a promising avenue
+towards placing stronger constraints on the nature of the stream and, if
+confirmed, the particular pertuber.
+
+% theoretically, we need to understand the co-evolution of the complex stream system under different origin scenarios, while observing radial velocities of candidate GD-1 members would facilitate data interpretation by remove the remaining contamination and enabling studies in the space of conserved quantities.
+% -- sentence on getting more from off-track than gaps alone, but somehow more general and not tied specifically to the encounter scenario / or some other nice parting thought
 
@@
 It is a pleasure to thank
+Lauren Anderson,
 Vasily Belokurov,
@@
 Edward Schlafly,
-and David N. Spergel for useful discussions and feedback.
+and David N. Spergel. % for useful discussions and feedback.
 
@@
 at Harvard University.
-All code used in this work and all results are available at
-\url{https://github.com/adrn/GD1-DR2}.
+% All code used in this work and all results are available at
+% \url{https://github.com/adrn/GD1-DR2}.
 }

### 2018-05-01 paper/gd1.tex
S
line

@@
 
-\keywords{Galaxy: halo --- dark matter --
+\keywords{Galaxy: halo --- dark matter ---
           Galaxy: kinematics and dynamics}
@@
 \figurename~\ref{fig:selection}.
-We discuss each of these in more detail in \sectionname
-s~\ref{sec:results}--\ref{sec:discussion} below.
+We discuss each of these in more detail in \S\,
+\ref{sec:results}--\ref{sec:discussion} below.
 

### 2018-05-01 paper/gd1.tex
ab trims

@@
 \begin{abstract}\noindent % trust me
-Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
-streams of stars that are extremely valuable tracers of the large- and
-small-scale distribution of mass in the Galaxy.
-Most stellar streams discovered around the Milky Way reside in the Galactic halo
-and are therefore primarily sensitive to the dark matter distribution.
+Using the exquisite data from the \gaia\ second data release combined with
+\pans\ photometry,  we present a sample of highly-probable members of the
+longest cold stream in the Milky Way -- GD-1.
+The resulting map of GD-1: (1) extends the apparent length of the stream by
+$20^\circ$, (2) reveals a plausible location for the progenitor, (3) detects 
+significant density variations along the stream, and (4) indicates
+the existence of stream members off the main stream track.
+% Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
+% streams of stars that are extremely valuable tracers of the large- and
+% small-scale distribution of mass in the Galaxy.
+% Most stellar streams discovered around the Milky Way reside in the Galactic halo
+% and are therefore primarily sensitive to the dark matter distribution.
 % The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
@@
 % signatures of dark matter subhalo interactions.
-Using the exquisite data from the \gaia\ second data release combined with
-\pans\ photometry,  we present a sample of highly-probable members of the
-longest cold stream, GD-1, that (1) extend the apparent length of the stream by
-$20^\circ$, (2) reveal a plausible location for the progenitor, and (3) indicate
-the existence of stream members off of the main stream track.
 % selected to have
@@
 % along the stream in unprecedented detail.
-This new view of the stream reveals significant density variations
-along the stream in unprecedented detail.
+% This new view of the stream reveals significant density variations
+% along the stream in unprecedented detail.
 % In addition to confirming several prominent gaps detected in photometric maps of
@@
 The phase-space density and mean track of stars in streams therefore encode
-information about the underlying distribution and shape of mass on galactic
+information about the underlying distribution mass on galactic
 scales (e.g., \citealt{Johnston:1999, Bonaca:2018}).
@@
 overdensity of faint blue stars in the SDSS photometry \citep{Grillmair:2006}.
-GD-1 spans $\sim 60^\circ$ on the sky, which, at its distance ($\sim
-7$--$10~\textrm{kpc}$) translates to a physical length of $\sim 15~\kpc$.
+Initially detected to span $\sim 60^\circ$ on the sky, the physical length of GD-1 at a distance of $\sim
+7$--$10~\textrm{kpc}$ is at least $\sim 15~\kpc$.
 No remnant progenitor for the stream has been found, but its width ($\sigma
@@
 A definitive ruling on the presence of dark matter subhalos in this mass regime
-would directly inform the nature of dark matter \citep[e.g.][]{Bullock:2017}.
+would directly inform the nature of dark matter \citep[e.g.,][]{Bullock:2017}.
 
@@
 system, $(\phi_1, \phi_2)$, defined in \cite{Koposov:2010}.
-We convert the sky coordinates and proper motions of the parallax-selected stars
+We convert the equatorial sky coordinates and proper motions %of the parallax-selected stars
 to the GD-1 stream system, and correct the proper motions for solar reflex
@@
 To improve the contrast of the stream over the background, we cross-match the
-sample to the \pans\ photometric catalog.% and use the photometry to clean the sample.
+sample to the \pans\ photometric catalog. % and use the photometry to clean the sample.
 \figurename~\ref{fig:selection} (bottom right) shows the proper-motion-selected
@@
 color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
-The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
-Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
-$12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
-$7.8~\kpc$.
+GD-1 stellar population stands out from the Milky Way foreground, and is well matched by a 12\,Gyr, $\feh = -1.35$ MIST isochrone at a distance of $7.8\,\kpc$ \citep[red line in Figure~\ref{fig:selection}][]{Dotter:2016, Choi:2016, Paxton:2011}.
+% The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
+% Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
+% $12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
+% $7.8~\kpc$.
 We use this isochrone to generate a polygonal selection in de-reddened $g-i$
-color and apparent $g$-band magnitude, as visualized by the shaded region in the
-bottom right panel of \figurename~\ref{fig:selection}.
+color and apparent $g$-band magnitude (shaded region in the
+bottom right panel of \figurename~\ref{fig:selection}).
 
@@
 candidate stream members after selecting on both proper motion and photometry.
-The GD-1 stream is clearly visible in the point density of \emph{individual}
+The GD-1 stream is clearly visible as an overdensity of \emph{individual}
 stars (the positions are not binned); this is the most pure view of the stream
@@
 \figurename~\ref{fig:selection}.
-We discuss each of these in more detail in \S\,
-\ref{sec:results}--\ref{sec:discussion} below.
+% We discuss each of these in more detail in \S\,
+% \ref{sec:results}--\ref{sec:discussion} below.
 
@@
 
-We extract high-confidence stream stars from the proper motion and CMD selected
-sample by further selecting stars close to the main stream track.
-We use these main stream track stars to study the global properties of the
-stream (\S\,\ref{sec:res_global}), and find interesting substructures off of the main stream track (\S\,\ref{sec:res_gap}).
+% We extract high-confidence stream stars from the proper motion and CMD selected
+% sample by further selecting stars close to the main stream track.
+% We use these main stream track stars to study the global properties of the
+% stream (\S\,\ref{sec:res_global}), and find interesting substructures off of the main stream track (\S\,\ref{sec:res_gap}).
 
-\subsection{Global properties}
+\subsection{Properties of the main GD-1 track}
 \label{sec:res_global}
@@
 We extract properties of the stream from the region defined above as a function
-of $\phi_1$ by computing median stream properties in overlapping windows:
-we set the window size to $3^\circ$ in $\phi_1$, and shift the window center by
-$1^\circ$ in longitude for each subsequent measurement.
+of $\phi_1$ by computing median stream properties in (overlapping) $3^\circ$ windows.
+% we set the window size to $3^\circ$ in $\phi_1$, and shift the window center by
+% $1^\circ$ in longitude for each subsequent measurement.
 Panels (c)--(e) in \figurename~\ref{fig:track-and-model} show the
@@
 manifest as sharp features in the surface density estimates along the stream.
-These are not a result of variable dust extinction or completeness variations
-from the \gaia\ scanning pattern.
-First, the visible portions of the GD-1 stream are located at high Galactic
-latitudes ($b > 20^\circ$).
+These do not map to features in dust extinction nor are they caused by completeness variations
+in the \gaia\ scanning pattern, as 99\% of our sample has at least 8 \gaia\ visibility periods, 
+promising a robust estimate of the proper motions.
+% First, the visible portions of the GD-1 stream are located at high Galactic
+% latitudes ($b > 20^\circ$).
 %, and we therefore do not expect significant dust
@@
 % select apart from the background.
-Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
-periods (median and root-variance of 14 and 3, respectively), promising a robust
-estimate of the proper motions.
+% Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
+% periods (median and root-variance of 14 and 3, respectively), promising a robust
+% estimate of the proper motions.
 % The variance in the number density of faint stars originating from the \gaia\
@@
 % The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they
-
 The observed density variations are therefore likely real morphological changes
-along the GD-1 stream, as have been suggested before photometric selection alone
-(\citealt{Carlberg:2013, DeBoer:2018}).
+along the GD-1 stream.
+% as have been suggested before photometric selection alone
+% (\citealt{Carlberg:2013, DeBoer:2018}).
 The deep under-densities we see from the kinematically-cleaned sample correspond
-to previously reported gaps in the stream.
+to previously reported gaps in the stream (\citealt{Carlberg:2013, DeBoer:2018}).
 There is also a well-defined, sharp overdensity in the surface density profile
@@
 
-To compute initial conditions for the stream model, we fit an orbit to the
-observed properties of the stream in a fixed Milky Way model consisting of a
-disk (\citealt{Miyamoto:1975}), a bulge (\citealt{Hernquist:1990}), and a
-spherical dark matter halo (\citealt{Navarro:1996}).
-We set the disk scale-length parameters to match the disk model found in
-\citet{Bovy:2015}; other parameters are set to produce an approximately flat
-disk rotation curve with a circular velocity at the solar circle,
-$v_{\textrm{circ}, \odot} \approx 230~\kms$.
-%use L-BFGS-B optimization as implemented in the \package{scipy}
-% \texttt{Python} package to
-We maximize the likelihood of the orbit given sky track and proper motions from
-this work, with mean distance and giant star radial velocities from
-\cite{Koposov:2010}.
-We do not vary the parameters of the Milky Way model in the fit because we are
-only interested in producing a comparison stream model. % of the stream properties.
-
-\figurename~\ref{fig:orbitfit} shows the best-fitting orbit from the procedure
-outlined above.
-The GD-1 stream crosses the midplane at large radius, far from the bulk of the
-stellar density:
-the grey line in the left panel shows the radius that contains $\approx 90\%$ of
-the mass, assuming an exponential radial density profile with a scale radius
-$\rm R_d = 3~\kpc$ (an intermediate value of recent measurements, e.g.,
-\citealt{Bovy:2012}).
-
 \begin{figure}
@@
 
+To compute initial conditions for the stream model, we fit an orbit to the
+observed properties of the stream in a fixed Milky Way model consisting of a
+disk (\citealt{Miyamoto:1975}), a bulge (\citealt{Hernquist:1990}), and a
+spherical dark matter halo (\citealt{Navarro:1996}).
+We set the disk scale-length parameters to match the disk model found in
+\citet{Bovy:2015}; other parameters are set to produce an approximately flat
+disk rotation curve with a circular velocity at the solar circle,
+$v_{\textrm{circ}, \odot} \approx 230~\kms$.
+%use L-BFGS-B optimization as implemented in the \package{scipy}
+% \texttt{Python} package to
+We maximize the likelihood of the orbit given sky track and proper motions from
+this work, with mean distance and radial velocities from
+\cite{Koposov:2010}.
+% We do not vary the parameters of the Milky Way model in the fit because we are
+% only interested in producing a comparison stream model. % of the stream properties.
+
+\figurename~\ref{fig:orbitfit} shows the best-fitting orbit from the procedure
+outlined above.
+The GD-1 stream crosses the midplane at large radius, far from the bulk of the
+stellar density:
+the grey line in the left panel shows the radius that contains $\approx 90\%$ of
+the mass, assuming a \citet{Bovy:2012} model of the stellar disk.
+% an exponential radial density profile with a scale radius
+% $\rm R_d = 3~\kpc$ (an intermediate value of recent measurements, e.g.,
+% \citealt{Bovy:2012}).
+
 To initialize the model stream generation, we assume that the present day
@@
 We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
-the mass of the progenitor until present day, where we assume $M = 0~\msun$
-(i.e. fully disrupted).
+the mass of the progenitor until its full disruption at the present.
+% day, where we assume $M = 0~\msun$
+% (i.e. fully disrupted).
 Stream properties computed from this resulting model stream are plotted in
@@
 the stream.
-We therefore don't expect that the observed surface density variations could be
-caused by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).
+We therefore don't expect that the observed surface density variations can be fully explained
+by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).
 
@@
 \label{sec:res_gap}
-
-Here we test whether the overdensities apparent beyond the main stream track
-(\figurename~\ref{fig:selection}) are truly associated with the GD-1 stream by
-comparing color-magnitude diagrams and proper motions of the spur and blob
-fields to the stream and background fields.
-% this would be a novel insight into disruption from the \gaia\ mission
-% Some contamination still remains present in our selection, so it is possible that these features are due to the Milky Way population randomly over-populating our selection boxes.
 
@@
 \end{figure}
+
+Here we test whether the overdensities apparent beyond the main stream track
+(\figurename~\ref{fig:selection}) are truly associated with the GD-1 stream by
+comparing color-magnitude diagrams and proper motions of the spur and blob
+fields to the stream and background fields.
+% this would be a novel insight into disruption from the \gaia\ mission
+% Some contamination still remains present in our selection, so it is possible that these features are due to the Milky Way population randomly over-populating our selection boxes.
 

### 2018-05-01 paper/gd1.tex
further trims

@@
 \begin{abstract}\noindent % trust me
+Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
+streams of stars that are extremely valuable tracers of the large- and
+small-scale distribution of mass in the Galaxy.
 Using the exquisite data from the \gaia\ second data release combined with
@@
 the existence of stream members off the main stream track.
-% Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
-% streams of stars that are extremely valuable tracers of the large- and
-% small-scale distribution of mass in the Galaxy.
+The additional length and proper treatment of the progenitor will aid in dynamical modeling of GD-1 and map the Galaxy on large scales, while its complex morphology points to a turbulent history, potentially informing on massive, small-scale perturbers within 30\,kpc.
+
 % Most stellar streams discovered around the Milky Way reside in the Galactic halo
@@
 phase-space coordinates \citep{Bowden:2015, Bovy:2016}, it appears that the dark
-matter distribution within $\approx20\,\rm kpc$ is close to spherical.
-This is consistent with findings from simulated Milky Way-like galaxies:
-simulated dark-matter-only halos tend to be triaxial \citep{Allgood:2006}, but
-baryonic processes tend to make their inner regions more spherical
-\citep{Zhu:2016}.
+matter distribution within $\approx20\,\rm kpc$ is close to spherical, consistent 
+with findings from simulated Milky Way-like galaxies \citep{Zhu:2016}.
+% This is consistent with findings from simulated Milky Way-like galaxies:
+% simulated dark-matter-only halos tend to be triaxial \citep{Allgood:2006}, but
+% baryonic processes tend to make their inner regions more spherical
+% \citep{Zhu:2016}.
 
@@
 \citep[e.g.,][]{Ngan:2014}.
-By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
-mass of $10^5~\msun$, and by assuming properties of the subhalo population from
-% dark matter-only simulations \citep[scaled to the Milky Way halo mass and
-% reduced in number by a factor of 3 to account for disruption by the Galactic
-% disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could
-dark matter-only simulations \citep{Springel:2008, Diemand:2008},
-\citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
+% By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
+% mass of $10^5~\msun$, and by assuming properties of the subhalo population from
+% dark matter-only simulations \citep{Springel:2008, Diemand:2008},
+\citet{Erkal:2016} predict that a 4\,Gyr old GD-1-like stream in a Milky Way-like galaxy
+could have up to one significant, wide
 ($\approx 5$--$7^\circ$) gap caused by an interaction with a
@@
 would directly inform the nature of dark matter \citep[e.g.,][]{Bullock:2017}.
+% dark matter-only simulations \citep[scaled to the Milky Way halo mass and
+% reduced in number by a factor of 3 to account for disruption by the Galactic
+% disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could
 
@@
 from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
-selecting sources with small parallax, $\varpi < 1~\textrm{mas}$, and select
-batches of stars that have small latitude in the heliocentric GD-1 coordinate
-system, $(\phi_1, \phi_2)$, defined in \cite{Koposov:2010}.
+selecting sources with small parallax, $\varpi < 1~\textrm{mas}$ in fields along the GD-1 stream
+% , and select
+% batches of stars that have small latitude in the heliocentric GD-1 coordinate
+% system, $(\phi_1, \phi_2)$, defined in \cite{Koposov:2010}.
 We convert the equatorial sky coordinates and proper motions %of the parallax-selected stars
-to the GD-1 stream system, and correct the proper motions for solar reflex
+to the GD-1 coordinate system \citep[$\phi_1, \phi_2$,][]{Koposov:2010}, and correct the proper motions for solar reflex
 motion by assuming that stars at a given stream longitude, $\phi_1$, have a
@@
 
-As a first selection of GD-1 member stars, we select stars with proper motions
+As a first selection of GD-1 member stars, we choose stars with proper motions
 $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot} <
-1~\masyr$, as visualized by the rectangle in the top right panel of
-\figurename~\ref{fig:selection}.
+1~\masyr$ (rectangle in the top right panel of
+\figurename~\ref{fig:selection}).
 \figurename~\ref{fig:selection} (top left) shows all stars that pass the
@@
 color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
-GD-1 stellar population stands out from the Milky Way foreground, and is well matched by a 12\,Gyr, $\feh = -1.35$ MIST isochrone at a distance of $7.8\,\kpc$ \citep[red line in Figure~\ref{fig:selection}][]{Dotter:2016, Choi:2016, Paxton:2011}.
+GD-1 stellar population stands out from the Milky Way foreground, and is well matched by a 12\,Gyr, $\feh = -1.35$ MIST isochrone at a distance of $7.8\,\kpc$ \citep[red line in Figure~\ref{fig:selection},][]{Dotter:2016, Choi:2016, Paxton:2011}.
 % The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
@@
 To compute initial conditions for the stream model, we fit an orbit to the
-observed properties of the stream in a fixed Milky Way model consisting of a
-disk (\citealt{Miyamoto:1975}), a bulge (\citealt{Hernquist:1990}), and a
-spherical dark matter halo (\citealt{Navarro:1996}).
-We set the disk scale-length parameters to match the disk model found in
-\citet{Bovy:2015}; other parameters are set to produce an approximately flat
-disk rotation curve with a circular velocity at the solar circle,
-$v_{\textrm{circ}, \odot} \approx 230~\kms$.
+observed properties of the stream in a fixed Milky Way model similar to \citet{Bovy:2015}.
+% consisting of a
+% disk (\citealt{Miyamoto:1975}), a bulge (\citealt{Hernquist:1990}), and a
+% spherical dark matter halo (\citealt{Navarro:1996}).
+% We set the disk scale-length parameters to match the disk model found in
+% \citet{Bovy:2015}; other parameters are set to produce an approximately flat
+% disk rotation curve with a circular velocity at the solar circle,
+% $v_{\textrm{circ}, \odot} \approx 230~\kms$.
 %use L-BFGS-B optimization as implemented in the \package{scipy}
@@
 The top and bottom left panels show properties in stream fields (darkest rectangles in middle panel), both middle panels present spur and blob (medium rectangles), while the right-most top and bottom panels are selected opposite from spur and blob (lightest rectangles).
-Spatial selection boxes are shown in the middle panel as rectangles, and in all CMD and proper-motion panels as shaded regions.
+Selection boxes are shown in all CMD and proper-motion panels as shaded regions.
 Both the spur and the blob fields have more stars in the CMD selection box than the corresponding control fields.

### 2018-05-01 paper/gd1.tex
abstract edit
fig1 annotations

@@
 significant density variations along the stream, and (4) indicates
-the existence of stream members off the main stream track.
-The additional length and proper treatment of the progenitor will aid in dynamical modeling of GD-1 and map the Galaxy on large scales, while its complex morphology points to a turbulent history, potentially informing on massive, small-scale perturbers within 30\,kpc.
+the existence of stream members perturbed off the main stream track.
+\gaia\
+The additional length and proper treatment of the progenitor will aid in dynamical modeling of GD-1 and map the Galaxy on large scales.
+The complex morphology of the stream points to a turbulent history; the detailed morphology in phase space of the perturbed stream members could potentially provide constraints on dark matter substructure that might have interacted with the stream.
 
@@
 % track.
-These features could be an indication of complex internal kinematics within the
-progenitor unlike any of the known globular clusters, or a result of a stream
-interaction with a massive perturber.
+% These features could be an indication of complex internal kinematics within the
+% progenitor unlike any of the known globular clusters, or a result of a stream
+% interaction with a massive perturber.
 % Detailed modeling of the positional and velocity offsets
@@
 It is a pleasure to thank
+Marla Geha,
+David W. Hogg,
+Kathryn V. Johnston;
 Lauren Anderson,
@@
 Andrew R. Casey,
-Marla Geha,
-David W. Hogg,
 Benjamin D. Johnson,
-Kathryn V. Johnston,
 Sergey Koposov,

### 2018-05-01 paper/gd1.tex
abstract including dwh comments

@@
 small-scale distribution of mass in the Galaxy.
-Using the exquisite data from the \gaia\ second data release combined with
+Using data from the \gaia\ second data release combined with
 \pans\ photometry, we present a sample of highly-probable members of the
@@
 $20^\circ$, (2) reveals a plausible location for the progenitor, (3) detects 
-significant density variations along the stream, and (4) indicates
+high-contrast gaps along the stream, and (4) indicates
 the existence of stream members perturbed off the main stream track.
-\gaia\
+These discoveries were only possible because the exquisite proper motions from \gaia\ permit a clean separation of the stream from the Milky Way stars.
 The additional length and proper treatment of the progenitor will aid in dynamical modeling of GD-1 and map the Galaxy on large scales.
-The complex morphology of the stream points to a turbulent history; the detailed morphology in phase space of the perturbed stream members could potentially provide constraints on dark matter substructure that might have interacted with the stream.
+The complex morphology of the stream points to a turbulent history; the detailed phase-space of the perturbed stream members could potentially constrain dark matter substructure in the Milky Way.
 
@@
 
-Over 30 candidate stellar streams have been discovered in the Milky Way \citep[see][for a recent review]{Grillmair:2016, Newberg:2016}.
+Over 30 candidate stellar streams have been discovered in the Milky Way \citep{Grillmair:2016, Newberg:2016, Malhan:2018}.
 % thanks to large-area, multi-band photometric surveys (especially

### 2018-05-01 paper/gd1.tex
tweaks

@@
 The additional length and proper treatment of the progenitor will aid in dynamical modeling of GD-1 and map the Galaxy on large scales.
-The complex morphology of the stream points to a turbulent history; the detailed phase-space of the perturbed stream members could potentially constrain dark matter substructure in the Milky Way.
+The complex morphology of the stream points to a turbulent history; detailed phase-space properties of the perturbed stream members could potentially constrain dark matter substructure in the Milky Way.
 
@@
 positions to find $\phi_{2, \textrm{track}}(\phi_1)$.
-We then define the stream region as $\left| \phi_2 - \phi_{2,
+We define the stream region as $\left| \phi_2 - \phi_{2,
 \textrm{track}}(\phi_1) \right| < 0.75^\circ$.

### 2018-05-01 paper/gd1.tex
Minor tweaks

@@
 \pans\ photometry, we present a sample of highly-probable members of the
-longest cold stream in the Milky Way -- GD-1.
+longest cold stream in the Milky Way, GD-1.
 The resulting map of GD-1: (1) extends the apparent length of the stream by
@@
 the existence of stream members perturbed off the main stream track.
-These discoveries were only possible because the exquisite proper motions from \gaia\ permit a clean separation of the stream from the Milky Way stars.
+These discoveries were only possible because of the exquisite proper motions
+from \gaia, which permit a clean separation of the stream from the Milky Way
+stars.
 The additional length and proper treatment of the progenitor will aid in dynamical modeling of GD-1 and map the Galaxy on large scales.

### 2018-05-01 paper/gd1.tex
width

@@
 This region is also the thinnest part of the stream, as would be expected for
-recently stripped material.
+recently stripped material, but we defer a robust analysis of variations in the
+stream width to future work.
 % that has not yet phase-mixed in multiple dimensions.
@@
 the stream.
+The broadening in the observed stream at negative longitudes is also seen in the model stream and is expected for older parts of the stream that have phase-mixed for longer times.
 We therefore don't expect that the observed surface density variations can be fully explained

### 2018-05-01 paper/gd1.tex
trimmy

@@
 the existence of stream members perturbed off the main stream track.
-These discoveries were only possible because of the exquisite proper motions
-from \gaia, which permit a clean separation of the stream from the Milky Way
+These discoveries are only possible because of the exquisite astrometry
+from \gaia, which permits a clean separation of the stream from Milky Way
 stars.
-The additional length and proper treatment of the progenitor will aid in dynamical modeling of GD-1 and map the Galaxy on large scales.
+The additional length and a proper treatment of the progenitor will aid in dynamical modeling of GD-1 for mapping the large-scale dark matter distribution.
 The complex morphology of the stream points to a turbulent history; detailed phase-space properties of the perturbed stream members could potentially constrain dark matter substructure in the Milky Way.
@@
 
-We retrieve \gaia\ data along the previously-identified track of the GD-1 stream
+We retrieve data along the previously-identified track of the GD-1 stream
 from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
@@
 color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
-GD-1 stellar population stands out from the Milky Way foreground, and is well matched by a 12\,Gyr, $\feh = -1.35$ MIST isochrone at a distance of $7.8\,\kpc$ \citep[red line in Figure~\ref{fig:selection},][]{Dotter:2016, Choi:2016, Paxton:2011}.
+The GD-1 stellar population stands out from the Milky Way foreground, and is well matched by a 12\,Gyr, $\feh = -1.35$ MIST isochrone at a distance of $7.8\,\kpc$ \citep[red line in \figurename~\ref{fig:selection},][]{Dotter:2016, Choi:2016, Paxton:2011}.
 % The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
@@
 manifest as sharp features in the surface density estimates along the stream.
-These do not map to features in dust extinction nor are they caused by completeness variations
-in the \gaia\ scanning pattern, as 99\% of our sample has at least 8 \gaia\ visibility periods,
-promising a robust estimate of the proper motions.
+We have verified that these do not map to features in dust extinction nor are
+they caused by completeness variations in the \gaia\ scanning pattern:
+the visible portions of the GD-1 stream are located at high Galactic latitudes
+($b > 20^\circ$), and 99\% of our sample has at least 8 \gaia\ visibility
+periods, promising a robust astrometric solution.
 % First, the visible portions of the GD-1 stream are located at high Galactic
@@
 This region is also the thinnest part of the stream, as would be expected for
-recently stripped material, but we defer a robust analysis of variations in the
+recently stripped material; we defer a robust analysis of variations in the
 stream width to future work.
@@
 observed surface density profile.
-We numerically integrate the backwards from this location for $4~\textrm{Gyr}$,
-then forward-generate the model stream, and release stream particles at each
+We numerically integrate backwards from this location for $4~\textrm{Gyr}$,
+then forward-generate the model stream, releasing stream particles at each
 timestep.
@@
 In both cases, we analyze the stream region (darkest shade), overdensity off from the stream (medium shade) and use a field on the opposite side of the stream from the overdensity as a control (lightest shade).
-The $\phi_1$ range is the same for all three fields associated with a given feature, and the control and feature field have the same area, while the stream field is smaller to avoid cross-contamination.
+The $\phi_1$ range is the same for all three fields for a given feature, and the control and feature fields have the same area (the stream field is smaller to avoid cross-contamination).
 
@@
 
-On the other hand, the control fields have some stars entering both the CMD and proper-motion selection boxes, but without any significant overdensities in either.
+The control fields have some stars in both the CMD and proper-motion selection boxes, but without any significant overdensities in either.
 CMDs of the spur and blob fields have more stars in the selection box than their respective control fields, and the distribution within the box is qualitatively similar to that of the corresponding stream field, suggesting a similar distance.
@@
 For example, the median proper motion of the blob is offset from the adjacent stream field by $\approx0.3$\,mas\,yr$^{-1}$ in the positive $\mu_{\phi_1}$ direction.
-We defer detailed modeling of these features to infer their co-evolution with the stream for a future study.
+We defer detailed modeling of these features to infer their co-evolution with the stream for future work.
 
@@
 First, the absence of a clear progenitor has been a long-standing hurdle in dynamical modeling of GD-1, mostly due to the unknown true extent of the stream.
-Confirming the progenitor location suggested here, $(\phi_1, \phi_2) = (-13.5^\circ,-0.5^\circ)$, for example by measuring the (canonical) angle variations along the stream \citep{Bovy:2014}, would motivate further searches for stream members along the trailing arm (positive $\phi_1$) to symmetrically match the extent of the leading arm.
+Confirming the progenitor location suggested here, $(\phi_1, \phi_2) =
+(-13.5^\circ,-0.5^\circ)$, would motivate further searches for stream members
+along the trailing arm (positive $\phi_1$) to symmetrically match the extent of
+the leading arm.
+% for example measuring the (canonical) angle variations along the stream \citep{Bovy:2014}
 The stream length is directly proportional to the information it provides on the underlying gravitational potential \citep{Bonaca:2018}, making GD-1 a top modeling priority to map the inner 30\,kpc of the Galaxy, even at the length of $\sim90^\circ$ inferred in this work.
-Properly accounting for the progenitor would also unleash the full constraining power of detailed density structure within the tails, as advocated by \citet{Kupper:2015}.
+Properly accounting for the progenitor would also allow for studying detailed density structure within the tails, as advocated by \citet{Kupper:2015}.
 
@@
 Radial velocities will help remove the remaining contamination, further
-improving the contrast of the stream, and will enable measurements of energy and
-angular momentum along the stream and off-track features.
+improving the contrast of the stream, and will enable measurements of
+differential energy and angular momentum along the stream and off-track
+features.
 While constraints on subhalo interactions from individual gaps leave serious

### 2018-05-01 paper/gd1.tex
rephrase fig4 caption

@@
 Color-magnitude and proper-motion diagrams of fields associated with a spur above GD-1 (top), and fields related to the blob below GD-1 (bottom).
-The top and bottom left panels show properties in stream fields (darkest rectangles in middle panel), both middle panels present spur and blob (medium rectangles), while the right-most top and bottom panels are selected opposite from spur and blob (lightest rectangles).
+Panels on the left show properties in stream fields (darkest rectangles in middle panel), middle panels present spur and blob (medium rectangles), and the right-most panels are control fields on the opposite sides from spur and blob (lightest rectangles).
+% The top and bottom left panels show properties in stream fields (darkest rectangles in middle panel), middle panels present spur and blob (medium rectangles), while the right-most top and bottom panels are selected opposite from spur and blob (lightest rectangles).
 Selection boxes are shown in all CMD and proper-motion panels as shaded regions.

### 2018-05-02 paper/gd1.tex
fixed typo

@@
 from the \gaia\ science archive\footnote{\url{https://gea.esac.esa.int/}} by
-selecting sources with small parallax, $\varpi < 1~\textrm{mas}$ in fields along the GD-1 stream
+selecting sources with small parallax, $\varpi < 1~\textrm{mas}$ in fields along the GD-1 stream.
 % , and select

### 2018-05-27 paper/submit-aas/referee-response.txt
response to ref

We thank the referee for a constructive report that has improved the quality of
this work. Below, we have copied the remarks from the referee (indented by >)
and have responded to each point in text that immediately follows.

> The analysis of the Gaia DR2 data requires some care, especially at the faint
> end. For example, the authors do not discuss the amplitude of the correlations
> between the proper motion components and how these may affect their selection
> of tentative members. (By the way, the plots show relatively large ranges,
> particularly those of the proper motions and it would be useful if a few
> zoom-ins are included, so that the reader can assess better the data.).
>
> It would also be useful if the authors would explore how many stars in the
> regions around the stream have 2 parameter instead of 5 parameter solutions
> (i.e. they have position measured on the sky but no proper motions).
>
> Although the authors comment that all the stars have 8 or more visibility
> periods, this does not reflect completeness of the dataset but merely the
> reliability of the astrometric solution. Therefore maps showing the quality of
> the astrometric data in this region of the sky should be included as they
> enhance the credibility of the various features the authors claim to find. As
> a guide to what are the relevant diagnostics, the authors may wish to use
> Arenou et a. 2018 or Lindegren et al 2018.
>
> The presence of an overdensity that could correspond to the location of the
> progenitor is not so clear from the plots shown in the paper. Again a zoom-in
> might help here.

* APW:
* TODO: final figure summarize data quality and dust
* Push back on Zoom-ins? -> plot them for the referee, don't add much
* TODO: send Ana file with extinction estimates

> The authors need to include a measurement of the proper motion in the
> equatorial reference frame, for example, at the location of the highest
> density of the stream. This is clearly missing from the paper, and would make
> the paper useful to other readers.

* APW: easy, but stupid

> The authors do not use the Gaia photometry, which although shallower is likely
> to be significantly more accurate. How do the CMDs look like in the Gaia
> bands?

* AB: will check CMD
* Respond with plot and comment
* Main sequence becomes flatter

> The authors do not re-derive a distance to the stream, is there a reason for
> this? It would perhaps be useful to overlay the isochrone on the Gaia data to
> try to determine the distance to GD-1 more accurately, and hence also the
> tangential velocities of the stars in the stream.

* AB: Punt to next paper: respond that distance not the focus

> It was not clear to me how the solar reflex motion was corrected via the
> equation given, where does this stem from?

We have added a sentence to clarify how this was done.

> The significance of the various features (the blob, the spur) needs to be
> quantified. For example, Fig.4 seems to show a small overdensity in the "off
> blob" region on the proper motion diagram but there is no mention of this.

* APW/AB: Define significance

> The authors state that the proximity of the features to the location of the
> large gaps suggests that they might be related. Please elaborate on how.

* AB: explain the reference to Yoon et al., be explicit about the possible
  connection to the gap

> The orbit shown in Fig.3 is not integrated sufficiently long to be able to
> establish whether the stream has crossed the disc. Please include a comparison
> to an orbit integrated for 6-8 Gyr.

The orbit shown in Figure 3 crosses the disk midplane (z=0) twice in the segment of the orbit shown. Given uncertainties in the Milky Way's gravitational potential, and uncertainties in the full-space motion of the stream progenitor, we don't think that including a 6-8 Gyr integration in the paper will be too meaningful. However, we have extended the integration time for the paper figure to 1 Gyr, and attach a figure with a 6 Gyr integration with this response for the referee to see.

* Comments from the AAS statistics editor:

> In Fig 2, the authors choose a parabolic model for the GD-1 stream with
> (arbitrary?) lower/upper boundaries. This band clearly includes some
> background stars, and may miss some stream stars around \phi_1=-30d. It would
> be nice to identify stream members by some statistical technique, either in
> the 2D sky or 4D sky+kinematics space.

> I suggest the authors consult the excellent summary of modern `spatial point
> processes' methodology with cookbook to the comprehensive 'spatstat' R/CRAN
> package, by A.Baddeley et al. 'Spatial Point Patterns: Methodology and
> Applications with R' (2015). For example, I suspect a good list of stream
> members could emerge from a Voronoi tessellation with truncation of tile
> areas. Other possibilities include: nearest neighbors within a distance
> threshold, nonparametric density estimation, Allard-Fraley feature detection
> (MLE on Voronoi tessellation), and Byers-Raftery feature detection (MLE
> mixture model on k-NN distribution).

* APW: simple statistical technique in 4D, punt

### 2018-05-28 paper/submit-aas/referee-response.txt
sketch distance, cmd responses

@@
 * AB: Punt to next paper: respond that distance not the focus
+- we do lay an isochrone in fig 1, consistent with a previously reported distance and stellar population
+- however, we note that fig 4 indicates a distance gradient, closer at trailing than leading tail, also consistent with the literature
+- while re-deriving a distance to the stream using would be useful, it is rather involved and would need to be done along the stream, so we are leaving it for our next contribution where we will derive more general properties of GD-1, while focusing on the new features in this one
 

### 2018-05-31 paper/submit-aas/referee-response.txt
working on completeness question

@@
 * Push back on Zoom-ins? -> plot them for the referee, don't add much
-* TODO: send Ana file with extinction estimates
+
+- Dust map done
+- 2-parameter solution done
+- Source density at phi1 = [-40, -20, 0], phi2 = 0 is ~8000 sources/deg^2 -- should be close to 100% complete according to Arenou.
+https://arxiv.org/pdf/1804.09375.pdf
+https://arxiv.org/pdf/1804.09366.pdf
 

### 2018-05-31 paper/submit-aas/referee-response.txt
cmd response

--- 2018-05-28 paper/submit-aas/referee-response.txt
@@
 
-* AB: will check CMD
-* Respond with plot and comment
-* Main sequence becomes flatter
+The referee is correct that the Gaia photometry is significantly more accurate than PanSTARRS, however, at faint magnitudes, the regime where GD-1 members are most easily separable from the Milky Way field stars, the Gaia BP-RP color is much more uncertain than the PanSTARRS g-i color.
+As a part of the response, we attach a figure that compares Gaia (left) and PanSTARRS (right) color-magnitude diagrams of GD-1 stars, selected with -1<phi_2<1, -50<phi_1<-10, and our fiducial proper motion selection.
+Typical photometric uncertainties are shown as a function of magnitude, centered on 0 color in both panels.
+PanSTARRS photometry is systematics dominated in the entire range, and the uncertainty in both g and i bands is 0.01 mag (0.014 mag for g-i color). 
+Uncertainties in Gaia's G band are smaller in the entire range, but BP-RP's vary from 0.001 at G~15 to 0.05 at G~20, approximately 5 times larger than PanSTARRS at comparably faint magnitudes.
+This leads to GD-1's main sequence being much better defined in PanSTARRS color-magnitude diagram than in Gaia's, so we opt to retain PanSTARRS photometry for selection of likely GD-1 members.
+
+Q: do we comment on using g vs G?
 

### 2018-06-03 paper/submit-aas/referee-response.txt
distance response

@@
 
-The referee is correct that the Gaia photometry is significantly more accurate than PanSTARRS, however, at faint magnitudes, the regime where GD-1 members are most easily separable from the Milky Way field stars, the Gaia BP-RP color is much more uncertain than the PanSTARRS g-i color.
-As a part of the response, we attach a figure that compares Gaia (left) and PanSTARRS (right) color-magnitude diagrams of GD-1 stars, selected with -1<phi_2<1, -50<phi_1<-10, and our fiducial proper motion selection.
+The referee is correct that the Gaia photometry is more accurate than PanSTARRS, however, at faint magnitudes, the regime where GD-1 members are most easily separable from the Milky Way field stars, the Gaia BP-RP color is much more uncertain than the PanSTARRS g-i color.
+As a part of the response, we attach a figure that compares Gaia (left) and PanSTARRS (right) color-magnitude diagrams of GD-1 stars, selected within -1<phi_2<1, -50<phi_1<-10, and our fiducial proper motion selection.
 Typical photometric uncertainties are shown as a function of magnitude, centered on 0 color in both panels.
@@
 This leads to GD-1's main sequence being much better defined in PanSTARRS color-magnitude diagram than in Gaia's, so we opt to retain PanSTARRS photometry for selection of likely GD-1 members.
-
-Q: do we comment on using g vs G?
 
@@
 
-* AB: Punt to next paper: respond that distance not the focus
-- we do lay an isochrone in fig 1, consistent with a previously reported distance and stellar population
-- however, we note that fig 4 indicates a distance gradient, closer at trailing than leading tail, also consistent with the literature
-- while re-deriving a distance to the stream using would be useful, it is rather involved and would need to be done along the stream, so we are leaving it for our next contribution where we will derive more general properties of GD-1, while focusing on the new features in this one
+An isochrone at a fiducial distance to the stream is already overlaid on the GD-1 color-magnitude diagram (cmd) in Figure 1, and used to define a cmd selection box, which is very useful for improving the selection of likely GD-1 members.
+Since the focus of this paper is reporting on new features of GD-1 detected in a cleaner map of the stream, using literature distance to the stream as a means to improve member selection seems adequate.
+Of course, more precise distances, and hence tangential velocities, would be especially useful for dynamical modeling of GD-1, but improving on the current measurements would require a more involved calculation, so we leave it for future work where we plan to comprehensively derive stream properties in all six dimensions of the phase space.
 

### 2018-06-13 paper/submit-aas/referee-response.txt
response: spur-gap association

@@
 
-* AB: explain the reference to Yoon et al., be explicit about the possible
-  connection to the gap
+To clarify the possible association, we've changed this statement to:
+---
+Both the spur and the blob are close to large gaps in the stream, so a mechanism that displaces stars from their original orbits along the stream to orbits beyond the stream would explain both sets of features.
+Future observations of the full 3-D kinematics combined with detailed dynamical modeling of the GD-1 system will be able to test this scenario.
+---
 

### 2018-06-13 paper/gd1.tex
updated figure
response: spur-gap association

@@
 \newcommand{\todo}[1]{{\color{red} TODO: #1}}
+\newcommand{\changes}[1]{{\color{green} #1}}
 
@@
        Gaia reveals GD-1 stars outside of the main stream}
-% Gaia DR2 reveals stream stars...
+% Gaia DR2 reveals stream dmlocal/tex/stars...
 % A first look at the GD-1 stellar stream with Gaia DR2
@@
 (\citealt{Schonrich:2010, Bovy:2015}).
+\changes{In detail, we project the solar velocity vector onto the tangent space at each star's sky position, convert to angular motion using the assumed distance, and subtract the solar component from each star's proper motion (this is done using the \package{astropy} and \package{gala} packages).}
 \figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <
@@
 
-Stars from the spur and the blob are similar to those along the main stream track in the color-magnitude diagram and in proper motions, which confirms their association to GD-1.
-The proximity of these off-stream features to large gaps in the stream is suggestive of the gaps being their origin, further implied by hints of relative velocity differences.
-For example, the median proper motion of the blob is offset from the adjacent stream field by $\approx0.3$\,mas\,yr$^{-1}$ in the positive $\mu_{\phi_1}$ direction.
-We defer detailed modeling of these features to infer their co-evolution with the stream for future work.
+Stars from the spur and the blob are similar to those along the main stream track in the color-magnitude diagram and in proper motions, which confirms their likely association to GD-1.
+Both the spur and the blob are close to large gaps in the stream, so a mechanism that displaces stars from their original orbits along the stream to orbits beyond the stream would explain both sets of features.
+Future observations of the full 3-D kinematics combined with detailed dynamical modeling of the GD-1 system will be able to test this scenario.
+
+% The proximity of these off-stream features to large gaps in the stream is suggestive of the gaps being their origin, further implied by hints of relative velocity differences.
+% For example, the median proper motion of the blob is offset from the adjacent stream field by $\approx0.3$\,mas\,yr$^{-1}$ in the positive $\mu_{\phi_1}$ direction.
+% We defer detailed modeling of these features to infer their co-evolution with the stream for future work.
 

### 2018-06-14 paper/submit-aas/referee-response.txt
new figure and response

@@
 
-* APW:
-* TODO: final figure summarize data quality and dust
-* Push back on Zoom-ins? -> plot them for the referee, don't add much
+We thank the referee for these suggestions and agree that it is useful and necessary to show the quality of the astrometric solution in this region of the sky.
+We have added a new Figure to the Appendix that contains maps of (1) the density of 2-parameters solutions in this region, (2) the unit weight error (as defined in Lindegren+2018, C.1), and (3) the SFD dust extinction around GD-1.
 
-- Dust map done
-- 2-parameter solution done
-- Source density at phi1 = [-40, -20, 0], phi2 = 0 is ~8000 sources/deg^2 -- should be close to 100% complete according to Arenou.
-https://arxiv.org/pdf/1804.09375.pdf
-https://arxiv.org/pdf/1804.09366.pdf
+We have also computed the source density at several locations along the stream to assess whether source crowding could impact the completeness (as in Arenou+2018).
+We find that, at each location along the stream between phi1 = (-50, -10) the source density is never larger than ~8000 sources/deg^2 (estimated using the 2-parameter and 5-parameter sources), and is far below the densities where crowding becomes a problem.
+
+TODO: correlations between proper motion components?
+
+We have decided not to include further zoom-in panels, as our figures already contain many panels.
 
@@
 
-We have added a sentence to clarify how this was done.
+We have added a sentence to clarify how this was done:
+---
+In detail, we project the solar velocity vector onto the tangent space at each star's sky position, convert to angular motion using the assumed distance, and subtract the solar component from each star's proper motion (this is done using the \package{astropy} and \package{gala} packages).
+---
 
@@
 
-* APW: simple statistical technique in 4D, punt
+The parabolic model (in phi1, phi2) is fit to the stream ridge points, then shifted up and down to define a selection region (as described in Section 3.1).
+
+We agree that, to characterize the stream, it would be better to build a probabilistic model of the stream in all available data dimensions.
+However, to be useful for future work, this would require simultaneously modeling the phase-space density of the stream so that the variations with distance and distribution of proper motions could be taken into account in a robust way.
+As the focus of this work is mainly to point out unexpected, off-track features that appear related to the stream, we defer this more sophisticated modeling to future work.

### 2018-06-15 notebooks/orbitfit.py
orbit fit script

@@ -1,0 +1,227 @@
+# coding: utf-8
+
+""" Utilities for fitting orbits to stream data. """
+
+# Third-party
+import astropy.coordinates as coord
+import astropy.units as u
+import numpy as np
+from scipy.interpolate import InterpolatedUnivariateSpline
+from scipy.stats import norm
+from gala.units import galactic
+
+# Project
+from . import galactocentric_frame, vcirc, vlsr
+from .coordinates import Ophiuchus
+from .util import integrate_forward_backward
+
+__all__ = ['ln_prior', 'ln_likelihood', 'ln_posterior']
+
+# ----------------------------------------------------------------------------
+# For inference:
+
+def _unpack(p, freeze=None):
+    if freeze is None:
+        freeze = dict()
+
+    # these are for the initial conditions
+    phi2, d, mu1, mu2, vr = p[:5]
+    count_ix = 5
+
+    # time to integrate forward and backward
+    if 't_forw' not in freeze:
+        t_forw = p[count_ix]
+        count_ix += 1
+    else:
+        t_forw = freeze['t_forw']
+
+    if 't_back' not in freeze:
+        t_back = p[count_ix]
+        count_ix += 1
+    else:
+        t_back = freeze['t_back']
+
+    # prior on instrinsic width of stream
+    if 'phi2_sigma' not in freeze:
+        phi2_sigma = p[count_ix]
+        count_ix += 1
+    else:
+        phi2_sigma = freeze['phi2_sigma']
+
+    # prior on instrinsic depth (distance) of stream
+    if 'd_sigma' not in freeze:
+        d_sigma = p[count_ix]
+        count_ix += 1
+    else:
+        d_sigma = freeze['d_sigma']
+
+    # prior on instrinsic LOS velocity dispersion of stream
+    if 'vr_sigma' not in freeze:
+        vr_sigma = p[count_ix]
+        count_ix += 1
+    else:
+        vr_sigma = freeze['vr_sigma']
+
+    return phi2, d, mu1, mu2, vr, t_forw, t_back, phi2_sigma, d_sigma, vr_sigma
+
+def ln_prior(p, ophdata, potential, dt, freeze=None):
+    """
+    Evaluate the prior over stream orbit fit parameters.
+
+    See docstring for `ln_likelihood()` for information on args and kwargs.
+
+    """
+
[…]

### 2018-06-19 paper/submit-aas/referee-response.txt
update response and paper text

@@
-We thank the referee for a constructive report that has improved the quality of
-this work. Below, we have copied the remarks from the referee (indented by >)
-and have responded to each point in text that immediately follows.
+We thank the referee for a constructive report that has improved the quality of this work.
+Below, we have copied the remarks from the referee (indented by >) and have responded to each point in text that immediately follows.
 
@@
 
-We thank the referee for these suggestions and agree that it is useful and necessary to show the quality of the astrometric solution in this region of the sky.
-We have added a new Figure to the Appendix that contains maps of (1) the density of 2-parameters solutions in this region, (2) the unit weight error (as defined in Lindegren+2018, C.1), and (3) the SFD dust extinction around GD-1.
-
+We thank the referee for these suggestions and agree that it is useful and important to discuss the quality of the astrometric solution in this region of the sky.
+We have attached a figure to this report that shows maps of (1) the density of 2-parameters solutions in this region, (2) the unit weight error (as defined in Lindegren+2018, C.1), and (3) the SFD dust extinction around GD-1.
 We have also computed the source density at several locations along the stream to assess whether source crowding could impact the completeness (as in Arenou+2018).
@@
 
-TODO: correlations between proper motion components?
+We don't see any correlations between astrometric issues or incompleteness and the features that we see associated with the stream.
+We do, however, see density variations in the main stream track due to the scanning pattern (especially between phi1=[-80, -55]).
+Given this and the above, we have added some text to Section TODO to explain this, but have decided not to add the attached figure to the text because we have instead included a new figure to address a later point (see below).
 
-We have decided not to include further zoom-in panels, as our figures already contain many panels.
+We have also looked at the magnitude of correlations between proper motion components in the selection region and found TODO.
+
+Lastly, we have decided not to include further zoom-in panels, as our figures already contain many panels, but we have tightened the axis ranges on the proper motion panels of (what is now) Figure TODO and hope that this makes the distribution in proper-motion space clearer.
 
@@
 
-The referee is correct that the Gaia photometry is more accurate than PanSTARRS, however, at faint magnitudes, the regime where GD-1 members are most easily separable from the Milky Way field stars, the Gaia BP-RP color is much more uncertain than the PanSTARRS g-i color.
+The referee is correct that the Gaia photometry is more accurate than PanSTARRS, however, at faint magnitudes (G > 18) --- the regime where GD-1 members are most easily separable from the Milky Way field stars --- the Gaia BP-RP color is much more uncertain than the PanSTARRS g-i color.
 As a part of the response, we attach a figure that compares Gaia (left) and PanSTARRS (right) color-magnitude diagrams of GD-1 stars, selected within -1<phi_2<1, -50<phi_1<-10, and our fiducial proper motion selection.
 Typical photometric uncertainties are shown as a function of magnitude, centered on 0 color in both panels.
-PanSTARRS photometry is systematics dominated in the entire range, and the uncertainty in both g and i bands is 0.01 mag (0.014 mag for g-i color).
+PanSTARRS photometry is systematics-dominated in the entire range, and the uncertainty in both g and i bands is 0.01 mag (0.014 mag for g-i color).
 Uncertainties in Gaia's G band are smaller in the entire range, but BP-RP's vary from 0.001 at G~15 to 0.05 at G~20, approximately 5 times larger than PanSTARRS at comparably faint magnitudes.
@@
 The orbit shown in Figure 3 crosses the disk midplane (z=0) twice in the segment of the orbit shown.
-Given uncertainties in the Milky Way's gravitational potential, and uncertainties in the full-space motion of the stream progenitor, we don't think that including a 6-8 Gyr integration in the paper will be too meaningful.
+Given uncertainties in the Milky Way's gravitational potential, and uncertainties in the full-space motion of the stream progenitor, we don't think that including a 6-8 Gyr integration in the paper will be more illustrative.
 However, we have extended the integration time for the paper figure to 1 Gyr, and attach a figure with a 6 Gyr integration with this response for the referee to see.
@@
 As the focus of this work is mainly to point out unexpected, off-track features that appear related to the stream, we defer this more sophisticated modeling to future work.
+
+=========================================
+For the journal editor:
+
+We have addressed each of the referee's comments in the attached response, but in doing so we have added a new figure that we believe improves the paper and directly addresses one of the referee's points.
+We realize that now the word count is slightly over the limit for ApJ letters (we count XX words).
+However, we do still feel that this result and paper is in the spirit of the Letters and would like to keep it in consideration for this journal.
+Please let us know if we should work on trimming the text, or else please advise.
+Thanks!

### 2018-06-19 paper/gd1.tex
update response and paper text

@@
 \newcommand{\article}{\textsl{Letter}}
+\newcommand{\given}{\,|\,}
 
@@
 \newcommand{\todo}[1]{{\color{red} TODO: #1}}
-\newcommand{\changes}[1]{{\color{green} #1}}
+\newcommand{\changes}[1]{{\textbf{#1}}}
 
@@
 
+\changes{
+To quantify the significance of these features, we construct a statistical model of the linear density of the filtered sky positions as a function of $\phi_2$ around each of the features and at a control region.
+We select three equal ranges of $\phi_1$ centered on the spur, $\phi_1 \in ()^\circ$, the blob, $\phi_1 \in ()^\circ$, and in between, as a control field, $\phi_1 \in ()^\circ$, and model the one-dimensional density in $\phi_2$ in each of these regions to see if the number of stars at the location of each feature is consistent with the background level.
+We use a four-component mixture model to represent the density consisting of a uniform background over the range $\phi_2 \in (-10, 5)^\circ$, a two-component Gaussian mixture to represent the main stream with mean $\mu_{\textrm{s}}$ and variances $\sigma_{\textrm{s}, 1}^2$ and $\sigma_{\textrm{s}, 2}^2$, and a single Gaussian for each feature with mean and variance $(\mu_{\textrm{f}}, \sigma_{\textrm{f}}^2)$.
+The full density model given all parameters $\bs{\theta} = (\alpha_{\textrm{bg}}, \alpha_{\textrm{s}}, \mu_{\textrm{s}}, \sigma_{\textrm{s}, 1}, \sigma_{\textrm{s}, 2}, \mu_{\textrm{f}}, \sigma_{\textrm{f}})$ is
+\begin{equation}
+    \begin{split}
+    p(\phi_2 \given \bs{\theta}) &=
+            \alpha_{\textrm{bg}} \, \mathcal{U}(-10, 5) \\ & \quad +
+            \alpha_{\textrm{s}} \,
+                (\mathcal{N}(\phi_2 \given \mu_{\textrm{s}}, \sigma_{\textrm{s}, 1}) + \mathcal{N}(\phi_2 \given \mu_{\textrm{s}}, \sigma_{\textrm{s}, 2})) \\ & \quad +
+            \alpha_{\textrm{f}} \,
+                \mathcal{N}(\phi_2 \given \mu_{\textrm{f}}, \sigma_{\textrm{f}})
+    \end{split}
+\end{equation}
+where $\alpha_{\textrm{bg}}$ is the fraction of background stars in the field, $\alpha_{\textrm{s}}$ is the fraction of stars associated with the mean stream track, and $\alpha_{\textrm{f}} = (1 - \alpha_{\textrm{bg}} - \alpha_{\textrm{s}})$ is the fraction of stars associated with either the spur, the blob, or at the location of the spur/blob in the control field.
+% and ignore uncertainties on the sky position $\phi_2$.
+
+We generate posterior samples in the parameters of this model in each $\phi_1$ range by specifying uniform priors over ...
+
+}
+
 Stars from the spur and the blob are similar to those along the main stream track in the color-magnitude diagram and in proper motions, which confirms their likely association to GD-1.
-Both the spur and the blob are close to large gaps in the stream, so a mechanism that displaces stars from their original orbits along the stream to orbits beyond the stream would explain both sets of features.
-Future observations of the full 3-D kinematics combined with detailed dynamical modeling of the GD-1 system will be able to test this scenario.
+\changes{Both the spur and the blob are close to large gaps in the stream, so a mechanism that displaces stars from their original orbits along the stream to orbits beyond the stream would explain both sets of features.
+Future observations of the full 3-D kinematics combined with detailed dynamical modeling of the GD-1 system will be able to test this scenario.}
 

### 2018-06-19 paper/gd1.tex
describe inference

@@
 To quantify the significance of these features, we construct a statistical model of the linear density of the filtered sky positions as a function of $\phi_2$ around each of the features and at a control region.
-We select three equal ranges of $\phi_1$ centered on the spur, $\phi_1 \in ()^\circ$, the blob, $\phi_1 \in ()^\circ$, and in between, as a control field, $\phi_1 \in ()^\circ$, and model the one-dimensional density in $\phi_2$ in each of these regions to see if the number of stars at the location of each feature is consistent with the background level.
+We select three equal ranges of $\phi_1$ centered on the spur, $\phi_1 \in ()^\circ$, the blob, $\phi_1 \in ()^\circ$, and in between (as a control field), $\phi_1 \in ()^\circ$, and model the one-dimensional density in $\phi_2$ in each of these regions.
+% to see if the number of stars at the location of each feature is consistent with the background level.
 We use a four-component mixture model to represent the density consisting of a uniform background over the range $\phi_2 \in (-10, 5)^\circ$, a two-component Gaussian mixture to represent the main stream with mean $\mu_{\textrm{s}}$ and variances $\sigma_{\textrm{s}, 1}^2$ and $\sigma_{\textrm{s}, 2}^2$, and a single Gaussian for each feature with mean and variance $(\mu_{\textrm{f}}, \sigma_{\textrm{f}}^2)$.
@@
 % and ignore uncertainties on the sky position $\phi_2$.
-
-We generate posterior samples in the parameters of this model in each $\phi_1$ range by specifying uniform priors over ...
-
+}
+\changes{
+We generate posterior samples in the parameters of this model in each $\phi_1$ range by specifying uniform priors over all parameters except the variances, for which we use priors that are uniform in the logarithm of the variance.
+We use an ensemble Markov Chain Monte Carlo (MCMC) sampler with 64 walkers to generate posterior samples (\citealt{Foreman-Mackey:2013}) and run the sampler for a total of 4096 steps, discarding the first 2048 samples as ``burn-in.''
+\figurename~\ref{fig:density-model} summarizes the results of this inference: top row shows histograms of the $\phi_2$ positions of stars in each $\phi_1$ range (left to right) and bottom row shows marginal posterior probability distributions for $\alpha_{\textrm{bg}}$, $\alpha_{\textrm{s}}$, and $\alpha_{\textrm{f}}$ estimated from the posterior samples for each $phi_1$ range.
+TODO: shows what
 }

### 2018-06-19 paper/gd1.tex
move fig and implications

@@
 We generate posterior samples in the parameters of this model in each $\phi_1$ range by specifying uniform priors over all parameters except the variances, for which we use priors that are uniform in the logarithm of the variance.
-We use an ensemble Markov Chain Monte Carlo (MCMC) sampler with 64 walkers to generate posterior samples (\citealt{Foreman-Mackey:2013}) and run the sampler for a total of 4096 steps, discarding the first 2048 samples as ``burn-in.''
-\figurename~\ref{fig:density-model} summarizes the results of this inference: top row shows histograms of the $\phi_2$ positions of stars in each $\phi_1$ range (left to right) and bottom row shows marginal posterior probability distributions for $\alpha_{\textrm{bg}}$, $\alpha_{\textrm{s}}$, and $\alpha_{\textrm{f}}$ estimated from the posterior samples for each $phi_1$ range.
-TODO: shows what
+We use an ensemble Markov Chain Monte Carlo (MCMC) sampler (\citealt{Foreman-Mackey:2013}) to generate posterior samples and run the sampler with 64 walkers for a total of 4096 steps, discarding the first 2048 samples as ``burn-in.''
+\figurename~\ref{fig:density-model} summarizes the results of this inference: top row shows histograms of the $\phi_2$ positions of stars in each $\phi_1$ range (left to right) with maximum a posteriori inferred density over-plotted (black curve) and bottom row shows marginal posterior probability distributions for $\alpha_{\textrm{bg}}$, $\alpha_{\textrm{s}}$, and $\alpha_{\textrm{f}}$ estimated from the posterior samples for each $phi_1$ range.
+In each of the feature fields, the posterior samples for the feature mixture component amplitude are inconsistent with 0, implying that the features are not consistent with the background stellar density.
 }
+
+% \begin{figure}
+% \begin{center}
+% \includegraphics[width=\textwidth]{density-model.pdf}
+% \end{center}
+% \caption{
+% TODO
+% }
+% \label{fig:density-model}
+% \end{figure}
 

### 2018-06-19 paper/submit-aas/referee-response.txt
more words

@@
 We do, however, see density variations in the main stream track due to the scanning pattern (especially between phi1=[-80, -55]).
-Given this and the above, we have added some text to Section TODO to explain this, but have decided not to add the attached figure to the text because we have instead included a new figure to address a later point (see below).
+Given this and the above, we have added some text to Section 3.1 to describe this, but have decided not to add the attached figure to the text because we have instead included a new figure to address a later point (see below).
 
-We have also looked at the magnitude of correlations between proper motion components in the selection region and found TODO.
+We have also looked at the magnitude of correlations between proper motion components in the selection region and find that TODO.
 
-Lastly, we have decided not to include further zoom-in panels, as our figures already contain many panels, but we have tightened the axis ranges on the proper motion panels of (what is now) Figure TODO and hope that this makes the distribution in proper-motion space clearer.
+Lastly, we have decided not to include further zoom-in panels, as our figures already contain many panels, but we have tightened the axis ranges on the proper motion panels of (what is now) Figure 5 and hope that this makes the distribution in proper-motion space clearer.
 
@@
 
-* APW: easy, but stupid
+TODO:
 
@@
 
-* APW/AB: Define significance
+We have added a new figure and some discussion to Section 3.1 to address this.
+In brief, we construct a model of the phi_2 distribution in three areas of phi_1: one centered on the spur, one on the blob, and one in between as a control field.
+We find that in the feature fields, the model prefers adding an additional component to explain the spur and the blob (rather than just background variations), whereas in the control field the amplitude of the additional component is consistent with 0.
 

### 2018-06-19 paper/gd1.tex
more words

@@
 
-Clear density variations, also apparent in the 2D positions of the sources,
-manifest as sharp features in the surface density estimates along the stream.
-We have verified that these do not map to features in dust extinction nor are
-they caused by completeness variations in the \gaia\ scanning pattern:
-the visible portions of the GD-1 stream are located at high Galactic latitudes
-($b > 20^\circ$), and 99\% of our sample has at least 8 \gaia\ visibility
-periods, promising a robust astrometric solution.
+Clear density variations, also apparent in the 2D positions of the sources, manifest as sharp features in the surface density estimates along the stream.
+We have verified that these do not correspond to features in dust extinction nor are they caused by completeness variations in the \gaia\ scanning pattern:
+the visible portions of the GD-1 stream are located at high Galactic latitudes ($b > 20^\circ$), and 99\% of our sample has at least 8 \gaia\ visibility periods \changes{and satisfy a selection on the unit weight error (see Equation~C.1 in \citealt{Lindegren:2018})}, promising a robust astrometric solution.
+\changes{
+Small-scale density variations, however, especially apparent between $-80^\circ < \phi_1 < -50^\circ$, are a consequence of the \gaia\ scanning pattern.
+}
+
 % First, the visible portions of the GD-1 stream are located at high Galactic
@@
 
-\figurename~\ref{fig:orbitfit} shows the best-fitting orbit from the procedure
-outlined above.
+\figurename~\ref{fig:orbitfit} shows \changes{a $1 \textrm{Gyr}$ segment of} the best-fitting orbit from the procedure outlined above.
 The GD-1 stream crosses the midplane at large radius, far from the bulk of the

### 2018-06-19 paper/gd1.tex
polynomial order

@@
 To study the global properties of the stream as a function of stream longitude,
-we extract stream stars around a second-order polynomial model for the stream
+we extract stream stars around a fourth-order polynomial model for the stream
 latitude, $\phi_2$, as a function of stream longitude, $\phi_1$.
@@
 \textit{Panel (a)}: Sky positions of candidate GD-1 stars.
-Lines show a 2nd-order polynomial fit to the median stream track offset by $\pm
-0.75^\circ$; the enclosed region is adopted as the nominal stream region.
+Lines show a \changes{4th}-order polynomial fit to the median stream track offset by $\pm 0.75^\circ$; the enclosed region is adopted as the nominal stream region.
 \textit{Panel (b)}: Sky positions of simulated star particles from a mock stream

### 2018-06-20 paper/gd1.tex
new figure

@@
 To quantify the significance of these features, we construct a statistical model of the linear density of the filtered sky positions as a function of $\phi_2$ around each of the features and at a control region.
-We select three equal ranges of $\phi_1$ centered on the spur, $\phi_1 \in ()^\circ$, the blob, $\phi_1 \in ()^\circ$, and in between (as a control field), $\phi_1 \in ()^\circ$, and model the one-dimensional density in $\phi_2$ in each of these regions.
+We select three equal ranges of $\phi_1$ centered on the spur, $\phi_1 \in (-36, -30)^\circ$, the blob, $\phi_1 \in (-18, -12)^\circ$, and in between (as a control field), $\phi_1 \in (-51, -45)^\circ$, and model the one-dimensional density in $\phi_2$ in each of these regions.
 % to see if the number of stars at the location of each feature is consistent with the background level.
@@
             \alpha_{\textrm{bg}} \, \mathcal{U}(-10, 5) \\ & \quad +
-            \alpha_{\textrm{s}} \,
-                (\mathcal{N}(\phi_2 \given \mu_{\textrm{s}}, \sigma_{\textrm{s}, 1}) + \mathcal{N}(\phi_2 \given \mu_{\textrm{s}}, \sigma_{\textrm{s}, 2})) \\ & \quad +
+            \alpha_{\textrm{s}, 1} \, \mathcal{N}(\phi_2 \given \mu_{\textrm{s}}, \sigma_{\textrm{s}, 1}) +
+            \alpha_{\textrm{s}, 2} \, \mathcal{N}(\phi_2 \given \mu_{\textrm{s}}, \sigma_{\textrm{s}, 2}) \\ & \quad +
             \alpha_{\textrm{f}} \,
@@
 \end{equation}
-where $\alpha_{\textrm{bg}}$ is the fraction of background stars in the field, $\alpha_{\textrm{s}}$ is the fraction of stars associated with the mean stream track, and $\alpha_{\textrm{f}} = (1 - \alpha_{\textrm{bg}} - \alpha_{\textrm{s}})$ is the fraction of stars associated with either the spur, the blob, or at the location of the spur/blob in the control field.
+where $\alpha_{\textrm{bg}}$ is the fraction of background stars in the field, $\alpha_{\textrm{s}} = \alpha_{\textrm{s}, 1} + \alpha_{\textrm{s}, 2}$ is the fraction of stars associated with the mean stream track, and $\alpha_{\textrm{f}} = (1 - \alpha_{\textrm{bg}} - \alpha_{\textrm{s}})$ is the fraction of stars associated with either the spur, the blob, or at the location of the spur/blob in the control field.
 % and ignore uncertainties on the sky position $\phi_2$.
@@
 We use an ensemble Markov Chain Monte Carlo (MCMC) sampler (\citealt{Foreman-Mackey:2013}) to generate posterior samples and run the sampler with 64 walkers for a total of 4096 steps, discarding the first 2048 samples as ``burn-in.''
-\figurename~\ref{fig:density-model} summarizes the results of this inference: top row shows histograms of the $\phi_2$ positions of stars in each $\phi_1$ range (left to right) with maximum a posteriori inferred density over-plotted (black curve) and bottom row shows marginal posterior probability distributions for $\alpha_{\textrm{bg}}$, $\alpha_{\textrm{s}}$, and $\alpha_{\textrm{f}}$ estimated from the posterior samples for each $phi_1$ range.
+\figurename~\ref{fig:density-model} summarizes the results of this inference: top row shows histograms of the $\phi_2$ positions of stars in each $\phi_1$ range (left to right) with maximum a posteriori inferred density over-plotted (black curve) and bottom row shows marginal posterior probability distributions for $\alpha_{\textrm{bg}}$, $\alpha_{\textrm{s}}$, and $\alpha_{\textrm{f}}$ estimated from the posterior samples for each $\phi_1$ range.
 In each of the feature fields, the posterior samples for the feature mixture component amplitude are inconsistent with 0, implying that the features are not consistent with the background stellar density.
@@
 
-% \begin{figure}
-% \begin{center}
-% \includegraphics[width=\textwidth]{density-model.pdf}
-% \end{center}
-% \caption{
-% TODO
-% }
-% \label{fig:density-model}
-% \end{figure}
+\begin{figure}
+\begin{center}
+\includegraphics[width=\textwidth]{density-model.pdf}
+\end{center}
+\caption{
+TODO
+}
+\label{fig:density-model}
+\end{figure}
 

### 2018-06-20 paper/gd1.tex
todo
add pm measurement and update density fig

@@
 fully-disrupted progenitor.
+\changes{At this location, between $\phi_1 \in (-18, -10)^\circ$, we find that the mean equatorial sky position of the stream stars is $(\alpha, \delta) = (177.01, 53.99)^\circ$ and the mean proper motion is $(\mu_\alpha \, \cos\delta, \mu_\delta) = (-7.78, -7.85) \pm 0.03~\masyr$, taking into account the covariance matrix for proper motions provided with the \gaia\ data.}
+
 Several of the under-densities and gaps hinted at from photometric selection
@@
 
-We extract properties of the stream from the region defined above as a function
-of $\phi_1$ by computing median stream properties in (overlapping) $3^\circ$ windows.
+We extract properties of the stream from the region defined above as a function of $\phi_1$ by computing median stream properties in (overlapping) $3^\circ$ windows.
 % we set the window size to $3^\circ$ in $\phi_1$, and shift the window center by
@@
 stream width to future work.
+\changes{With present data, however, we cannot rule out a scenario in which the progenitor system disrupted longer ago and resulted in one of the larger gaps (e.g., \citealt{DeBoer:2018}).}
 % that has not yet phase-mixed in multiple dimensions.
@@
 \caption{
-TODO
+\emph{Top row}: Density of stars in $\phi_2$ (grey) and inferred density (black curve) for regions around the spur, blob, and a control field.
+The model for the density includes components for the stream, background, and a feature overdensity (the spur or the blob), but with amplitudes that are allowed to go to zero (see \sectionname~\ref{sec:res_gap} for more details.)
+\emph{Bottom row}: Inferred posterior probability distributions over the amplitude ($\alpha$) of each component in each field: $\alpha_{\textrm{s}}$ for the stream, $\alpha_{\textrm{bg}}$ for the background, and $\alpha_{\textrm{f}}$ for the feature.
+In both feature fields, the model prefers an additional component to describe the feature, whereas in the control field the density is consistent with stream + background.
 }

### 2018-06-20 paper/submit-aas/referee-response.txt
todo
final touches

@@
 
-We have also looked at the magnitude of correlations between proper motion components in the selection region and find that TODO.
-
 Lastly, we have decided not to include further zoom-in panels, as our figures already contain many panels, but we have tightened the axis ranges on the proper motion panels of (what is now) Figure 5 and hope that this makes the distribution in proper-motion space clearer.
@@
 
-TODO:
+We have added this quantity to the paper in Section 2:
+---
+At this location, between $\phi_1 \in (-18, -10)^\circ$, we find that the mean equatorial sky position of the stream stars is $(\alpha, \delta) = (177.01, 53.99)^\circ$ and the mean proper motion is $(\mu_\alpha \, \cos\delta, \mu_\delta) = (-7.78, -7.85) \pm 0.03~\masyr$, taking into account the covariance matrix for proper motions provided with the \gaia\ data.
+---
 

### 2018-06-20 paper/submit-aas/referee-response.txt
word

@@
 We do, however, see density variations in the main stream track due to the scanning pattern (especially between phi1=[-80, -55]).
-Given this and the above, we have added some text to Section 3.1 to describe this, but have decided not to add the attached figure to the text because we have instead included a new figure to address a later point (see below).
+Given this and the above, we have added some text to Section 3.1 to note this, but have decided not to add the attached figure to the text because we have instead included a new figure to address a later point (see below).
 
-Lastly, we have decided not to include further zoom-in panels, as our figures already contain many panels, but we have tightened the axis ranges on the proper motion panels of (what is now) Figure 5 and hope that this makes the distribution in proper-motion space clearer.
+Lastly, we have decided not to include further zoom-in panels, as our figures already contain many panels, but we have tightened the axis ranges on the proper motion panels of (what is now) Figure 5.
+We hope that this makes the distribution in proper-motion space clearer.
 

### 2018-06-20 paper/gd1.tex
software citations

@@
 (\citealt{Schonrich:2010, Bovy:2015}).
-\changes{In detail, we project the solar velocity vector onto the tangent space at each star's sky position, convert to angular motion using the assumed distance, and subtract the solar component from each star's proper motion (this is done using the \package{astropy} and \package{gala} packages).}
+\changes{In detail, we project the solar velocity vector onto the tangent space at each star's sky position, convert to angular motion using the assumed distance, and subtract the solar component from each star's proper motion (this is done using the \package{astropy} \citep{astropy} and \package{gala} \citep{gala} packages).}
 \figurename~\ref{fig:selection} (top right) shows stars with $|\phi_2| <

### 2018-06-20 paper/gd1.tex
typos

--- 2018-06-20 paper/gd1.tex
@@
 
-\figurename~\ref{fig:orbitfit} shows \changes{a $1 \textrm{Gyr}$ segment of} the best-fitting orbit from the procedure outlined above.
+\figurename~\ref{fig:orbitfit} shows \changes{a $1\,\textrm{Gyr}$ segment of} the best-fitting orbit from the procedure outlined above.
 The GD-1 stream crosses the midplane at large radius, far from the bulk of the
@@
 \caption{
-\emph{Top row}: Density of stars in $\phi_2$ (grey) and inferred density (black curve) for regions around the spur, blob, and a control field.
+{\bf \emph{Top row}: Density of stars in $\phi_2$ (grey) and inferred density (black curve) for regions around the spur, blob, and a control field.
 The model for the density includes components for the stream, background, and a feature overdensity (the spur or the blob), but with amplitudes that are allowed to go to zero (see \sectionname~\ref{sec:res_gap} for more details.)
 \emph{Bottom row}: Inferred posterior probability distributions over the amplitude ($\alpha$) of each component in each field: $\alpha_{\textrm{s}}$ for the stream, $\alpha_{\textrm{bg}}$ for the background, and $\alpha_{\textrm{f}}$ for the feature.
-In both feature fields, the model prefers an additional component to describe the feature, whereas in the control field the density is consistent with stream + background.
+In both feature fields, the model prefers an additional component to describe the feature, whereas in the control field the density is consistent with stream + background.}
 }

### 2018-06-21 paper/gd1.tex
add sentence about percentiles

@@
 In each of the feature fields, the posterior samples for the feature mixture component amplitude are inconsistent with 0, implying that the features are not consistent with the background stellar density.
+In each of the $\phi_1$ regions, the 15th and 85th percentile values of the feature amplitude, $\alpha_{\textrm{f}}$, are $(0.12, 0.19)$ (spur), $(0.07, 0.22)$ (blob), and $(0.005, 0.103)$ (control).
 }

### 2018-06-21 paper/gd1.tex

@@
 \caption{
-{\bf \emph{Top row}: Density of stars in $\phi_2$ (grey) and inferred density (black curve) for regions around the spur, blob, and a control field.
+\changes{\emph{Top row}: Density of stars in $\phi_2$ (grey) and inferred density (black curve) for regions around the spur, blob, and a control field.
 The model for the density includes components for the stream, background, and a feature overdensity (the spur or the blob), but with amplitudes that are allowed to go to zero (see \sectionname~\ref{sec:res_gap} for more details.)

### 2018-06-21 paper/gd1.tex
reword end of significance

@@
 \figurename~\ref{fig:density-model} summarizes the results of this inference: top row shows histograms of the $\phi_2$ positions of stars in each $\phi_1$ range (left to right) with maximum a posteriori inferred density over-plotted (black curve) and bottom row shows marginal posterior probability distributions for $\alpha_{\textrm{bg}}$, $\alpha_{\textrm{s}}$, and $\alpha_{\textrm{f}}$ estimated from the posterior samples for each $\phi_1$ range.
-In each of the feature fields, the posterior samples for the feature mixture component amplitude are inconsistent with 0, implying that the features are not consistent with the background stellar density.
-In each of the $\phi_1$ regions, the 15th and 85th percentile values of the feature amplitude, $\alpha_{\textrm{f}}$, are $(0.12, 0.19)$ (spur), $(0.07, 0.22)$ (blob), and $(0.005, 0.103)$ (control).
+The 15th and 85th percentile values of the feature amplitude in different fields are $(0.12, 0.19)$ (spur), $(0.07, 0.22)$ (blob), and $(0.005, 0.103)$ (control).
+The posterior samples for the amplitude of the feature mixture component are consistent with 0 in the control field, marginally inconsistent with 0 in the blob field, and significantly inconsistent with 0 in the spur field.
+% In each of the feature fields, the posterior samples for the feature mixture component amplitude are inconsistent with 0, implying that the features are not consistent with the background stellar density.
+% In each of the $\phi_1$ regions, the 15th and 85th percentile values of the feature amplitude, $\alpha_{\textrm{f}}$, are $(0.12, 0.19)$ (spur), $(0.07, 0.22)$ (blob), and $(0.005, 0.103)$ (control).
 }

### 2018-06-21 paper/submit-aas/referee-response.txt
response typo
minor change

@@
 
-The parabolic model (in phi1, phi2) is fit to the stream ridge points, then shifted up and down to define a selection region (as described in Section 3.1).
+The polyomial model (in phi1, phi2) is fit to the stream ridge points, then shifted up and down to define a selection region (as described in Section 3.1).
 
@@
 We realize that now the word count is slightly over the limit for ApJ letters (we count XX words).
-However, we do still feel that this result and paper is in the spirit of the Letters and would like to keep it in consideration for this journal.
+However, we do still feel that this result and paper are in the spirit of the Letters and would like to keep it in consideration for this journal.
 Please let us know if we should work on trimming the text, or else please advise.

### 2018-06-21 paper/gd1.tex
minor change

@@
 The 15th and 85th percentile values of the feature amplitude in different fields are $(0.12, 0.19)$ (spur), $(0.07, 0.22)$ (blob), and $(0.005, 0.103)$ (control).
-The posterior samples for the amplitude of the feature mixture component are consistent with 0 in the control field, marginally inconsistent with 0 in the blob field, and significantly inconsistent with 0 in the spur field.
+The posterior samples for the amplitude of each feature mixture component are consistent with zero in the control field, marginally significant in the blob field, and significantly inconsistent with the background in the spur field.
 % In each of the feature fields, the posterior samples for the feature mixture component amplitude are inconsistent with 0, implying that the features are not consistent with the background stellar density.

### 2018-06-21 paper/gd1.tex
proper motion selection

@@
 
-As a first selection of GD-1 member stars, we choose stars with proper motions
+As a first selection of GD-1 member stars, we choose stars \changes{that lie in the orange polygon in proper motions (top right panel of \figurename~\ref{fig:selection}; approximately bounded by the rectangle
 $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot} <
-1~\masyr$ (rectangle in the top right panel of
-\figurename~\ref{fig:selection}).
+1~\masyr$).}
 \figurename~\ref{fig:selection} (top left) shows all stars that pass the
@@
 \textit{Panel (d)}: Median proper motion along the stream computed in the same
-bins as previous panel.
+bins as previous panel \changes{(not solar reflex corrected)}.
 \textit{Panel (e)}: Same as Panel (d), but for proper motion in stream latitude.

### 2018-06-21 paper/gd1.tex
polygon

@@
 
-As a first selection of GD-1 member stars, we choose stars \changes{that lie in the orange polygon in proper motions (top right panel of \figurename~\ref{fig:selection}; approximately bounded by the rectangle
+As a first selection of GD-1 member stars, we choose stars \changes{that lie in a polygon in proper motions (orange shaded polygon in top right panel of \figurename~\ref{fig:selection}; approximately bounded by the rectangle
 $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot} <

### 2018-07-09 paper/submit-aas/referee-response2.txt
stuff

> However, the maps sent to the referee but not included in the paper do
> not ease my concerns on how robust their conclusions are. For example,
> the figures show that the assumption of a uniform density may be
> flawed given the number of features in the data. Especially the region
> around the spur shows a high density of large values of the unit
> weight error.
>
> Since their procedure starts with a selection of tentative members
> based on proper motions, it would be good if the authors could make a
> figure that shows both the stars selected according to the proper
> motion (i.e. the sample they already have) and the stars with 2
> parameter solutions that fall within the CMD selection. It would also
> be useful to assess the relative number of good and bad solutions in
> the blob and the spur, as well as in the regions around the gaps.
>
> The issue is not so much the quality of the data that is kept (the 5
> parameter solutions), but what does incompleteness in this dataset
> imply? The presence of stars with only 2 parameter solutions implies
> that such stars would not be selected according to the procedure
> followed by the authors, and their exclusion could potentially lead to
> "features". A plot as I propose would help to double check that the
> exclusion of such stars (if any) does not give rise to gaps or
> spurious over-densities.

TODO: Ana needs to crossmatch the 2par solutions with pan-starrs

> I would also like to ask the authors to overplot their selected GD-1
> stream stars (i.e. those shown in the bottom panel of Fig. 1), in the
> maps of the density of 2 parameter solutions, the unit weight
> error and the extinction map.

We have attached these figures to the response.

> Finally a small point regarding the presence of a progenitor. Given
> that it is very difficult to identify the feature the authors
> associated to the progenitor clearly from the figures presented, and
> its significance is not demonstrated in the paper, it does not really
> seem to deserve a mention in the abstract.

We have removed this statement from the abstract.

### 2018-07-19 paper/submit-aas/referee-response2.txt
ref response 2
fraction

@@
 
-TODO: Ana needs to crossmatch the 2par solutions with pan-starrs
+We have attached the requested figure to the report. We do not see any correlation between the off-track features described in this paper and the CMD-selected stars with 2-parameter solutions.
+
+We have compared the fraction of good solutions in the spur, in the region between the stream and the spur, and in the stream at the same range of longitudes.
+We consider the longitude range -35 < phi1 < -30, and 3 ranges in phi2: (-0.4, 0.4), (0.4, 0.9), (0.9, 1.7) for the stream, region between stream and spur, and spur, respectively.
+In these 3 regions, we find 84%, 88%, 89% of the 5 parameter solution stars pass the unit weight error criteria.
 
@@
 We have attached these figures to the response.
+Again, while there is clearly structure in these maps, we do not see any relation between the data quality or incompleteness and the new off-track features reported in this paper.
+We do see issues with completeness at phi1 < -50 due to the scan pattern, and we discuss this in the text.
 

### 2018-07-23 paper/submit-aas/referee-response.txt
proper motion selection

@@
 Below, we have copied the remarks from the referee (indented by >) and have responded to each point in text that immediately follows.
+A few other minor changes have been made directly to the text and are indicated as bold text in the updated manuscript.
 
@@
 We have addressed each of the referee's comments in the attached response, but in doing so we have added a new figure that we believe improves the paper and directly addresses one of the referee's points.
-We realize that now the word count is slightly over the limit for ApJ letters (we count XX words).
+We realize that now the word count is a bit over the limit for ApJ letters (we count 4265 words).
 However, we do still feel that this result and paper are in the spirit of the Letters and would like to keep it in consideration for this journal.

### 2018-07-23 paper/gd1.tex
doi for dataset

@@
 The resulting map of GD-1: (1) extends the apparent length of the stream by
-$20^\circ$, (2) reveals a plausible location for the progenitor, (3) detects
-high-contrast gaps along the stream, and (4) indicates
+$20^\circ$, (2) detects high-contrast gaps along the stream, and (3) indicates
 the existence of stream members perturbed off the main stream track.
@@
 \changes{At this location, between $\phi_1 \in (-18, -10)^\circ$, we find that the mean equatorial sky position of the stream stars is $(\alpha, \delta) = (177.01, 53.99)^\circ$ and the mean proper motion is $(\mu_\alpha \, \cos\delta, \mu_\delta) = (-7.78, -7.85) \pm 0.03~\masyr$, taking into account the covariance matrix for proper motions provided with the \gaia\ data.}
+\changes{The full data for the GD-1 region along with selection masks are available through Zenodo (\dataset[10.5281/zenodo.1295543]{https://doi.org/10.5281/zenodo.1295543})\footnote{See also \url{https://github.com/adrn/gd1-dr2}.}}
 

### 2018-07-24 paper/gd1.tex
update model figure

@@
 The resulting map of GD-1: (1) extends the apparent length of the stream by
-$20^\circ$, (2) detects high-contrast gaps along the stream, and (3) indicates
-the existence of stream members perturbed off the main stream track.
+$20^\circ$, (2) \changes{reveals plausible locations for the progenitor}, (3) detects high-contrast gaps along the stream, and (4) indicates the existence of stream members perturbed off the main stream track.
 These discoveries are only possible because of the exquisite astrometry
@@
 Lines show a \changes{4th}-order polynomial fit to the median stream track offset by $\pm 0.75^\circ$; the enclosed region is adopted as the nominal stream region.
-\textit{Panel (b)}: Sky positions of simulated star particles from a mock stream
-model of the GD-1 stream.
-\textit{Panel (c)}: Background-subtracted surface-density estimates computed in
+\changes{\textit{Panels (b) and (c)}: Sky positions of simulated star particles from mock stream models of the GD-1 stream in which the progenitor is currently disrupting (toy model 1, panel b) or the progenitor fully disrupted 500 Myr ago and results in the observed under-density at $\phi_1 \sim -20^\circ$ (toy model 2, panel c).}
+\textit{Panel (d)}: Background-subtracted surface-density estimates computed in
 successive, overlapping $3^\circ$ windows for both GD-1 data and the model
 stream.
-\textit{Panel (d)}: Median proper motion along the stream computed in the same
+\textit{Panel (e)}: Median proper motion along the stream computed in the same
 bins as previous panel \changes{(not solar reflex corrected)}.
-\textit{Panel (e)}: Same as Panel (d), but for proper motion in stream latitude.
+\textit{Panel (f)}: Same as Panel (d), but for proper motion in stream latitude.
 }
@@
 
-To initialize the model stream generation, we assume that the present day
-position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak of the
-observed surface density profile.
-We numerically integrate backwards from this location for $4~\textrm{Gyr}$,
-then forward-generate the model stream, releasing stream particles at each
+\changes{To initialize the model stream generation, we assume that the present day position of the progenitor system is either at $\phi_1 = -13.5^\circ$, the peak of the observed surface density profile (toy model 1), or was fully-disrupted 500 Myr in the past and the under-density at $\phi_1 = -20^\circ$ represents the fully-disrupted progenitor (toy model 2).}
+We numerically integrate backwards from these locations for $4~\textrm{Gyr}$,
+then forward-generate the model streams, releasing stream particles at each
 timestep.
 We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
-the mass of the progenitor until its full disruption at the present.
+the mass of the progenitor until its full disruption at the \changes{present (toy model 1), or at $t = -500~\textrm{Myr}$ (toy model 2)}.
 % day, where we assume $M = 0~\msun$
 % (i.e. fully disrupted).
-Stream properties computed from this resulting model stream are plotted in
-\figurename~\ref{fig:track-and-model} as dashed (orange) lines.
-The model stream qualitatively matches the stream track in all phase-space
-dimensions, but has a much smoother surface density profile along the extent of
+Stream properties computed from the resulting model streams are plotted in
+\figurename~\ref{fig:track-and-model} as dashed (orange and purple) lines.
+The model streams qualitatively match the stream track in all phase-space
+dimensions, but have much smoother surface density profiles along the extent of
 the stream.
 The broadening in the observed stream at negative longitudes is also seen in the model stream and is expected for older parts of the stream that have phase-mixed for longer times.
-We therefore don't expect that the observed surface density variations can be fully explained
-by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).
+We therefore don't expect that the observed surface density variations can be fully explained by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).
 

### 2018-07-24 paper/submit-aas/referee-response2.txt

@@
 
-We have attached the requested figure to the report. We do not see any correlation between the off-track features described in this paper and the CMD-selected stars with 2-parameter solutions.
+We have attached the requested figure to the report. We do not see correlations between the off-track features described in this paper and the CMD-selected stars with 2-parameter solutions.
 
@@
 
-We have removed this statement from the abstract.
+We agree that the uniqueness of the previous claim was not strong, and we have both rephrased the text and added an alternate model stream for comparison.
+With Figure 2, we now show two plausible locations for the progenitor: one in which the progenitor disrupts at present day and results in the over-density, and one in which the progenitor disrupted several orbits ago and can explain one of the observed under-densities.

### 2018-07-24 paper/gd1.tex
toy model 2 rephrases

@@
 stream width to future work.
-\changes{With present data, however, we cannot rule out a scenario in which the progenitor system disrupted longer ago and resulted in one of the larger gaps (e.g., \citealt{DeBoer:2018}).}
+\changes{With present data, however, we cannot rule out a scenario in which the progenitor system disrupted longer ago and resulted in a gap at $\phi_1\approx-20^\circ$ (see Figure~\ref{fig:track-and-model} and, e.g., \citealt{DeBoer:2018}).}
 % that has not yet phase-mixed in multiple dimensions.
@@
 
-\changes{To initialize the model stream generation, we assume that the present day position of the progenitor system is either at $\phi_1 = -13.5^\circ$, the peak of the observed surface density profile (toy model 1), or was fully-disrupted 500 Myr in the past and the under-density at $\phi_1 = -20^\circ$ represents the fully-disrupted progenitor (toy model 2).}
+\changes{We present two progenitor configurations which reproduce certain aspects of the observed GD-1 density variations: the progenitor system is either at $\phi_1 = -13.5^\circ$, the peak of the observed surface density profile (toy model 1), or was fully-disrupted 500\,Myr in the past and the under-density at $\phi_1 = -20^\circ$ represents the fully-disrupted progenitor (toy model 2).}
+% \changes{To initialize the model stream generation, we assume that the present day position of the progenitor system is either at $\phi_1 = -13.5^\circ$, the peak of the observed surface density profile (toy model 1), or was fully-disrupted 500 Myr in the past and the under-density at $\phi_1 = -20^\circ$ represents the fully-disrupted progenitor (toy model 2).}
 We numerically integrate backwards from these locations for $4~\textrm{Gyr}$,

### 2018-07-25 paper/submit-aas/referee-response2.txt
ref response

@@
+We thank the referee for the suggestions.
+We believe we have now addressed all of the referee's comments either in the manuscript text or response below.
+We have also attached a number of figures to the response that were suggested or requested by the referee.
+
 > However, the maps sent to the referee but not included in the paper do
@@
 
+(we have added text to the manuscript, bolded, to describe the points above)
+
 > I would also like to ask the authors to overplot their selected GD-1

### 2018-07-25 paper/gd1.tex
fix d and add model lines
ref response

@@
 bins as previous panel \changes{(not solar reflex corrected)}.
+\changes{The two toy models have consistent median proper motions (overlapping dashed lines).}
 \textit{Panel (f)}: Same as Panel (d), but for proper motion in stream latitude.
@@
 \changes{
+In detail, we have compared the fraction of good solutions in the region near the spur, in the region between the stream and the spur, and in the stream at the same range of longitudes.
+We consider the longitude range $-35^\circ < \phi_1 < -30^\circ$, and 3 ranges in $\phi_2$: $(-0.4, 0.4)^\circ$, $(0.4, 0.9)^\circ$, $(0.9, 1.7)^\circ$ for the stream, region between stream and spur, and spur, respectively.
+In these 3 regions, we find 84\%, 88\%, 89\% of the stars with 5-parameter solutions pass the unit weight error criteria.
 Small-scale density variations, however, especially apparent between $-80^\circ < \phi_1 < -50^\circ$, are a consequence of the \gaia\ scanning pattern.
